### 2019-02-25 report.tex
Initial commit

\documentclass[11pt,a4paper]{article}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\DeclareMathOperator*{\argmax}{arg\,max}

\pagestyle{fancy}
\begin{document}
\author{F\'elix GONTIER \\ \texttt{felix.gontier@ls2n.fr}}
\title{CENSE - Perceptual test II}
\maketitle

\section{Introduction}

This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}. From recordings of a sensor network with low information content, this study aims at developing a processing chain for the estimation of pleasantness in urban sound environments.

\section{Test description}
\subsection{Objectives}
\label{sec:test_obj}

The objectives of this test are three-fold. First, the study aims at verifying the hypothesis that recorded and equivalent replicated sound scenes result in the same perceptual response, meaning that the development of models based on simulated data for real-life scenarios is feasible.\\

Very few recorded scenes are fully annotated in terms of backgrounds, event onsets-offsets and sound levels. However, the prediction of source-specific acoustic indicators through deep learning models requires large amounts of data for which the ground truth of these indicators is known. An alternative to the subjective annotation of corpora of sound scenes is the use of fully simulated scenes, where source activity is controlled. Thus, it is important to verify that new simulated scenes generate a perceptual space similar to that of real scenarios, even with low source complexity.\\

Finally, this work aims at finding physical indicators useful for the estimation of pleasantness. Previous studies consistently model pleasantness as a function of perceptual source activity descriptors such as the time of presence. These quantities are somewhat closer to the physical properties of the scene than the more abstract concept of pleasantness. This work thus primarily focuses on the relation of perceived source activity with physical indicators which can be predicted using deep learning architectures.

\subsection{Data}
\label{sec:test_data}

With respect to the three discussed objectives, a test corpus is created based on recordings from one of the four soundwalks studied in~\cite{aumond}. The scenes correspond to 19 locations in the 13th district of Paris. For each recording 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception.
The first objective requires the comparison of perceptual assessments for recorded and replicated sound scenes. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) with diverse environments are picked to explore different real-life situations, as a function of the classifications proposed in~\cite{jrg}: \textit{parks}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For the second part of the corpus sound scenes are simulated based on the 19 45~s extracts. The recordings are first annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and are associated with a sound level.
\item Events characterized for each occurence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound (NOTE?) contributions for all other sources. For this experiment, the playback sound level is the same as during recordings, and ranges from 63.9~dB to 79.4~dB.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/tvb_pres.eps}
    \caption{Estimated time of presence space for simulated scenes.}\label{fig:tvb_pres}
\end{figure}

To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios. Because of the lack of large isolated sources datasets and the previously proposed models of pleasantness, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most encounterable real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}. These statistics include, conditionally to the ambiance:
\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurences in seconds, also in the form of normal distributions.
\end{itemize}
Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to enhance realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. For each environment diverse new scenarios are generated by sampling the background and event properties distributions. For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus is thus comprised of a total of 100 scenes.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.4\textwidth]{figures/leqa.png}
    \caption{Scale used to generate sound levels as a function of the environment for new simulated scenes.}\label{fig:amb_levels}
\end{figure}

\subsection{Questions}

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes (CITE). While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies (CITE) show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/interface.png}
    \caption{Screenshot of the test interface used by all participants.}\label{fig:interface}
\end{figure}

\subsection{Test conditions}

Prior to the test, participants are given an introduction and shown the interface before the test to ensure that the quantities are understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 simulated, then to 25 of the 75 simulated. The evaluated simulated scenes are chosen so that each scene is evaluated by the same number of participants. A random pre-generated order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes to calibrate their answers. For the remainder of the corpus all participants experience different scene ordering. Participants can only listen to each scene once. They can start answering questions before the end of the sample, but have to answer all questions and listen to the full extract before proceeding to the next.\\

The scenes are played at a given sound level as discussed in Section~\ref{sec:test_data}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber, resulting in a relation between playback Leq and LAeq and sound card output voltage for pink noise. This ensures that correctly scaled scenes will be heard at the desired sound level for all participants.\\

A total of 23 students at Ecole Centrale de Nantes completed the test.

\section{Perceptual analysis}

This section studies the test results with regard to the objectives presented in Section~\ref{sec:test_obj}.\\

Participants evaluate each scene on criteria coded on discrete scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

\subsection{Correspondance between recorded and replicated scenes}

The relevance of the study relies on the perceptual equivalence of recorded and synthetic scenes. To verify this hypothesis, the 6 recorded and corresponding replicated scenes are compared in terms of perceptual responses.\\

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This particular test is chosen because while it assumes that distributions are continuous, it is non-parametric. Pleasantness is found to differ between the two groups only for one location, P4, which is explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is significantly louder in the replicated scene than it is in the recording for about half
of its duration. Although to a lesser extent the same issue explains the high difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

Overall, most perceptual responses for replicated scenes are close to those for recordings. No major difference is consistently found in a particular scale, and the proposed hypothesis cannot be invalidated.

\begin{table}[h]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Kolmogorov-Smirnov test are in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & -1.04 & 0.43 & 0.13 & -0.91 & 0.39 & -2.09 & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & \textbf{0.91} & 0 & \textbf{-1.83} & 0.48 & 1.30 & -0.96 & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & -0.91 & 0.09 & \textbf{-1.43} \\
	P15 & -1.35 & 0.52 & 0.52 & -1.17 & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & 1.13 & -0.30 & -1.17 & -0.43 & 1.39 & -1.04 & -1.83 & 0.83 & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table}

\subsection{Perceptual space}

The perceptual space generated by the five general scales can also be studied and compared to previous studies to further validate the test material. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. This confirms that the simulation process generates perceptually realist scenes and can be used for further research.\\

Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by black segments. The space covered by scenes based on new scenarios covers that of the studied real-life environments, demonstrating the efficiency of the scene generation procedure in terms of diversity.

% FIGURE
\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_rep.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
    \end{subfigure}
    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252). Each dot represent the projection of one scene on the resulting space. The components space is distorted compared to the litterature.}\label{fig:pspace}
\end{figure}

\subsection{Perceptual scales relationships}
\label{sec:percm}

\begin{table}
\centering
\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
\label{tab:percc}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
\end{tabular}
%}
\end{table}

Finally, we study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond, lavandier}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while here no direct relation is visible. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds. The perceptual assessment of the sound level of passing vehicles $L_{T, p}$ is a potential alternative to that of the time of presence of traffic $T_{T, p}$. Except for the interest (I) it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\

To further compare this experiment's results to previous work, multilinear models of pleasantness are built as a function of both the overall loudness (OL) source-specific parameters. To ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to model regression. Only models for which all predictors verify $VIF<5$ are considered valid.\\

The best resulting models with statistically significant estimates ($p<0.05$ for all predictors) are:

\begin{align}
\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p}\\
\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p}
\end{align}

Both models have very similar expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
\begin{equation}
\hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
\end{equation}
although in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness. This is expected given the observed differences in correlations values.\\

Table~\ref{tab:percm} summarizes the performance of the three models. Again, the two models computed on the present corpus display similar, low estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model. However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus.

\begin{table}[t]
\centering
\caption{Performance of perceptual models for pleasantness prediction (**: p<0.01).}
\label{tab:percm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{p1}$ & 0.59 & 0.91 & 0.95** \\
	$\hat P_{p2}$ & 0.61 & 0.90 & 0.95** \\
	$\hat P_{p3}$~\cite{ricciardi2014} & 1.05 & 0.77 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

\section{Physical analysis}

From results obtained in the previous section, a perceptual model is assumed where pleasantness is a function of the perceived overall loudness and time of presence of traffic, voices and birds. These parameters are not easily predictable from recordings because of their subjectivity. Thus, there is a need to propose a similar model of pleasantness from physically-based indicators that can either 1) be computed directly from the audio signal, or 2) be predicted using a learned parametric model taking as input the audio signal.

\subsection{Indicators}

Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond, gontier, lavandier} to correlate well with content-related perceptual parameters: the overall loudness and the source-specific time of presence.\\

First, indicators can be computed directly from the mixed audio, without needing information about its composition. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{ita}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond}, that is found to correlate well with the perceptual time of presence of birds.
\end{itemize}

Furthermore, the simulation process outputs ground truth source contributions in the form of separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

\begin{itemize}
\item $L_{eq, s}$: Equivalent sound level for source $s$
\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined
\end{itemize}

Finally, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed. $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:

\begin{align}
\hat T_s(\alpha) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}_{\Delta_s(t)>\alpha}\\
\alpha_{opt} &= \argmax_{\alpha}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha)\right)
\end{align}

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. While this indicator has a simple expression, it does not reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\

Similarly, $\hat T_s(\alpha, \beta)$ is based on a binary source emergence model, but is computed on the third-octave band emergence $\Delta_s(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$ as a simple spectral masking model. It is parametrized by $\alpha$ and $\beta$ thresholds:

\begin{align}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
\end{align}

The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\

In the remainder of this report, the subscript $s$ is replaced with the corresponding source initial: $T$ for traffic, $V$ for voices and $B$ for birds.

\subsection{Relation between perceptual an physical data}
\label{sec:physm}

Analysis is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.\\

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study.\\

Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source. Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weaker for other sources. Both proposed estimates of the time of presence $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ show strong relations to their perceptual counterparts with high correlations (r>0.8), though the latter is consistently better. An interesting result is the high correlation between the $L_{10}$ which is interpreted as a measurement of emergent events to the perceived sound level of passing vehicles $L_{T, p}$. If in perceptual models there is a need to include the sound level of traffic events, the $L_{10}$ may be sufficient instead of a source separation and level estimation algorithm.\\

\begin{table}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\ \hline
	$TFSD_{4kHz}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha, \beta)$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
	$\hat T_V(\alpha, \beta)$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
	$\hat T_B(\alpha, \beta)$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
	$\hat T_T(\alpha)$ & -0.47** & NS & 0.31** & -0.52** & -0.26* & 0.52** & 0.78** & -0.34** & -0.33** \\
	$\hat T_V(\alpha)$ & NS & 0.45** & NS & 0.30** & NS & NS & -0.30** & 0.81** & NS \\
	$\hat T_B(\alpha)$ & 0.61** & -0.37** & -0.50** & 0.57** & 0.56** & -0.46** & -0.56** & NS & 0.88** \\ \hline
\end{tabular}
}
\end{table}

Subsequently, multilinear models are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:percm} a VIF check is performed on predictors to ensure that no multi-collinearity exists. The best pleasantness model optimized on the present data is:
\begin{equation}
\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta)
\end{equation}
The absence of indicators related to traffic and voices activity is explainable:
\begin{itemize}
\item The $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively
\item In this corpus no strong contribution on pleasantness is identified from voices, with no significant correlation for both perceptual (Table~\ref{tab:percc}) and physical (Table~\ref{tab:physc}) indicators.
\end{itemize}
It is also possible to use the models discussed in Section~\ref{sec:percm} as is. In this situation, instead of predicting pleasantness directly, each intermediate perceptual parameter is modeled as a function of available physical variables. Here, there is already a one-to-one association between perceptual and physical quantities, and the models are linear:
\begin{align}
\hat{OL} &= 0.22~L_{50} - 8.4\\
\hat{T}_{T, p} &= 6.6~\hat T_{T}(\alpha, \beta) + 0.25\\
\hat{T}_{V, p} &= 6.3~\hat T_{V}(\alpha, \beta) + 1.54\\
\hat{T}_{B, p} &= 7~\hat T_{B}(\alpha, \beta) + 2.5
\end{align}
Substituting these equations in equation (3), the pleasantness model becomes:
\begin{equation}
\hat P_{\varphi 2} = 10.95 - 0.08~L_{50} - 0.99~\hat T_{T}(\alpha, \beta) + 1.26~\hat T_{V}(\alpha, \beta) + 1.05~\hat T_{B}(\alpha, \beta)
\end{equation}

Table~\ref{tab:physm} compares the performance of the two models. Both obtain good estimates of pleasantness for a 10-point scale. Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its contruction process. It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness. This is particularly important in the case of voices, found in multiple previous studies to have a positive effect. Conversely, the $\hat P_{\varphi 2}$ model uses a perceptual model constructed on a much larger corpus of recorded scenes, and is less prone to these specificity properties.

\begin{table}[t]
\centering
\caption{Performance of physical models for pleasantness prediction (**: p<0.01).}
\label{tab:physm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{\varphi 1}$ & 0.83 & 0.82 & 0.91** \\
	$\hat P_{\varphi 2}$ & 1.30 & 0.62 & 0.80** \\ \hline
\end{tabular}
%}
\end{table}

\section{Physical indicators prediction}

\subsection{Objective}

In the previous section we identified indicators based on masking models that can be computed from separated sources audio in the scene. These indicators provide estimates of the perceived time of presence that can be used to approximate pleasantness using a simple linear model with relatively low error. The objective of this section is to propose a deep learning architecture to predict the source-specific quantities needed for this approximation from mixed sound scenes.

\subsection{Choice of indicators}

Based on the results outlined by the perceptual experiment, a processing chain is obtained for the estimation of pleasantness in urban sound scenes as shown in Figure~\ref{fig:ppred}. Pleasantness is modeled as a function of overall loudness and the perceived times of presence of traffic, voices and birds sources. However, predicting these quantities directly using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels. However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}. While the same need for annotation of large datasets exists, it can be circumvented in the case of simulated data, for which the contribution of each source in the mix is available. Thus, a deep learning model can be used to predict either of the first steps of the pleasantness estimation chain.\\

An advantage of predicting sound levels is the accuracy of the labels as they are computed directly on available separated sources audio. On the condition that isolated samples used during scene generation are of good quality in terms of low other source content, the resulting ground truth labels can be considered exact. However, the prediction is a regression task which is typically difficult to learn. The best studied masking model relies on third-octave band sound levels, meaning that the learned model needs to perform an implicit source separation. Moreover, once an estimations of the sound levels are obtained the selected masking model must be used as the next step of pleasantness prediction. This model is not fully accurate and thus leads to small errors in pleasantness values even for perfect sound level estimations.\\

Instead, we can choose to directly predict if each of the three sources is present (heard) at a given time. This is either a detection problem if the data is processed asynchronously or a multi-label classification problem if the data is processed in frames, both of which are much simpler tasks. The corresponding labels are directly the outputs of the best studied masking model $\hat T(\alpha, \beta)$. Its uncertainties are thus transferred to "weak" labels rather than present as part of the constant processing chain used to obtain pleasantness estimations.\\

In this report two models for source-specific sound level and presence estimation are implemented and compared.

\subsection{Data}

Two datasets of simulated scenes are created for the proposed task:
\begin{itemize}
\item The development set containing 400 45~s scenes (total 5 hours) is used during the training process
\item The evaluation set containing 200 45~s scenes (total 2.5 hours) is used to compute the model's performance and generalisation capabilities
\end{itemize}

The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus used during the perceptual test. Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples for other sources. The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\

The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation. Thus, the model should take as input the same representation. Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration. This scale is considered relevant for perception and was already used to compute the masking models proposed in the perceptual experiment results analysis. The input frames of dimension 29x8 are processed independantly in this study.\\

Target outputs differ for the level estimation and source detection. In the former case labels are obtained by computing the full-band sound level of source-specific audio channels. For the latter task the weak presence labels given by $\hat T(\alpha, \beta)$ before averaging along time are used. Estimates of pleasantness are obtained as follows:
\begin{itemize}
\item Sound level model: the outputs are used to compute the $\hat T(\alpha)$ indicator, which is then applied to equation (13),
\item Presence model: outputs are averaged along time and considered as $\hat T(\alpha, \beta)$ in equation (13).
\end{itemize}

\subsection{Model}

Deep learning models are implemented and trained separately for both sound level and presence prediction tasks, with architectures shown in Figure~\ref{fig:deep_arch}. For both tasks, the models include 4 blocks of convolutional layers followed by LeakyReLU activations. The convolutional layers have respectively 128, 64, 32 and 32 output maps (channels), and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. For the model trained to predict sound levels these 3 values are taken directly as the model output, corresponding to the estimated level of traffic, voices and birds respectively. An additional sigmoid activation is added for the model predicting source presence to obtain outputs in the 0-1 range.

\subsection{Training details}

Training the models is achieved by progressively minimizing a loss function $L(y, \hat y)$, xhere $y$ and $\hat y$ represent the target and predicted outputs respectively. For sound level estimation we use the MSE or L2 loss:
\begin{equation}
L(y, \hat y) = \sum_s \left(\hat y_s - y_s\right)^2
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted sound level of source $s$ in dB. For presence prediction the cost function is the binary cross-entropy of equation:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.\\

Models are optimized using the Adam algorithm~\cite{adam} on mini-batches of 64 1s examples. For all experiments a constant learning rate of $\lambda = 0.001$ is used. Training is performed during 10 epochs, that is iterations over the whole training set. As shown in Figure~\ref{fig:mdl_loss}, at that point the training loss is still decreasing but not the validation loss, indicating that the model starts overfitting the training data.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/mdl_loss.eps}
    \caption{Evolution of the presence detection model's loss as a function of training iterations.}\label{fig:mdl_loss}
\end{figure}

\subsection{Metrics}

Performance metrics can be computed at either part of the processing chain shown in Figure~\ref{fig:ppred}. At the physical level, the following are considered:

\begin{itemize}
\item Sound level error for the sound level prediction model only as the presence prediction model bypasses this step,
\item Accuracy of source presence detection in 1~s frames with labels taken from the computation of $\hat T_s(\alpha, \beta)$ before averaging,
\item Error of the resulting time of presence with $\hat T_s(\alpha, \beta)$ as the assumed ground truth.
\end{itemize}

Additionally, the pleasantness estimates resulting from each model can be compared to assessments, although on the perceptual experiment corpus only as no subjective annotation was performed on the constructed evaluation dataset.

\subsection{Results}

Table~\ref{tab:perf_cmp} displays the performance of the two studied models for metrics at the physical level. The first model outputs very poor estimations of the source-specific sound level, but the resulting source presence accuracy is high at 92.51\%. As a result, the time of presence estimates are precise with a root mean-squared error of 0.12 on a 0-1 scale. This behaviour may be explained by the permissiveness of the $\hat T_s(\alpha)$ indicator: if sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values. Thus, when using $\hat T_s(\alpha)$ it is sufficient to obtain good presence detection that the model output either very low values when the source is absent or poor estimates of the level when it is present. Here, the model fails at the source separation and level estimation task and is probably closer to a detection model.\\

The presence detection model performs similarly well for both presence accuracy (92.11\%) and resulting $\hat T_s(\alpha, \beta)$ estimation (0.12 RMSE). Both models obtain lower accuracies for birds than for traffic or voices. However, the origin of errors varies. The sound level model has an extremely low false positive rate consistently across sources while its false negative rate varies from 7.8\% for traffic to more than 18\% for birds. The reverse is visible for the presence model which displays a higher false positive rate for birds. Overall, traffic is found to be the most easily detected source, which may be due to its location in the lower frequency bands while voice and bird information is usually concentrated in narrower frequency ranges when using third-octave bands.\\

\begin{table}[t]
\centering
\caption{Comparison of performances of models predicting source sound level or presence with metrics at the physical level. Levels and presence metrics are computed for n=8600 1s frames and time of presence metrics on n=200 45s scenes. (TP: true positive, TN: true negative, FP: false positive, FN: false negative)}
\label{tab:perf_cmp}
\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c c c | c | c c c }
\hline
	Model & \multicolumn{4}{|c}{Sound level} & \multicolumn{4}{|c}{Presence} \\ \hline
	 & All sources & Traffic & Voices & Birds & All sources & Traffic & Voices & Birds \\ \hline
	Sound level RMSE (dB) & 12.91 & - & - & - & - & - & - & - \\
	Sound level MAE (dB) & 6.91 & - & - & - & - & - & - & - \\ \hline
	Presence accuracy (\%) & 92.51 & 93.65 & 92.90 & 90.98 & 92.11 & 94.09 & 92.15 & 90.08 \\
	Presence TP (\%) & 88.20 & 92.16 & 88.08 & 81.62 & 91.96 & 92.70 & 89.93 & 93.37 \\
	Presence TN (\%) & 97.86 & 97.34 & 98.32 & 97.74 & 92.30 & 97.54 & 94.76 & 87.70 \\
	Presenct FP (\%) & 2.14 & 2.66 & 1.68 & 2.26 & 7.70 & 2.46 & 5.24 & 12.30 \\
	Presence FN (\%) & 11.80 & 7.84 & 11.92 & 18.38 & 8.04 & 7.30 & 10.17 & 6.63 \\ \hline
	$\hat T_s(\alpha, \beta)$ RMSE & 0.12 & 0.14 & 0.11 & 0.12 & 0.12 & 0.13 & 0.10 & 0.11 \\
	$\hat T_s(\alpha, \beta)$ MAE & 0.06 & 0.06 & 0.06 & 0.07 & 0.06 & 0.05 & 0.06 & 0.08 \\ \hline
\end{tabular}
}
\end{table}

Finally, a comparison is made of the models's capacity to result in relevant pleasantness estimates. For this analysis the perceptual experiment is used as assessments are available. It is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
The results are shown in Table~\ref{tab:pppred}. Again, both models lead to similarly correct estimations of pleasantness with errors lower than 1.20 on a 10-point scale. Interestingly, predictions are better on the first tested sub-corpus which contains other sources, indicating that models are quite robust to this type of perturbation in the data. Though, a larger set of scenes and environments with perceptual annotations would be required to further validate this point.\\

The third part of Table~\ref{tab:pppred} refers to values of pleasantness obtained when using directly the $\hat T_s(\alpha, \beta)$ time of presence computed from ground truth separated channels (equation(13)). For the first sub-corpus only the 19 replicated scenes are used as the ground truth information is not available for the recorded part. For all the computed metrics the resulting error is higher than for both studied models. This is explained by the uncertainties of the $\hat T_s(\alpha, \beta)$ indicator that result from the simplicity of its masking model. The predictions from both models' outputs are in fact closer to the performance of the $\hat Pp3$ perceptual model presented in Section~\ref{sec:percm} (equation (3)).\\

Considering the abovementioned potential unstability of the sound level model, the presence detection model is to be preferred for further development.

\begin{table}[t]
\centering
\caption{Comparison of pleasantness prediction quality from predicted sound level and presence of sources. The experiment corpus is split into respectively the 25 recorded and replicated scenes and 75 simulated scenes.}
\label{tab:pppred}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c | c c | c c }
\hline
	Model & \multicolumn{2}{|c}{Sound level} & \multicolumn{2}{|c}{Presence} & \multicolumn{2}{|c}{$\hat T_s(\alpha, \beta)$} \\ \hline
	Sub-corpus & 1 & 2 & 1 & 2 & 1 & 2 \\ \hline
	RMSE & 1.02 & 1.17 & 1.12 & 1.18 & 1.33 & 1.29 \\ 
	MAE & 0.79 & 0.97 & 0.82 & 0.95 & 1.00 & 1.02 \\ \hline
	r & 0.85** & 0.79** & 0.86** & 0.80** & 0.76** & 0.77** \\ \hline
\end{tabular}
%}
\end{table}

\section{Conclusion}

A perceptual experiment is conducted with the objectives of 1) demonstrating that scenes replicated from annotations of recordings yield similar perceptual responses to these recordings, 2) simulated scenes with new scenarios can be used to develop perceptual models, 3) physical indicators can be computed from audio signals from which pleasantness can be efficiently predicted. These three hypotheses are verified, and a pleasantness model relying on presence detection for traffic, voices and bird sources is proposed.\\

Subsequently, a deep learning model with simple, state-of-the-art architecture is developed for this task. The model takes as input the fast third-octave band representation of the mixed scene and is trained on a corpus of simulated scenes to output presence labels in short time frames of 1s. Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean-squared error on a 10-point scale.\\

Future work will be mainly focused on the study of the robustness of developed models to new environment and data. Here only three sources were considered during training, and a limited isolated sound sources database. Thus, pleasantness estimation chain needs to be adapted to the city of Lorient, in which the Cense sensor network is to be implemented, as it contains specific sound environments as well as unique sources that were not considered in this study.

\clearpage

\textbf{TODO}

\textbf{Text}

\begin{itemize}
\item Appendix A
\end{itemize}

\textbf{Figures}

\begin{itemize}
\item Pleasantness prediction chain
\item Deep learning architecture
\item Headphones calibration
\end{itemize}

\clearpage

%\begin{itemize}
%\item 
%\item 
%\item 
%\end{itemize}

%\begin{table}
%\centering
%\caption{}
%\label{tab:}
%%\resizebox{\columnwidth}{!}{
%\begin{tabular}{ c | c c c c c c c c c }
%\hline
%	 & P & L & OL & I & C & $V_{V, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
%	P1 &  &  &  &  &  &  &  &  &  \\
%	P3 &  &  &  &  &  &  &  &  &  \\
%	P4 &  &  &  &  &  &  &  &  &  \\
%	P8 &  &  &  &  &  &  &  &  &  \\
%	P15 &  &  &  &  &  &  &  &  &  \\
%	P18 &  &  &  &  &  &  &  &  &  \\ \hline
%\end{tabular}
%%}
%\end{table}

\begin{thebibliography}{9}
%\bibitem{axelsson2010}
%Axelsson, \"O.; Nilsson, M.E.; Berglund, B.
%A principal components model of soundscape perception. \textit{J. Ac. Soc. Am.}, 2010
%
%\bibitem{jeon2018}
%Jeon, J.Y.; Hong, J.Y.; Lavandier, C; Lafon, J.; Axelsson, \"O.; Hurtig, M.
%A cross-national comparison in assessment of urban park soundscapes in France, Korea, and Sweden through laboratory experiments. \textit{Applied Acoustics}, 2018
%
%
%\bibitem{aumond}
%Aumond, P.; Can, A.; De Coensel, B.; Botteldooren, D.; Ribeiro, C.; Lavandier, C.
%Modeling soundscape pleasantness using perceptual assessments and acoustic measurements along paths in urban context. \textit{Acta Acustica united with Acustica}, 2017
%
%\bibitem{ita}
%Berzborn, M.; Bomhardt, R.; Klein, J.; Richter, J.-G.; Vorlander, M.
%The ITA-Toolbox: An Open Source MATLAB Toolbox for Acoustic Measurements and Signal Processing. \textit{In proceedings of the 43th Annual German Congress on Acoustics}, 2017
%
%\bibitem{simscene}
%Rossignol, M.; Lafay, G.; Lagrange, M.; Misdariis, N.
%SimScene : a web-based acoustic scenes simulator, \textit{In proceedings of the 1st Web Audio Conference (WAC)}, 2015
%
%\bibitem{gloaguen}
%Gloaguen, J.-R.; Can, A.; Lagrange, M.; Petiot, J.-F.
%Creation of a corpus of realistic urban sound scenes with controlled acoustic properties. \textit{In proceedings of the 173rd Meeting of the Acoustical Society of America and the 8th Forum Acusticum (Acoustics'17)}, 2017
\end{thebibliography}

\end{document}

### 2019-02-25 report.tex
proff read 1

@@
 \usepackage{bbm}
+\usepackage{color}
 \DeclareMathOperator*{\argmax}{arg\,max}
-
+\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
 \pagestyle{fancy}
@@
 This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}.
-From recordings of a sensor network with low information content, this study aims at developing a processing chain for the estimation of pleasantness in urban sound environments.
+From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.
 
@@
 Very few recorded scenes are fully annotated in terms of backgrounds, event onsets-offsets and sound levels.
-However, the prediction of source-specific acoustic indicators through deep learning models requires large amounts of data for which the ground truth of these indicators is known.
+At the same time, the prediction of source-specific acoustic indicators through deep learning models requires large amounts of data for which the ground truth of these indicators is known.
 An alternative to the subjective annotation of corpora of sound scenes is the use of fully simulated scenes, where source activity is controlled.
-Thus, it is important to verify that new simulated scenes generate a perceptual space similar to that of real scenarios, even with low source complexity.\\
+It is therefore important to verify that new simulated scenes generate a perceptual space similar to that of real scenarios, even with low source complexity.\\
 
-Finally, this work aims at finding physical indicators useful for the estimation of pleasantness.
+Next, this work aims at finding physical indicators that can be considered for the estimation of pleasantness.
 Previous studies consistently model pleasantness as a function of perceptual source activity descriptors such as the time of presence.
@@
 \item Background sources that are present throughout the whole scene, and are associated with a sound level.
-\item Events characterized for each occurence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
+\item Events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
 \end{itemize}
@@
 The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources.
-This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound (NOTE?) contributions for all other sources.
-For this experiment, the playback sound level is the same as during recordings, and ranges from 63.9~dB to 79.4~dB.\\
+This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources.
+For this experiment, the playback sound level is the same as during recordings \ml{of the references scenes ?}, and ranges from 63.9~dB to 79.4~dB.\\
 
@@
 To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios.
-Because of the lack of large isolated sources datasets and the previously proposed models of pleasantness, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
-Considering the aim of the study, newly generated scenarios should cover most encounterable real-life situations while remaining perceptually plausible.
+Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
+Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
 Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}.
@@
 \item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
-\item The inter-onset of event occurences in seconds, also in the form of normal distributions.
+\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
 \end{itemize}
-Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to enhance realism.
+Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to improve realism.
 An additional \textit{place} ambiance is added with properties derived empirically from available data.
@@
 
-The experiment corpus is thus comprised of a total of 100 scenes.
+The experiment corpus totals 100 scenes.
 
@@
 Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales.
-They are presented in french and translated in this report.
+They are presented in french and translated in this report using standard terminology.
 The first 5 scales relate to general properties of the scene:
@@
 
-Prior to the test, participants are given an introduction and shown the interface before the test to ensure that the quantities are understood.
+Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood.
 Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 simulated, then to 25 of the 75 simulated.
@@
 
-Participants evaluate each scene on criteria coded on discrete scales.
+Participants evaluate each scene on criteria coded on a discrete scales \ml{how many steps ?}.
 Many statistical tools assume that the studied data follows continuous, single mode distributions.
@@
 
-\subsection{Correspondance between recorded and replicated scenes}
+\subsection{Correspondence between recorded and replicated scenes}
 
-The relevance of the study relies on the perceptual equivalence of recorded and synthetic scenes.
+The relevance of the remaining study relies on the perceptual equivalence of recorded and synthetic scenes.
 To verify this hypothesis, the 6 recorded and corresponding replicated scenes are compared in terms of perceptual responses.\\
@@
 Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold.
-This particular test is chosen because while it assumes that distributions are continuous, it is non-parametric.
-Pleasantness is found to differ between the two groups only for one location, P4, which is explained by high discrepancies in the perceptual time of presence of traffic and voices.
+This non parametric test is chosen because it assumes that distributions are continuous.
+Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices.
 For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half.
 Replicating this scene using simScene imposes a constant sound level for background sources.
-Thus, the background traffic is significantly louder in the replicated scene than it is in the recording for about half
+Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
 of its duration.
-Although to a lesser extent the same issue explains the high difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly.
+To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, \ml{trop long : but as voices have a less distinct influence on pleasantness this is not reflected directly}.
 Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy.
-For example no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\
+For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\
 
 Overall, most perceptual responses for replicated scenes are close to those for recordings.
-No major difference is consistently found in a particular scale, and the proposed hypothesis cannot be invalidated.
+No major difference is found in a particular scale, and the proposed hypothesis, \ml{that is that ...} cannot be invalidated.
 
@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-This confirms that the simulation process generates perceptually realist scenes and can be used for further research.\\
+This confirms that the simulation process generates perceptually convincing scenes and can be used for further research.\\
 
@@
 
-Finally, we study the relations between perceptual scales with respect to existing work.
+We then study the relations between perceptual scales with respect to existing work.
 Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
@@
 Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
-In previous studies a small positive contribution of voices to pleasantness was found, while here no direct relation is visible.
+In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
 This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
@@
 
-Both models have very similar expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
+Both models have very similar formal expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
 \begin{equation}
 \hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
-\end{equation}
-although in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness.
+\end{equation}.
+Though, in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness.
 This is expected given the observed differences in correlations values.\\
@@
 
-From results obtained in the previous section, a perceptual model is assumed where pleasantness is a function of the perceived overall loudness and time of presence of traffic, voices and birds.
+From results obtained in the previous section, a perceptual model is chosen where pleasantness is a function of the perceived overall loudness and time of presence of traffic, voices and birds.
 These parameters are not easily predictable from recordings because of their subjectivity.
-Thus, there is a need to propose a similar model of pleasantness from physically-based indicators that can either 1) be computed directly from the audio signal, or 2) be predicted using a learned parametric model taking as input the audio signal.
+Thus, there is a need to propose a model of pleasantness whose parameters can be estimated from physically-based indicators that can either 1) be computed directly from the audio signal, or 2) be predicted using a learned parametric model taking as input the audio signal.
 
@@
 
-First, indicators can be computed directly from the mixed audio, without needing information about its composition.
+First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} when the source are active and at which level.
 Typically this includes indicators derived from sound level measurements.
@@
 
-Finally, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed.
+Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed.
 $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:
@@
 where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments.
-While this indicator has a simple expression, it does not reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\
+While this indicator has a simple expression, it is not found to reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\
 
@@
 First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness.
-Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study.\\
+Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study \ml{because ??}.\\
 
@@
 Both proposed estimates of the time of presence $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ show strong relations to their perceptual counterparts with high correlations (r>0.8), though the latter is consistently better.
-An interesting result is the high correlation between the $L_{10}$ which is interpreted as a measurement of emergent events to the perceived sound level of passing vehicles $L_{T, p}$.
-If in perceptual models there is a need to include the sound level of traffic events, the $L_{10}$ may be sufficient instead of a source separation and level estimation algorithm.\\
+An interesting result is the large correlation between the $L_{10}$ which is interpreted as a measurement of emergent events to the perceived sound level of passing vehicles $L_{T, p}$.
+In perceptual models there may be a need to include the sound level of traffic events.
+In this case, the $L_{10}$ may be sufficient instead of a source separation and level estimation algorithm.\\
 
@@
 Both obtain good estimates of pleasantness for a 10-point scale.
-Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its contruction process.
+Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its construction process.
 It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness.

### 2019-02-25 report.tex
proof read done

@@
 
-In the previous section we identified indicators based on masking models that can be computed from separated sources audio in the scene.
-These indicators provide estimates of the perceived time of presence that can be used to approximate pleasantness using a simple linear model with relatively low error.
+In the previous section,  indicators based on masking models that can be computed from separated sources audio in the scene are proposed.
+These indicators provide estimates of the perceived time of presence that can be used to approximate pleasantness using a simple linear model with relatively low prediction error.
 The objective of this section is to propose a deep learning architecture to predict the source-specific quantities needed for this approximation from mixed sound scenes.
@@
 Pleasantness is modeled as a function of overall loudness and the perceived times of presence of traffic, voices and birds sources.
-However, predicting these quantities directly using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels.
+
+Predicting these quantities directly \ml{from what ?} using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels.
 However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}.
-While the same need for annotation of large datasets exists, it can be circumvented in the case of simulated data, for which the contribution of each source in the mix is available.
+While the same need for annotation of large datasets exists, it can be circumvented by the use of simulated data, for which the contribution of each source in the mix is available.
 Thus, a deep learning model can be used to predict either of the first steps of the pleasantness estimation chain.\\
@@
 An advantage of predicting sound levels is the accuracy of the labels as they are computed directly on available separated sources audio.
-On the condition that isolated samples used during scene generation are of good quality in terms of low other source content, the resulting ground truth labels can be considered exact.
+Assuming that isolated samples used during scene generation are of good quality in terms of low other source content, the resulting ground truth labels can be considered as reliable.
 However, the prediction is a regression task which is typically difficult to learn.
 The best studied masking model relies on third-octave band sound levels, meaning that the learned model needs to perform an implicit source separation.
-Moreover, once an estimations of the sound levels are obtained the selected masking model must be used as the next step of pleasantness prediction.
+Moreover, once an estimation of the sound levels are obtained the selected masking model must be used as the next step of pleasantness prediction.
 This model is not fully accurate and thus leads to small errors in pleasantness values even for perfect sound level estimations.\\
@@
 The corresponding labels are directly the outputs of the best studied masking model $\hat T(\alpha, \beta)$.
-Its uncertainties are thus transferred to "weak" labels rather than present as part of the constant processing chain used to obtain pleasantness estimations.\\
+Its uncertainties are thus transferred to "weak" labels rather than considered as part of the constant processing chain used to obtain pleasantness estimations.\\
 
-In this report two models for source-specific sound level and presence estimation are implemented and compared.
+In this report, two models for source-specific sound level and presence estimation are implemented and compared. \ml{manque un lien entre les deux derniers par}
 
@@
 \item The development set containing 400 45~s scenes (total 5 hours) is used during the training process
-\item The evaluation set containing 200 45~s scenes (total 2.5 hours) is used to compute the model's performance and generalisation capabilities
+\item The evaluation set containing 200 45~s scenes (total 2.5 hours) is used to compute the model's performance and its generalization capabilities
 \end{itemize}
 
-The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus used during the perceptual test.
+The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test.
 Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples for other sources.
@@
 This scale is considered relevant for perception and was already used to compute the masking models proposed in the perceptual experiment results analysis.
-The input frames of dimension 29x8 are processed independantly in this study.\\
+The input frames of dimension 29x8 are processed independently in this study.\\ \ml{cite sensor paper}
 
@@
 
-Training the models is achieved by progressively minimizing a loss function $L(y, \hat y)$, xhere $y$ and $\hat y$ represent the target and predicted outputs respectively.
+Training the models is achieved by progressively minimizing a loss function $L(y, \hat y)$, where $y$ and $\hat y$ represent the target and predicted outputs respectively.
 For sound level estimation we use the MSE or L2 loss:
@@
 \end{equation}
-where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted sound level of source $s$ in dB. For presence prediction the cost function is the binary cross-entropy of equation:
+where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted sound level of source $s$ in dB.
+For the prediction of the presence, the cost function is the binary cross-entropy of equation:
 \begin{equation}
@@
 As a result, the time of presence estimates are precise with a root mean-squared error of 0.12 on a 0-1 scale.
-This behaviour may be explained by the permissiveness of the $\hat T_s(\alpha)$ indicator: if sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values.
-Thus, when using $\hat T_s(\alpha)$ it is sufficient to obtain good presence detection that the model output either very low values when the source is absent or poor estimates of the level when it is present.
-Here, the model fails at the source separation and level estimation task and is probably closer to a detection model.\\
+This behavior may be explained by the fact that $\hat T_s(\alpha)$ indicator is too permissive: if some sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values.
+Thus, when using $\hat T_s(\alpha)$, it is sufficient to obtain good presence detection that the model output either very low values when the source is absent or poor estimates of the level when it is present.
+Here, the model fails at the source separation and level estimation task and is probably closer to a detection model.\\ \ml{Clarifier les deux derineire phrases}
 
-The presence detection model performs similarly well for both presence accuracy (92.11\%) and resulting $\hat T_s(\alpha, \beta)$ estimation (0.12 RMSE).
+The presence detection model performs well for both presence accuracy (92.11\%) and resulting $\hat T_s(\alpha, \beta)$ estimation (0.12 RMSE).
 Both models obtain lower accuracies for birds than for traffic or voices.
 However, the origin of errors varies.
-The sound level model has an extremely low false positive rate consistently across sources while its false negative rate varies from 7.8\% for traffic to more than 18\% for birds.
-The reverse is visible for the presence model which displays a higher false positive rate for birds.
-Overall, traffic is found to be the most easily detected source, which may be due to its location in the lower frequency bands while voice and bird information is usually concentrated in narrower frequency ranges when using third-octave bands.\\
+The sound level model has an very low false positive rate across sources while its false negative rate varies from 7.8\% for traffic to more than 18\% for birds.
+The opposite is observed for the presence model which displays a higher false positive rate for birds.
+Overall, traffic is found to be the source that is most easily detected.
+This may be due to its power spectral density that is concentrated in the lower frequency bands while voice and bird information is usually concentrated in narrower frequency ranges when using third-octave bands.\\
 
@@
 
-Finally, a comparison is made of the models's capacity to result in relevant pleasantness estimates.
-For this analysis the perceptual experiment is used as assessments are available.
+Next, a comparison is made of the models's capacity to result in relevant pleasantness estimates.
+For this analysis, the perceptual experiment is used as assessments are available.
 It is further split in two parts:
@@
 
-Subsequently, a deep learning model with simple, state-of-the-art architecture is developed for this task.
+Subsequently, a deep learning model with state-of-the-art architecture is developed for this task.
 The model takes as input the fast third-octave band representation of the mixed scene and is trained on a corpus of simulated scenes to output presence labels in short time frames of 1s.
-Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean-squared error on a 10-point scale.\\
+Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean-squared error on a 10-point scale which is found satisfying \ml{est que l'on peut comparer a la variance de jugement perceptif ?}.\\
 
-Future work will be mainly focused on the study of the robustness of developed models to new environment and data.
-Here only three sources were considered during training, and a limited isolated sound sources database.
-Thus, pleasantness estimation chain needs to be adapted to the city of Lorient, in which the Cense sensor network is to be implemented, as it contains specific sound environments as well as unique sources that were not considered in this study.
+Future work will focus on the study of the robustness of developed models to new environment and data.
+Here only three sources were considered during training, and a limited isolated sound sources database have been considered.
+Thus, pleasantness estimation chain needs to be evaluated using data gathered in the experimentation site, where the Cense sensor network is to be implemented.
+%, as it contains specific sound environments as well as unique sources that were not considered in this study.
 

### 2019-02-26 report.tex
Corrections and proof read

@@
 
-This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}.
+This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}.
 From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.
@@
 
-With respect to the three discussed objectives, a test corpus is created based on recordings from one of the four soundwalks studied in~\cite{aumond}.
+With respect to the three discussed objectives, a test corpus is created based on recordings from one of the four soundwalks studied in~\cite{aumond2017}.
 The scenes correspond to 19 locations in the 13th district of Paris.
@@
 The first objective requires the comparison of perceptual assessments for recorded and replicated sound scenes.
-The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) with diverse environments are picked to explore different real-life situations, as a function of the classifications proposed in~\cite{jrg}: \textit{parks}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
+The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) with diverse environments are picked to explore different real-life situations, as a function of the classifications proposed in~\cite{gloaguen2017}: \textit{parks}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
 For the second part of the corpus sound scenes are simulated based on the 19 45~s extracts.
@@
 
-The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources.
-This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources.
-For this experiment, the playback sound level is the same as during recordings \ml{of the references scenes ?}, and ranges from 63.9~dB to 79.4~dB.\\
+The 19 scenes are then replicated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources.
+This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources.
+For this experiment, the playback sound level is the same as during recordings of the references scenes, and ranges from 63.9~dB to 79.4~dB.\\
 
@@
 Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
-Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}.
+Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
 These statistics include, conditionally to the ambiance:
@@
 For each environment diverse new scenarios are generated by sampling the background and event properties distributions.
-For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space.
+For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space.
 This is shown in Figure~\ref{fig:tvb_pres}.
@@
 \end{itemize}
-These quantities are typically studied in the perceptual characterization of sound scenes (CITE).
+These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
 While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
 Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters.
-Previous studies (CITE) show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds.
+Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds.
 The perception of source activity can be linked to notions of dominance, time of presence,  or emergence.
@@
 
-Participants evaluate each scene on criteria coded on a discrete scales \ml{how many steps ?}.
+Participants evaluate each scene on criteria coded on 10-point discrete scales.
 Many statistical tools assume that the studied data follows continuous, single mode distributions.
@@
 of its duration.
-To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, \ml{trop long : but as voices have a less distinct influence on pleasantness this is not reflected directly}.
+To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly.
 Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy.
@@
 Overall, most perceptual responses for replicated scenes are close to those for recordings.
-No major difference is found in a particular scale, and the proposed hypothesis, \ml{that is that ...} cannot be invalidated.
+No major difference is found in a particular scale, and the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be invalidated.
 
@@
 79.4\% - 18.1\% and 79.6\% - 15.2\%).
-Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson}.
+Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}.
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
@@
 Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
-The resulting values are consistent with the literature~\cite{aumond, lavandier}, both between general scales as discussed in the previous section and for the relations to source contributions.
+The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions.
 Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
@@
 In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model.
-However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus.
+However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus: the mean standard deviation of pleasantness assessments for this experiment is about 1.77.
 
@@
 
-Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond, gontier, lavandier} to correlate well with content-related perceptual parameters: the overall loudness and the source-specific time of presence.\\
+Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond2017, gontier2018} to correlate well with content-related perceptual parameters: the overall loudness and the source-specific time of presence.\\
 
@@
 Typically this includes indicators derived from sound level measurements.
-For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{ita}:
+For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
 
@@
 \item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
-\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond}, that is found to correlate well with the perceptual time of presence of birds.
+\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond2017}, that is found to correlate well with the perceptual time of presence of birds.
 \end{itemize}
@@
 
-Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed.
+Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier2018} are computed.
 $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:
@@
 First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness.
-Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study \ml{because ??}.\\
+Although any could be used, the $L_{50}$ displays the highest correlation to pleasanntness and is thus selected for pleasantness prediction in this study.\\
 
@@
 
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/ppred.pdf}
+    \caption{Processing chain applied to obtain an estimation of pleasantness in a recorded scene.}\label{fig:ppred}
+\end{figure}
+
 Based on the results outlined by the perceptual experiment, a processing chain is obtained for the estimation of pleasantness in urban sound scenes as shown in Figure~\ref{fig:ppred}.
@@
 
-Predicting these quantities directly \ml{from what ?} using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels.
+Predicting these quantities directly from audio recordings using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels.
 However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}.
@@
 
-In this report, two models for source-specific sound level and presence estimation are implemented and compared. \ml{manque un lien entre les deux derniers par}
+In the following sections, two models for source-specific sound level and presence estimation are implemented and compared.
 
 \subsection{Data}
+\label{sec:pred_data}
 
@@
 
-The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation.
+The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{gontier2017}.
 Thus, the model should take as input the same representation.
@@
 This scale is considered relevant for perception and was already used to compute the masking models proposed in the perceptual experiment results analysis.
-The input frames of dimension 29x8 are processed independently in this study.\\ \ml{cite sensor paper}
+The input frames of dimension 29x8 are processed independently in this study.\\
 
@@
 \subsection{Model}
+\label{sec:pred_mdl}
 
@@
 An additional sigmoid activation is added for the model predicting source presence to obtain outputs in the 0-1 range.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
+    \caption{Architecture of deep learning models used for sound level or source presence prediction.}\label{fig:deep_arch}
+\end{figure}
 
@@
 
-Models are optimized using the Adam algorithm~\cite{adam} on mini-batches of 64 1s examples.
+Models are optimized using the Adam algorithm~\cite{kingma2015} on mini-batches of 64 1s examples.
 For all experiments a constant learning rate of $\lambda = 0.001$ is used.
@@
 The first model outputs very poor estimations of the source-specific sound level, but the resulting source presence accuracy is high at 92.51\%.
-As a result, the time of presence estimates are precise with a root mean-squared error of 0.12 on a 0-1 scale.
+As a result, the time of presence estimates are precise with a root mean square error of 0.12 on a 0-1 scale.
 This behavior may be explained by the fact that $\hat T_s(\alpha)$ indicator is too permissive: if some sources are close in terms of full-band sound level, the threshold $\alpha = -19dB$ allows rougher estimates for binary presence values.
-Thus, when using $\hat T_s(\alpha)$, it is sufficient to obtain good presence detection that the model output either very low values when the source is absent or poor estimates of the level when it is present.
-Here, the model fails at the source separation and level estimation task and is probably closer to a detection model.\\ \ml{Clarifier les deux derineire phrases}
+Thus, because the sound level model output is used in $\hat T_s(\alpha)$, there are two cases for which the correct presence is obtained
+\begin{enumerate}
+\item The model outputs very low level values for sources that are not present
+\item The model outputs poor estimates but with sufficiently close values for all present sources
+\end{enumerate}
+Thus, the model may fail at the source separation and level estimation task and is here probably closer to a detection model.\\
 
@@
 The model takes as input the fast third-octave band representation of the mixed scene and is trained on a corpus of simulated scenes to output presence labels in short time frames of 1s.
-Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean-squared error on a 10-point scale which is found satisfying \ml{est que l'on peut comparer a la variance de jugement perceptif ?}.\\
+Low pleasantness prediction errors are achieved on the perceptual experiment corpus: around 1.15 root mean square error on the 10-point scale which is below the mean standard deviation of pleasantness assessments of 1.77.\\
 
@@
 Thus, pleasantness estimation chain needs to be evaluated using data gathered in the experimentation site, where the Cense sensor network is to be implemented.
-%, as it contains specific sound environments as well as unique sources that were not considered in this study.
-
-\clearpage
-
-\textbf{TODO}
-
-\textbf{Text}
-
-\begin{itemize}
-\item Appendix A
-\end{itemize}
-
-\textbf{Figures}
-
-\begin{itemize}
-\item Pleasantness prediction chain
-\item Deep learning architecture
-\item Headphones calibration
-\end{itemize}
 
@@
 
-\begin{thebibliography}{9}
-%\bibitem{axelsson2010}
-%Axelsson, \"O.; Nilsson, M.E.; Berglund, B.
-%A principal components model of soundscape perception. \textit{J. Ac. Soc. Am.}, 2010
-%
-%\bibitem{jeon2018}
-%Jeon, J.Y.; Hong, J.Y.; Lavandier, C; Lafon, J.; Axelsson, \"O.; Hurtig, M.
-%A cross-national comparison in assessment of urban park soundscapes in France, Korea, and Sweden through laboratory experiments. \textit{Applied Acoustics}, 2018
-%
-%
-%\bibitem{aumond}
-%Aumond, P.; Can, A.; De Coensel, B.; Botteldooren, D.; Ribeiro, C.; Lavandier, C.
-%Modeling soundscape pleasantness using perceptual assessments and acoustic measurements along paths in urban context. \textit{Acta Acustica united with Acustica}, 2017
-%
-%\bibitem{ita}
-%Berzborn, M.; Bomhardt, R.; Klein, J.; Richter, J.-G.; Vorlander, M.
-%The ITA-Toolbox: An Open Source MATLAB Toolbox for Acoustic Measurements and Signal Processing. \textit{In proceedings of the 43th Annual German Congress on Acoustics}, 2017
-%
-%\bibitem{simscene}
-%Rossignol, M.; Lafay, G.; Lagrange, M.; Misdariis, N.
-%SimScene : a web-based acoustic scenes simulator, \textit{In proceedings of the 1st Web Audio Conference (WAC)}, 2015
-%
-%\bibitem{gloaguen}
-%Gloaguen, J.-R.; Can, A.; Lagrange, M.; Petiot, J.-F.
-%Creation of a corpus of realistic urban sound scenes with controlled acoustic properties.
-\textit{In proceedings of the 173rd Meeting of the Acoustical Society of America and the 8th Forum Acusticum (Acoustics'17)}, 2017
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{refs}
 

### 2019-02-26 report.tex
Proof read

@@
 This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}.
-From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.
+From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of pleasantness in urban sound environments.
 
@@
 This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources.
-For this experiment, the playback sound level is the same as during recordings of the references scenes, and ranges from 63.9~dB to 79.4~dB.\\
+For this experiment, the playback sound level is the same as during recordings of the reference scenes, and ranges from 63.9~dB to 79.4~dB.\\
 

### 2019-03-21 report.tex
Final version

@@
 \begin{document}
-\author{F\'elix GONTIER \\ \texttt{felix.gontier@ls2n.fr}}
-\title{CENSE - Perceptual test II}
+\author{F\'elix Gontier, Catherine Lavandier, Pierre Aumond, Mathieu Lagrange, Jean-Francois Petiot}
+\title{Project CENSE - WP5 Advancement Report}
 \maketitle
@@
 This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}.
-From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of pleasantness in urban sound environments.
+From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of pleasantness in urban sound environments using corpora of simulated scenes.\\
 
 \section{Test description}
+\label{sec:test}
 \subsection{Objectives}
@@
 
-\begin{figure}[!h]
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.6\textwidth]{figures/tvb_pres.eps}
-    \caption{Estimated time of presence space for simulated scenes.}\label{fig:tvb_pres}
+     \begin{subfigure}[t]{0.33\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
+    \end{subfigure}%
+    \begin{subfigure}[t]{0.33\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.33\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
+    \end{subfigure}
+    \caption{Estimated time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
 \end{figure}
@@
 
-The experiment corpus totals 100 scenes.
+The experiment corpus totals 100 scenes, including 6 recorded, 19 replicated and 75 simulated extracts.
 
@@
 \subsection{Questions}
+\label{sec:test_questions}
 
@@
 \begin{itemize}
-\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
-\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
-\item Overall loudness (\textit{Quiet - Noisy}) - OL
-\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
-\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
+\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
+\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
+\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
+\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
+\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
 \end{itemize}
@@
 \begin{itemize}
-\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
-\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
+\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
+\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
 \end{itemize}
@@
 \subsection{Test conditions}
+\label{sec:test_conditions}
 
@@
 
-A total of 23 students at Ecole Centrale de Nantes completed the test.
+A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment.
 
 \section{Perceptual analysis}
+\label{sec:perc}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/subj_gr.eps}
+    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
     \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
@@
 \subsection{Correspondence between recorded and replicated scenes}
+\label{sec:perc_recrep}
 
@@
 Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold.
-This non parametric test is chosen because it assumes that distributions are continuous.
 Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices.
@@
 \subsection{Perceptual space}
+\label{sec:perc_space}
 
@@
 Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots.
-Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by black segments.
-The space covered by scenes based on new scenarios covers that of the studied real-life environments, demonstrating the efficiency of the scene generation procedure in terms of diversity.
+Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by black arrows.
+The standard deviations of responses are shown as ellipses in Appendix~\ref{sec:app_a}, which overlap in most cases confirming the equivalence discussed in Section~\ref{sec:perc_recrep}.
+The space covered by scenes based on new scenarios (red dots in Figure~\ref{fig:pspace}b) covers that of the studied real-life environments (blue dots), demonstrating the efficiency of the scene generation procedure in terms of diversity.
 
@@
 \subsection{Perceptual scales relationships}
-\label{sec:percm}
+\label{sec:perc_mdls}
 
-\begin{table}
+We then study the relations between perceptual scales with respect to existing work.
+Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
+The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions.
+Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
+In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
+This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
+Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\
+
+Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds.
+The perceptual assessment of the sound level of passing vehicles $L_{T, p}$ is a potential alternative to that of the time of presence of traffic $T_{T, p}$.
+Except for the interest (I) it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
+
+\begin{table}[t]
 \centering
@@
 
-We then study the relations between perceptual scales with respect to existing work.
-Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
-The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions.
-Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
-In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
-This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
-Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\
-
-Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds.
-The perceptual assessment of the sound level of passing vehicles $L_{T, p}$ is a potential alternative to that of the time of presence of traffic $T_{T, p}$.
-Except for the interest (I) it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
-
 To further compare this experiment's results to previous work, multilinear models of pleasantness are built as a function of both the overall loudness (OL) source-specific parameters.
@@
 \end{align}
-
 Both models have very similar formal expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
@@
 \hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
-\end{equation}.
+\end{equation}
 Though, in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness.
@@
 In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model.
-However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus: the mean standard deviation of pleasantness assessments for this experiment is about 1.77.
+However, the error stays within acceptable range as it is below the mean standard deviation of pleasantness assessments of about 1.77 for this experiment.
 
-\begin{table}[t]
+\begin{table}[h]
 \centering
@@
 \section{Physical analysis}
+\label{sec:phys}
 
@@
 \subsection{Indicators}
+\label{sec:phys_inds}
 
@@
 \item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
-\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond2017}, that is found to correlate well with the perceptual time of presence of birds.
 \end{itemize}
 
+The time and frequency second derivative for the 4kHz band with fast measurements ($TFSD_{4kHz, 1/8s}$) introduced by~\cite{aumond2017} is also computed as it represents well with the perceptual time of presence of birds.
 Furthermore, the simulation process outputs ground truth source contributions in the form of separate channels.
@@
 \end{align}
-
 where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments.
@@
 \end{align}
-
 The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
@@
 \subsection{Relation between perceptual an physical data}
-\label{sec:physm}
+\label{sec:phys_mdls}
 
@@
 	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\ \hline
-	$TFSD_{4kHz}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
+	$TFSD_{4kHz, 1/8s}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
 	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
@@
 Subsequently, multilinear models are constructed based on the results of the analysis of correlations.
-As in Section~\ref{sec:percm} a VIF check is performed on predictors to ensure that no multi-collinearity exists.
+As in Section~\ref{sec:perc_mdls} a VIF check is performed on predictors to ensure that no multi-collinearity exists.
 The best pleasantness model optimized on the present data is:
@@
 \end{itemize}
-It is also possible to use the models discussed in Section~\ref{sec:percm} as is.
+It is also possible to use the models discussed in Section~\ref{sec:perc_mdls} as is.
 In this situation, instead of predicting pleasantness directly, each intermediate perceptual parameter is modeled as a function of available physical variables.
@@
 It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness.
-This is particularly important in the case of voices, found in multiple previous studies to have a positive effect.
+This is particularly important in the case of voices, as their impact depends on their content and expressivity.
+In this work recordings of read books are used, yielding the observed neutral pleasantness contribution.
 Conversely, the $\hat P_{\varphi 2}$ model uses a perceptual model constructed on a much larger corpus of recorded scenes, and is less prone to these specificity properties.
@@
 \section{Physical indicators prediction}
+\label{sec:pred}
 
 \subsection{Objective}
+\label{sec:pred_obj}
 
@@
 \subsection{Choice of indicators}
+\label{sec:pred_inds}
 
@@
 Predicting these quantities directly from audio recordings using deep learning architectures is difficult: training this type of models requires large annotated datasets which would require extensive work by several humans to obtain relevant labels.
-However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:physm}.
+However, the perceived time of presence can be approximated by a simple masking model as demonstrated in Section~\ref{sec:phys_mdls}.
 While the same need for annotation of large datasets exists, it can be circumvented by the use of simulated data, for which the contribution of each source in the mix is available.
@@
 \subsection{Training details}
+\label{sec:pred_train}
 
@@
 \subsection{Metrics}
+\label{sec:pred_metrics}
 
@@
 \subsection{Results}
+\label{sec:pred_res}
 
@@
 This is explained by the uncertainties of the $\hat T_s(\alpha, \beta)$ indicator that result from the simplicity of its masking model.
-The predictions from both models' outputs are in fact closer to the performance of the $\hat Pp3$ perceptual model presented in Section~\ref{sec:percm} (equation (3)).\\
+The predictions from both models' outputs are in fact closer to the performance of the $\hat Pp3$ perceptual model presented in Section~\ref{sec:perc_mdls} (equation (3)).\\
 
@@
 \section{Conclusion}
+\label{sec:ccl}
 
@@
 
+\clearpage
+\appendix
+\section{Comparison of PCA plots for recorded and replicated scenes}
+\label{sec:app_a}
+
+\begin{figure}[h]
+    \centering
+     \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
+    \end{subfigure}%
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p3.eps}
+    \end{subfigure}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p4.eps}
+    \end{subfigure}%
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p8.eps}
+    \end{subfigure}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+     \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p15.eps}
+    \end{subfigure}%
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/pca_p18.eps}
+    \end{subfigure}
+\end{figure}
+
 \end{document}

### 2019-03-25 paper/paperDraft.tex
Paper V1

\documentclass[11pt,a4paper]{article}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$\\
$^2$\\
$^3$
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}
\label{sec:intro}
%- Du bruit au soundscape
%- La pleasantness premier axe sur les affects liés au soundscape
%- On a des modèles de pleasantness perceptifs qui font intervenir des temps de présence des sources
%- Peut-on à partir de la mesure physique faire des modèles de pleasantness sur la base d'indicateurs physiques?
%- Idée: si l'on fait de la reconnaissance de sources, on peut faire un meilleur modèle physique que les précédents
%- Première étape: On vérifie que le corpus créé est proche d'un corpus réel
%- Sur le corpus on fait de l'apprentissage profond
%- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)

\section{Methods}
\label{sec:methods}

\subsection{Perceptual validation}
\label{sec:pval}

\subsubsection{Sound scenes corpus}
\label{sec:pval_corp}

A first corpus is created to validate the use of simulated sound scenes in the development of pleasantness prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception. The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes. The recordings are first annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and are associated with a sound level.
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

Sound scenes are then simulated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\

\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Estimated time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
\end{figure}

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
\end{itemize}

Adjustments are made on the variance of the considered properties to better cover possible scenarios, and the EBR of voice events, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions. For each scene the time of presence for the traffic, voice and bird sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context shown in Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/lvls.jpg}
    \caption{Example of sound level scale a function of the environment for new simulated scenes. TODO CITE}\label{fig:amb_levels}
\end{figure}

\subsubsection{Experiment}
\label{sec:pval_exp}

A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point (0-10) scales. These scales are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\

Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}

Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. A random pre-generated listening order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15) to calibrate their answers. For the remainder of the corpus all participants experience different scene ordering. Participants can only listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

The scenes are played at a given sound level as discussed in Section~\ref{sec:pval_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber, resulting in a relation between playback Leq and LAeq and sound card output voltage for pink noise. This ensures that correctly scaled scenes will be heard at the desired sound level by listeners.\\

A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment.

\subsection{Indicators for pleasantness prediction}
\label{sec:inds}

Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond2017, gontier2018, ricciardi2014} to correlate well with perceptual parameters.\\

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 1~s (slow) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model in~\cite{ricciardi2014}.
\end{itemize}

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. To provide a baseline for pleasantness prediction the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\

The simulation process outputs ground truth source contributions in the form of separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

\begin{itemize}
\item $L_{eq, s}$: Equivalent sound level for source $s$
\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined
\end{itemize}

Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed. $\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta_s(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$. It is parametrized by $\alpha$ and $\beta$ thresholds:

\begin{align}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
\end{align}

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\

In the remainder of this paper, the subscript $s$ is replaced with the corresponding source initial: $T$ for traffic, $V$ for voices and $B$ for birds.

\subsection{Indicator prediction through deep learning}
\label{sec:deep}

\subsubsection{Learning corpus}
\label{sec:deep_corp}

To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
\begin{itemize}
\item The development set containing 400 45~s scenes (total 5 hours) which is used during the training process,
\item The evaluation set containing 200 45~s scenes (total 2.5 hours) which is used to compute the performance of the trained model and its generalization capabilities.
\end{itemize}

The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:pval_corp}. Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples and recording annotations for other sources. The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\

The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}. Thus, a prediction model should take as input the same representation. Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments.

\subsubsection{Model architecture}
\label{sec:deep_mdl}

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by LeakyReLU activations. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
\end{figure}

When using the model's outputs to estimate the time of presence of sources, a threshold of 0.5 is used to obtain binary presence values.

\section{Results}
\label{sec:results}

\subsection{Perceptual validation}
\label{sec:perc}
%=> 3.1, 3.2

Participants evaluate each scene on criteria coded on 11-point discrete scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

The perceptual responses of the 6 recorded and corresponding replicated scenes are then compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Discrepancies can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes is found for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\

\begin{table}[h]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Kolmogorov-Smirnov test are in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & -1.04 & 0.43 & 0.13 & -0.91 & 0.39 & -2.09 & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & \textbf{0.91} & 0 & \textbf{-1.83} & 0.48 & 1.30 & -0.96 & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & -0.91 & 0.09 & \textbf{-1.43} \\
	P15 & -1.35 & 0.52 & 0.52 & -1.17 & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & 1.13 & -0.30 & -1.17 & -0.43 & 1.39 & -1.04 & -1.83 & 0.83 & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table}

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the efficiency of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by arrows, and the standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

% FIGURE
\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.55\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
    \end{subfigure}
    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252). Each dot represent the projection of one scene on the resulting space. The components space is distorted compared to the litterature.}\label{fig:pspace}
\end{figure}

\subsection{Baseline pleasantness models}
\label{sec:base}
%=> 3.3, 4.2

\subsubsection{Perceptual models}
\label{sec:base_perc}

\begin{table}[h]
\centering
\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
\label{tab:percc}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
\end{tabular}
%}
\end{table}

We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds. The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$. Except for the interest (I) it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\

As a baseline for pleasantness prediction, multilinear models are built as a function of both the overall loudness (OL) and source-specific parameters. To ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to model regression. Only models for which all predictors verify $VIF<5$ are considered valid. The best resulting models with statistically significant estimates ($p<0.05$) for all predictors are:

\begin{align}
\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p}\\
\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p}
\end{align}

Both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity. A small positive contribution of the time of presence of voices is also found despite no significant correlation.\\

Table~\ref{tab:percm} summarizes performance metrics computed for the two models. $\hat P_{p1}$ and $\hat P_{p2}$ display similar estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. The mean standard deviation of pleasantness assessments for this experiment is about 1.77, thus perceptual models are considered accurate with about 0.6 RMSE on the 10-point scale.

\begin{table}[t]
\centering
\caption{Performance of perceptual models for pleasantness prediction (**: p<0.01).}
\label{tab:percm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{p1}$ & 0.59 & 0.91 & 0.95** \\
	$\hat P_{p2}$ & 0.61 & 0.90 & 0.95** \\ \hline
\end{tabular}
%}
\end{table}

\subsubsection{Physical models}
\label{sec:base_phys}

The analysis of physical indicators is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. Two scenes containing only one source are excluded for implementation concerns, resulting in $n=92$ scenes.\\

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source. Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weaker for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha, \beta)$ show strong correlations to their perceptual counterparts (r>0.8), along with good source discrimination properties. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity. Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$. In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.\\

\begin{table}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\
	$L_{10}-L_{90}$ & NS & NS & -0.24* & NS & -0.22* & NS & NS & NS & NS \\ \hline
	$TFSD_{500Hz, 1s}$ & NS & 0.41** & NS & 0.28** & NS & -0.24* & -0.39** & 0.74** & NS \\
	$TFSD_{4kHz, 1/8s}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha, \beta)$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
	$\hat T_V(\alpha, \beta)$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
	$\hat T_B(\alpha, \beta)$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
\end{tabular}
}
\end{table}

Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:base_perc} a VIF check is performed on predictors to ensure that no multi-collinearity exists. On the present corpus, the best model is:
\begin{equation}
\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta)
\end{equation}
Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:base_perc}. This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively, and no strong contribution of voices to pleasantness is identified in this corpus.\\

This model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal. Both models' coefficients are optimized on the studied data:
\begin{align}
\hat P_{\varphi 2} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
\hat P_{\varphi 3} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} + 5.00~TFSD_{4kHz, 1/8s}
\end{align}

Table~\ref{tab:physm} compares the performance of the three models. Expectedly, $\hat P_{\varphi 1}$ outperforms both baselines. Though, compared to the best perceptual models its explained variance is about 9\% lower. Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 10-point scale. In $\hat P_{\varphi 2}$ introducing the $L_{10}-L_{90}$ emergence has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\

\begin{table}[t]
\centering
\caption{Performance of physical models for pleasantness prediction (**: p<0.01).}
\label{tab:physm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{\varphi 1}$ & 0.83 & 0.82 & 0.91** \\
	$\hat P_{\varphi 2}$ & 0.90 & 0.79 & 0.89** \\
	$\hat P_{\varphi 3}$ & 0.90 & 0.78 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

\subsection{Prediction of pleasantness through deep learning}
\label{sec:pred}

Table~\ref{tab:perf_cmp} summarizes the performance of the source detection deep learning architecture presented in Section~\ref{sec:deep} on the evaluation dataset. The overall presence detection accuracy is 92.11\%. The model performs similarly well for the three sources, ranging from 90.08\% for birds to 94.09\% for traffic. Regarding the type of errors, they are equally split between false positives and false negatives. Though, these rates vary depending on the source. Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%. This is expected given the spectral content related to these sources. Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model. The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.\\

\begin{table}[t]
\centering
\caption{Performance of the model predicting source presence with $\hat T_s(\alpha, \beta)$ labels. Presence metrics are computed for n=8600 1s frames and time of presence metrics on n=200 45s scenes. (TP: true positive, TN: true negative, FP: false positive, FN: false negative)}
\label{tab:perf_cmp}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c c c }
\hline
	 & All sources & Traffic & Voices & Birds \\ \hline
	Presence accuracy (\%) &  92.11 & 94.09 & 92.15 & 90.08 \\
	Presence TP (\%) & 91.96 & 92.70 & 89.93 & 93.37 \\
	Presence TN (\%) & 92.30 & 97.54 & 94.76 & 87.70 \\
	Presenct FP (\%) & 7.70 & 2.46 & 5.24 & 12.30 \\
	Presence FN (\%) & 8.04 & 7.30 & 10.17 & 6.63 \\ \hline
	$\hat T_s(\alpha, \beta)$ RMSE & 0.12 & 0.13 & 0.10 & 0.11 \\
	$\hat T_s(\alpha, \beta)$ MAE & 0.06 & 0.05 & 0.06 & 0.08 \\ \hline
\end{tabular}
%}
\end{table}

Next, the capacity of these predictions to result in relevant pleasantness estimates is studied. For this analysis only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data. This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

\begin{table}[t]
\centering
\caption{Pleasantness prediction quality using the source detection model compared to labels. The experiment corpus is split into respectively the 25 recorded and replicated scenes and 75 simulated scenes.}
\label{tab:pppred}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c | c c c }
\hline
	Presence & \multicolumn{3}{|c}{Model output} & \multicolumn{3}{|c}{$\hat T_s(\alpha, \beta)$ labels} \\ \hline
	Sub-corpus & All & 1 & 2 & All & 1 & 2 \\ \hline
	RMSE & 0.84 & 0.80 & 0.85 & 0.83 & 0.73 & 0.85 \\ \hline
	r & 0.91** & 0.92** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
	$R^2_{adj}$ & 0.82 & - & - & 0.82 & - & - \\ \hline
\end{tabular}
%}
\end{table}

Pleasantness prediction using the detection model is as performant as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. Though, errors are significantly higher than for perceptual models in Section~\ref{sec:base_perc}.\\

\section{Discussion}
\label{sec:discussion}

The perceptual models of pleasantness models are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
\end{align}
Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result. First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\

The physical model on this corpus includes only the contribution of bird sources. This is attributed to the overall low diversity of the studied corpus:
\begin{itemize}
\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process. Quieter environments such as parks and quiet streets are less likely to contain continuous traffic, while high sound levels in busy streets are always attributed to traffic. Including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
\item The diversity of available isolated source extracts for scene simulation is low. Particularly, voice events are exclusively recordings of read english with low variations in speaker and consistently neutral expressiveness. These properties result in no significant contribution of voice events to pleasantness for this corpus.
\end{itemize}
Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness. The trained deep learning architecture already displays few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\

Results show that pleasantness estimates using the developped source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Because the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered "weak" labels. Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\\

The detection-based pleasantness predictions are worse on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy. This is expected as the architecture was never trained on complex scenes. Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes. The detection model is thus quite robust to scenes with additional sources as perturbators. A full study with a larger evaluation corpus would be needed to further validate the model's robustness to diverse real-life scenarios.

\section{Conclusion}
\label{sec:conclusion}

A perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual characterization. Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.\\

Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction in urban environments, leads to better estimates than previous models from acoustic indicators.\\

Future work will focus on the robustness of the source detection model, which heavily depends on the quantity and quality of data available for scene simulation. Furthermore, this methodology can be adapted to specific environments by extending or refining the source taxonomy in both the scene simulation process and detection model. For example, additional transport-related sources such as train and plane activity could be introduced, and a difference could be made depending on voice expressivity or bird species.

\section*{Acknowledgements}
\label{sec:ack}

This research is funded by the French National Agency for Research, under the CENSE project (convention ANR-16-CE22-0012)

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2019-03-26 paper/paperDraft.tex
relecture rapide

@@
 A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
-For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception.
+For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception.
 The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes.
@@
 Sound scenes are then simulated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources.
-This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources.
+This database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
 The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
@@
 
-Adjustments are made on the variance of the considered properties to better cover possible scenarios, and the EBR of voice events, mostly read English recordings, are reduced to improve realism.
+Adjustments are made on the variance of the considered properties to better cover possible scenarios.
+The EBR of voice events, mostly read English recordings, are reduced to improve realism.
 An additional \textit{place} ambiance is added with properties derived empirically from available data.
@@
 
-The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.
+The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project \ml{a citer si pas fait avant}, and 75 simulated scenes.
 
@@
 
-A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment.
+A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.
 
@@
 \item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
-\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model in~\cite{ricciardi2014}.
+\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
 \end{itemize}
@@
 
-The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}.
-Thus, a prediction model should take as input the same representation.
-Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes.
+\ml{ Je prendrais un point de vue plus general, tout le monde fait alors on utilise ca comme ca ca montre la capacité de generalisation de l'approche // The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}.
+Thus, a prediction model should take as input the same representation.} Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes.
 Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception.
@@
 Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments.
+\ml{les resultats seront discutés ensuite ou non ?
+Sinon le dire ici}
 
@@
 A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
-The model includes 4 blocks of convolutional layers followed by LeakyReLU activations.
+The model includes 4 blocks of convolutional layers followed by LeakyReLU \ml{un peu technique, utiliser les terms complets et citer} activations.
 The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
@@
 Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75).
-The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora.
+The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora.
 For both sets the variance explained by the first two components is similar (resp.
@@
 Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots.
-The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the efficiency of the scene generation procedure in terms of diversity.\\
+The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\
 
-Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by arrows, and the standard deviation of assessments is represented by ellipses for one scene.
+Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace} using arrows.
+The standard deviation of assessments is represented by ellipses for one scene.
 The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity.
@@
 Each dot represent the projection of one scene on the resulting space.
-The components space is distorted compared to the litterature.}\label{fig:pspace}
+The components space is distorted compared to the literature.}\label{fig:pspace}
 \end{figure}
@@
 
-Next, the capacity of these predictions to result in relevant pleasantness estimates is studied.
-For this analysis only the perceptual experiment is used as no assessments are available on the evaluation dataset.
+Next, the capacity of these predictions to give relevant pleasantness estimates is studied.
+For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset.
 To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
@@
 The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}.
-First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data.
-This performance of the detection model is expected given the low errors on time of presence estimates.
+\ml{Je ne comprends pas : First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data.} This performance of the detection model is expected given the low errors on time of presence estimates.
 Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training.
@@
 
-Pleasantness prediction using the detection model is as performant as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-Though, errors are significantly higher than for perceptual models in Section~\ref{sec:base_perc}.\\
+Pleasantness prediction using the detection model is as effective as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
+Though, errors are significantly higher than the one of perceptual models in Section~\ref{sec:base_perc}.\\
 
@@
 
-The perceptual models of pleasantness models are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
+The perceptual models of pleasantness \ml{models} are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
 \begin{align}
@@
 \end{align}
-Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result.
-First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment.
+\ml{Enlever ou reformuler: Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result.} First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment.
 However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models.
@@
 
-The physical model on this corpus includes only the contribution of bird sources. This is attributed to the overall low diversity of the studied corpus:
+The physical model on this corpus includes only the contribution of bird sources. This result is attributed to the overall low diversity of the studied corpus:
 \begin{itemize}
 \item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
-Quieter environments such as parks and quiet streets are less likely to contain continuous traffic, while high sound levels in busy streets are always attributed to traffic.
-Including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
-\item The diversity of available isolated source extracts for scene simulation is low.
-Particularly, voice events are exclusively recordings of read english with low variations in speaker and consistently neutral expressiveness.
+Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic.
+It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
+\item The diversity of available isolated source extracts for scene simulation is rather low.
+Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness.
 These properties result in no significant contribution of voice events to pleasantness for this corpus.
@@
 
-Results show that pleasantness estimates using the developped source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus.
+Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus.
 This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
-Because the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered "weak" labels.
-Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\\
+Since the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered as "weak" labels.
+Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\ml{faut faire attention au wording, je n'utiliserai pas bon et mauvais, mais plus le fait que les labels sont plus ou moins haut niveau ?}\\
 
-The detection-based pleasantness predictions are worse on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy.
-This is expected as the architecture was never trained on complex scenes.
+The detection-based pleasantness predictions are of lower quality on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy.
+This is expected as the architecture is not trained on complex scenes.
 Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes.

### 2019-03-27 paper/paperDraft.tex
Add intro

@@
 %- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)
+In urban areas, the decreasing quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
+Consequently, the 2002/49/CE European directive requires the development of noise maps in large cities.
+These maps aim at providing useful tools for urban planning and noise reduction purposes.
+As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1 (CITE) norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
+
+The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied (CITE).
+Specifically, the axis of pleasantness is increasingly associated with soundscape quality in the recent literature (CITE).
+Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments (CITE).
+While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness: traffic, human voices and birds.
+In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
+
+The perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the construction of perceptually motivated noise maps.
+With the advent of the Internet of Things and the availability of low-cost sensor networks (CITE), predicting pleasantness from acoustic measurements is an interesting and ongoing area of research as an alternative to systematic human quality assessment.
+Ricciardi et al.
+(CITE) present a physical model of pleasantness using only variations of the measured sound level.
+Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
+Aumond et al.
+(CITE) explore new indicators based on spectral variations in recordings of sound scenes.
+These indicators are designed to represent the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
+Following the recent advances in machine learning, it may be possible to better approximate the perception of sound sources contributing to pleasantness by developing new indicators relying on source recognition models.
+However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings.
+To our knowledge such databases do not currently exist in the literature.
+A possible alternative is the use of simulated corpora (CITE) where sound scenes are composed from semi-randomly generated annotations.\\
+
+The aim of this research is two-fold.
+First, a perceptual test is conducted where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction.
+Next, a deep learning architecture is designed and trained that performs multi-label sound source recognition.
+Estimations of pleasantness using these predictions are compared to existing models from physical indicators.
 
@@
 
-The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project \ml{a citer si pas fait avant}, and 75 simulated scenes.
+The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.
 
@@
 
-\ml{ Je prendrais un point de vue plus general, tout le monde fait alors on utilise ca comme ca ca montre la capacité de generalisation de l'approche // The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}.
-Thus, a prediction model should take as input the same representation.} Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes.
+The developped model should extract relevant source information from a representation of the audio signal.
+Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal.
+Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}.
+Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range.
 Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception.
@@
 Alternatively, it is possible to perform sound level estimation using ground truth labels.
-Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments.
-\ml{les resultats seront discutés ensuite ou non ?
-Sinon le dire ici}
+Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.
 
@@
 A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
-The model includes 4 blocks of convolutional layers followed by LeakyReLU \ml{un peu technique, utiliser les terms complets et citer} activations.
+The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$ .
 The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
@@
 The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}.
-\ml{Je ne comprends pas : First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data.} This performance of the detection model is expected given the low errors on time of presence estimates.
+First, pleasantness estimations are equally performant using the model's predictions or the ground truth labels , with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus.
+This performance of the detection model is expected given the low errors on time of presence estimates.
 Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training.
@@
 
-The perceptual models of pleasantness \ml{models} are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
+The perceptual models of pleasantness are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
 \begin{align}
@@
 \end{align}
-\ml{Enlever ou reformuler: Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result.} First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment.
+Several hypotheses can be drawn from this result.
+First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment.
 However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models.
@@
 This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
-Since the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered as "weak" labels.
-Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\ml{faut faire attention au wording, je n'utiliserai pas bon et mauvais, mais plus le fait que les labels sont plus ou moins haut niveau ?}\\
+Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
+Thus, in some cases the deep learning model's predictions are considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\
 

### 2019-03-28 paper/paperDraft.tex
input for intro

@@
 %- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)
-In urban areas, the decreasing quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
+In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
 Consequently, the 2002/49/CE European directive requires the development of noise maps in large cities.
@@
 
-The perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the construction of perceptually motivated noise maps.
-With the advent of the Internet of Things and the availability of low-cost sensor networks (CITE), predicting pleasantness from acoustic measurements is an interesting and ongoing area of research as an alternative to systematic human quality assessment.
+Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
+With the advent of the Internet of Things and the availability of low-cost sensor networks (CITE), it may be feasible to predict pleasantness or other perceptual factors from acoustic measurements.
+
+Bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is an interesting and ongoing area of research.
 Ricciardi et al.
@@
 (CITE) explore new indicators based on spectral variations in recordings of sound scenes.
-These indicators are designed to represent the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
-Following the recent advances in machine learning, it may be possible to better approximate the perception of sound sources contributing to pleasantness by developing new indicators relying on source recognition models.
-However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings.
-To our knowledge such databases do not currently exist in the literature.
-A possible alternative is the use of simulated corpora (CITE) where sound scenes are composed from semi-randomly generated annotations.\\
+These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
 
-The aim of this research is two-fold.
+In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.
+In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques CITE.
+However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
+In order to circumvent this issue, we propose in this paper to study the potential of the use of simulated corpora (CITE) \ml{j'enleverai c'est super ambigu, on a l'impression que l'on annote au hasard :) where sound scenes are composed from semi-randomly generated annotations}.
+
+The contributions of this paper are thus three-fold:
+\begin{enumerate}
+  \item provide controled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception
+  \item propose numerical means for predicting time of presence of sound sources from raw acoustic data
+  \item provide a comparative study of the behavior of the proposed approach with state of the art ones.
+\end{enumerate}
+
 First, a perceptual test is conducted where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction.
-Next, a deep learning architecture is designed and trained that performs multi-label sound source recognition.
-Estimations of pleasantness using these predictions are compared to existing models from physical indicators.
+Next, a deep learning architecture is designed and trained that performs multi-label sound source recognition at relevant time scales.
+Estimations of pleasantness using these predictions are next compared to existing models from physical indicators.
+
+\ml{mettre les noms de section dans ce par}
 

### 2019-04-03 paper/paperDraft.tex
input for intro
Add cites + proofread

@@
 \date{
-$^1$\\
-$^2$\\
-$^3$
+$^1$ LS2N, UMR 6004, Ecole Centrale de Nantes, CNRS, F-44300, France\\
+$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
+$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
 }
@@
 \label{sec:intro}
-%- Du bruit au soundscape
-%- La pleasantness premier axe sur les affects liés au soundscape
-%- On a des modèles de pleasantness perceptifs qui font intervenir des temps de présence des sources
-%- Peut-on à partir de la mesure physique faire des modèles de pleasantness sur la base d'indicateurs physiques?
-%- Idée: si l'on fait de la reconnaissance de sources, on peut faire un meilleur modèle physique que les précédents
-%- Première étape: On vérifie que le corpus créé est proche d'un corpus réel
-%- Sur le corpus on fait de l'apprentissage profond
-%- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)
+
 In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
-Consequently, the 2002/49/CE European directive requires the development of noise maps in large cities.
+Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities.
 These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1 (CITE) norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
+As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
 
-The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied (CITE).
-Specifically, the axis of pleasantness is increasingly associated with soundscape quality in the recent literature (CITE).
-Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments (CITE).
-While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness: traffic, human voices and birds.
+The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
+Specifically, the axis of pleasantness is found as the most explanatory of subjective assessments, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
+Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments.
+While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
 In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
@@
 Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
-With the advent of the Internet of Things and the availability of low-cost sensor networks (CITE), it may be feasible to predict pleasantness or other perceptual factors from acoustic measurements.
-
+With the advent of the Internet of Things and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017}, it may be feasible to predict pleasantness or other perceptual factors from acoustic measurements.
 Bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is an interesting and ongoing area of research.
-Ricciardi et al.
-(CITE) present a physical model of pleasantness using only variations of the measured sound level.
+Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
 Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
-Aumond et al.
-(CITE) explore new indicators based on spectral variations in recordings of sound scenes.
-These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
+Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes.
+These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.\\
 
 In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.
-In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques CITE.
+In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
 However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
-In order to circumvent this issue, we propose in this paper to study the potential of the use of simulated corpora (CITE) \ml{j'enleverai c'est super ambigu, on a l'impression que l'on annote au hasard :) where sound scenes are composed from semi-randomly generated annotations}.
+In order to circumvent this issue, we propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017}.\\
 
@@
 \begin{enumerate}
-  \item provide controled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception
-  \item propose numerical means for predicting time of presence of sound sources from raw acoustic data
-  \item provide a comparative study of the behavior of the proposed approach with state of the art ones.
+  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception
+  \item Propose numerical means for predicting the time of presence of sound sources from raw acoustic data
+  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
 \end{enumerate}
 
-First, a perceptual test is conducted where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction.
-Next, a deep learning architecture is designed and trained that performs multi-label sound source recognition at relevant time scales.
-Estimations of pleasantness using these predictions are next compared to existing models from physical indicators.
+First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction.
+Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
+Estimations of pleasantness using these predictions are next compared to existing models from physical indicators in Section~\ref{sec:results}.
 
-\ml{mettre les noms de section dans ce par}
+\section{Data}
+\label{sec:data}
 
-\section{Methods}
-\label{sec:methods}
+\subsection{Corpus generation}
+\label{sec:data_corp}
 
-\subsection{Perceptual validation}
-\label{sec:pval}
-
-\subsubsection{Sound scenes corpus}
-\label{sec:pval_corp}
-
-A first corpus is created to validate the use of simulated sound scenes in the development of pleasantness prediction models.
+A corpus is first created to validate the use of simulated sound scenes in the development of pleasantness prediction models.
 It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris.
@@
 Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
-Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
+Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
 Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
@@
 This is shown in Figure~\ref{fig:tvb_pres}.
-The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context shown in Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\
+The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB to 77.1~dB.\\
 
@@
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=0.8\textwidth]{figures/lvls.jpg}
-    \caption{Example of sound level scale a function of the environment for new simulated scenes. TODO CITE}\label{fig:amb_levels}
-\end{figure}
-
-\subsubsection{Experiment}
-\label{sec:pval_exp}
+\subsection{Perceptual experiment}
+\label{sec:data_exp}
 
@@
 
-The scenes are played at a given sound level as discussed in Section~\ref{sec:pval_corp}, through the same computer and sound card configurations.
+The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations.
 Beyerdynamics DT-990 Pro headphones are used for all participants.
@@
 
-\subsection{Indicators for pleasantness prediction}
-\label{sec:inds}
-
-Several indicators are computed from the audio signals that are found in previous studies~\cite{aumond2017, gontier2018, ricciardi2014} to correlate well with perceptual parameters.\\
-
-First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level.
-Typically this includes indicators derived from sound level measurements.
-For this study the following are computed with 1~s (slow) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
-
-\begin{itemize}
-\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
-\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level.
-The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
-\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
-\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
-\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
-\end{itemize}
-
-The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity.
-To provide a baseline for pleasantness prediction the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
-
-The simulation process outputs ground truth source contributions in the form of separate channels.
-Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
-
-\begin{itemize}
-\item $L_{eq, s}$: Equivalent sound level for source $s$
-\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined
-\end{itemize}
-
-Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed.
-$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta_s(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$.
-It is parametrized by $\alpha$ and $\beta$ thresholds:
-
-\begin{align}
-\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
-\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
-\end{align}
-
-where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments.
-The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
-This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\
-
-In the remainder of this paper, the subscript $s$ is replaced with the corresponding source initial: $T$ for traffic, $V$ for voices and $B$ for birds.
-
-\subsection{Indicator prediction through deep learning}
-\label{sec:deep}
-
-\subsubsection{Learning corpus}
-\label{sec:deep_corp}
-
-To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
-\begin{itemize}
-\item The development set containing 400 45~s scenes (total 5 hours) which is used during the training process,
-\item The evaluation set containing 200 45~s scenes (total 2.5 hours) which is used to compute the performance of the trained model and its generalization capabilities.
-\end{itemize}
-
-The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:pval_corp}.
-Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples and recording annotations for other sources.
-The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\
-
-The developped model should extract relevant source information from a representation of the audio signal.
-Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal.
-Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}.
-Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range.
-Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception.
-The input frames of dimension 29x8 are then processed independently in this study.\\
-
-Target outputs are associated to each input frame to train the model.
-Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator.
-Thus, the source presence values before averaging along time can be used as weak labels.
-Alternatively, it is possible to perform sound level estimation using ground truth labels.
-Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.
-
-\subsubsection{Model architecture}
-\label{sec:deep_mdl}
-
-A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
-The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$ .
-The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
-The output of the last block is flattened then goes through a fully connected layer with output size 3.
-A final sigmoid activation is used to obtain outputs in the 0-1 range.
-The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
-\begin{equation}
-L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
-\end{equation}
-where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
-This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
-    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
-\end{figure}
-
-When using the model's outputs to estimate the time of presence of sources, a threshold of 0.5 is used to obtain binary presence values.
-
-\section{Results}
-\label{sec:results}
-
-\subsection{Perceptual validation}
-\label{sec:perc}
-%=> 3.1, 3.2
+\subsection{Validation}
+\label{sec:data_val}
 
@@
 
-\subsection{Baseline pleasantness models}
-\label{sec:base}
-%=> 3.3, 4.2
+\section{Methods}
+\label{sec:methods}
 
-\subsubsection{Perceptual models}
-\label{sec:base_perc}
+\subsection{Indicators for pleasantness prediction}
+\label{sec:methods_inds}
+
+Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters.
+These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.\\
+
+First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level.
+Typically this includes indicators derived from sound level measurements.
+For this study the following are computed with 1~s (slow) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
+
+\begin{itemize}
+\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
+\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level.
+The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
+\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
+\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
+\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
+\end{itemize}
+
+The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity.
+To provide a baseline for pleasantness prediction the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
+
+The simulation process outputs ground truth source contributions in the form of separate channels.
+Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
+
+\begin{itemize}
+\item $L_{eq, s}$: Equivalent sound level for source $s$
+\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined
+\end{itemize}
+
+Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed.
+$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta_s(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$.
+It is parametrized by $\alpha$ and $\beta$ thresholds:
+
+\begin{align}
+\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
+\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
+\end{align}
+
+where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments.
+The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
+This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\
+
+In the remainder of this paper, the subscript $s$ is replaced with the corresponding source initial: $T$ for traffic, $V$ for voices and $B$ for birds.
+
+\subsection{Indicator prediction using deep learning}
+\label{sec:methods_deep}
+
+To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
+\begin{itemize}
+\item The development set containing 400 45~s scenes (total 5 hours) which is used during the training process,
+\item The evaluation set containing 200 45~s scenes (total 2.5 hours) which is used to compute the performance of the trained model and its generalization capabilities.
+\end{itemize}
+
+The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}.
+Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples and recording annotations for other sources.
+The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
+As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\
+
+The developped model should extract relevant source information from a representation of the audio signal.
+Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal.
+Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}.
+Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range.
+Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception.
+The input frames of dimension 29x8 are then processed independently in this study.\\
+
+Target outputs are associated to each input frame to train the model.
+In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator.
+Thus, the source presence values obtained before averaging along time can be used as weak labels.
+Alternatively, it is possible to perform sound level estimation using ground truth labels.
+Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.\\
+
+A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
+The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$ .
+The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
+The output of the last block is flattened then goes through a fully connected layer with output size 3.
+A final sigmoid activation is used to obtain outputs in the 0-1 range.
+The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
+\begin{equation}
+L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
+\end{equation}
+where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
+This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
+    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
+\end{figure}
+
+When using the model's outputs to estimate the time of presence of sources, a threshold of 0.5 is used to obtain binary presence values.
+
+\section{Results}
+\label{sec:results}
+
+\subsection{Perceptual pleasantness model}
+\label{sec:results_perc}
 
@@
 $\hat P_{p1}$ and $\hat P_{p2}$ display similar estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness.
-The mean standard deviation of pleasantness assessments for this experiment is about 1.77, thus perceptual models are considered accurate with about 0.6 RMSE on the 10-point scale.
+The mean standard deviation of pleasantness assessments for this experiment is about 1.77 on the 11-point scale, the obtained perceptual models are thus considered accurate with about 0.6 RMSE.
 
@@
 
-\subsubsection{Physical models}
-\label{sec:base_phys}
+\subsection{Physical pleasantness models}
+\label{sec:results_phys}
 
-The analysis of physical indicators is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
+The analysis of physical indicators presented in Section~\ref{sec:methods_inds} is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
 Two scenes containing only one source are excluded for implementation concerns, resulting in $n=92$ scenes.\\
@@
 Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations.
-As in Section~\ref{sec:base_perc} a VIF check is performed on predictors to ensure that no multi-collinearity exists.
+As in Section~\ref{sec:results_perc} a VIF check is performed on predictors to ensure that no multi-collinearity exists.
 On the present corpus, the best model is:
@@
 \end{equation}
-Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:base_perc}.
+Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}.
 This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively, and no strong contribution of voices to pleasantness is identified in this corpus.\\
@@
 
-\subsection{Prediction of pleasantness through deep learning}
-\label{sec:pred}
+\subsection{Pleasantness prediction through deep learning}
+\label{sec:results_deep}
 
-Table~\ref{tab:perf_cmp} summarizes the performance of the source detection deep learning architecture presented in Section~\ref{sec:deep} on the evaluation dataset.
+The performance of the source detection deep learning architecture presented in Section~\ref{sec:methods_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}.
 The overall presence detection accuracy is 92.11\%.
@@
 \end{enumerate}
-Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}.
+Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:results_phys}.
 Thus, only the outputs corresponding to the presence of birds are used.
@@
 
-Pleasantness prediction using the detection model is as effective as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-Though, errors are significantly higher than the one of perceptual models in Section~\ref{sec:base_perc}.\\
+Pleasantness prediction using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
+Though, errors are significantly higher than the one of perceptual models in Section~\ref{sec:results_perc}.\\
 
@@
 Several hypotheses can be drawn from this result.
-First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment.
-However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models.
+First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the \textit{ex situ} setting of the experiment.
+However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for \textit{in situ} models.
 This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{refs}

### 2019-04-15 paper/paperV1.tex
Proofread 2

--- 2019-04-03 paper/paperDraft.tex
@@
 These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
+As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
 
 The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
-Specifically, the axis of pleasantness is found as the most explanatory of subjective assessments, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
+Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
 Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments.
-While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
+While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
 In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
@@
 Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
-With the advent of the Internet of Things and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017}, it may be feasible to predict pleasantness or other perceptual factors from acoustic measurements.
-Bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is an interesting and ongoing area of research.
+Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost.
+Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\
+
 Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
@@
 Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes.
-These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.\\
+These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
+In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\
 
-In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.
 In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
-However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
-In order to circumvent this issue, we propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017}.\\
+Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
+We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
 
-The contributions of this paper are thus three-fold:
+More precisely, the contributions of this paper are three-fold:
 \begin{enumerate}
-  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception
-  \item Propose numerical means for predicting the time of presence of sound sources from raw acoustic data
+  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
+  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
   \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
@@
 
-First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction.
+First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness.
 Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
@@
 A corpus is first created to validate the use of simulated sound scenes in the development of pleasantness prediction models.
-It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris.
+It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France.
 A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
-For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception.
-The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes.
+For each of the 19 recordings, 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception.
+The resulting 45 second extracts are replicated to study perceptual differences between recorded and simulated scenes.
 The recordings are first annotated in terms of the following properties:
@@
 To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios.
-Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
+Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
+We choose to restrict the scene composition to these 3 sources.
 Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
@@
 \item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
-\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
-\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
+\item Event-to-background ratios in dB for all events and backgrounds besides the first, in terms of normal distributions,
+\item The inter-onset of event occurrences in seconds, also in terms of normal distributions.
 \end{itemize}
@@
 
-Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds.
-The perception of source activity can be linked to notions of dominance, time of presence, or emergence.
+The perception of source activity can be linked to the notions of dominance, time of presence, or emergence.
 We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study.
-The influence of traffic on pleasantness is further separated into that of background and emergent events.
+The influence of traffic on pleasantness is further separated into that of background and salient events.
 Thus, 4 additional questions are presented:
@@
 Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood.
-Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios.
+Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}.
 The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants.
-A random pre-generated listening order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes (resp.
+All participants are first presented with the most quiet then most noisy of the recorded scenes (resp.
 P3 and P15) to calibrate their answers.
-For the remainder of the corpus all participants experience different scene ordering.
+A random listening order is generated for each subject to attenuate ordering effects for the remaining of the test.
 Participants can only listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\
@@
 
-A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.
+A total of 23 students aged from 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.
 
@@
 
-Participants evaluate each scene on criteria coded on 11-point discrete scales.
+Participants evaluate each scene on criteria using 11-point Likert scales.
 Many statistical tools assume that the studied data follows continuous, single mode distributions.
 It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments.
-A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}.
+A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus.
+Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal.
+and shown in Figure~\ref{fig:hclusters}.
 No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\
@@
 It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level.
-Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75).
+Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75).
 The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora.
@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots.
+Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace_rec} as dots.
 The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\
 
-Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace} using arrows.
+Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows.
 The standard deviation of assessments is represented by ellipses for one scene.
-The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity.
+The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}.
 This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\
 
-% FIGURE
 \begin{figure}[h]
     \centering
-     \begin{subfigure}[t]{0.55\textwidth}
+    \includegraphics[width=0.8\textwidth]{figures/pca_p1.eps}
+    \caption{Principal components analysis of the four general questions for scenes with real scenarios at the assessment level (n=575).
+Arrows show discrepancies between corresponding recorded and replicated scenes.
+For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
+\end{figure}
+\begin{figure}[h]
     \centering
-        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
-    \end{subfigure}%
-    \begin{subfigure}[t]{0.5\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
-    \end{subfigure}
-    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252).
-Each dot represent the projection of one scene on the resulting space.
-The components space is distorted compared to the literature.}\label{fig:pspace}
+    \includegraphics[width=0.8\textwidth]{figures/pca_sim.eps}
+    \caption{Principal components analysis of assessments for the simulated scenes (red dots) at the assessment level (n=1725), with projections of recorded and replicated scenes (blue dots).}\label{fig:pspace_sim}
 \end{figure}
@@
 The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity.
-To provide a baseline for pleasantness prediction the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
+To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
 
-The simulation process outputs ground truth source contributions in the form of separate channels.
-Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
+The simulation process outputs ground truth source contributions as separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
 
@@
 \begin{itemize}
-\item The development set containing 400 45~s scenes (total 5 hours) which is used during the training process,
-\item The evaluation set containing 200 45~s scenes (total 2.5 hours) which is used to compute the performance of the trained model and its generalization capabilities.
+\item The development set of 5 hours containing 400 scenes of 45~s each, which is used during the training process,
+\item The evaluation set of 2.5 hours containing 200 scenes of 45~s each, which is used to compute the performance of the trained model and its generalization capabilities.
 \end{itemize}
@@
 The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}.
-Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples and recording annotations for other sources.
+Only traffic, human voices and birds sources compose the sound scenes.
 The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
@@
 Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range.
-Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception.
-The input frames of dimension 29x8 are then processed independently in this study.\\
+Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception.
+The spectral blocks of dimension 29x8 are then processed independently in this study.\\
 
@@
 In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator.
-Thus, the source presence values obtained before averaging along time can be used as weak labels.
-Alternatively, it is possible to perform sound level estimation using ground truth labels.
-Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.\\
+Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels.
+Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha, \beta)$.
+Though, training a model for sound level regression is more difficult than for source detection as it requires the architecture to perform implicit source separation.
+This will not be discussed here as it did not result in better time of presence estimations in our experiments.\\
 
@@
 Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
-The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions.
-Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
+The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions.
+Indeed, pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
 In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
@@
 Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 10-point scale.
-In $\hat P_{\varphi 2}$ introducing the $L_{10}-L_{90}$ emergence has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\
+Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{\varphi 2}$ has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\
 
@@
 Regarding the type of errors, they are equally split between false positives and false negatives.
-Though, these rates vary depending on the source.
+Though, these rates vary depending on the type of source.
 Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%.
@@
 The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}.
-First, pleasantness estimations are equally performant using the model's predictions or the ground truth labels , with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus.
+First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus.
 This performance of the detection model is expected given the low errors on time of presence estimates.
@@
 Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness.
-The trained deep learning architecture already displays few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\
+The trained deep learning architecture already displays only few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\
 
@@
 Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
-Thus, in some cases the deep learning model's predictions are considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\
+Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\
 
@@
 This is expected as the architecture is not trained on complex scenes.
-Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes.
+Though, pleasantness estimates are interestingly still better on average than on the corpus of simulated scenes.
 The detection model is thus quite robust to scenes with additional sources as perturbators.
-A full study with a larger evaluation corpus would be needed to further validate the model's robustness to diverse real-life scenarios.
 
@@
 
-Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction in urban environments, leads to better estimates than previous models from acoustic indicators.\\
+Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, leads to better estimates than previous state-of-the-art models.\\
 
 Future work will focus on the robustness of the source detection model, which heavily depends on the quantity and quality of data available for scene simulation.
+Together with additional recordings, data augmentation techniques for machine learning will be studied to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.
 Furthermore, this methodology can be adapted to specific environments by extending or refining the source taxonomy in both the scene simulation process and detection model.

### 2019-04-29 paper/paperV1.tex
Revision CL+PA

@@
 \label{sec:intro}
-
 In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
@@
 These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
+As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.\\
 
 The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
-Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
-Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments.
-While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
-In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
+Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}.
+In this paper, we will focus on the first affective dimension which is pleasantness.\\
 
-Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
+Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}.
+While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} and traffic noise, human voices and birds calls can respectively be used as a proxy of its three types~\cite{lavandier2006, ricciardi2014, aumond2017}.
+In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}.
+Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.\\
+
 Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost.
-Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\
-
+Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.
 Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
 Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
-Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes.
+Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes.
 These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
-In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\
+In order to tackle this issue, the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.
 
-In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
-Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
+In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
+Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings.
+To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels.
 We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
@@
 
-First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness.
+First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate the relevance of the proposed protocol for applications in \textit{in situ} sound environments.
+Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
+Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.
+
+%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
+Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities.
+These maps aim at providing useful tools for urban planning and noise reduction purposes.
+As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
+%
+%The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
+Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
+Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments.
+While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
+In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
+%
+%Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
+Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost.
+Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\
+%
+%Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
+Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
+Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes.
+These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
+In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\
+%
+%In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
+Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
+We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
+%
+%More precisely, the contributions of this paper are three-fold:
+%\begin{enumerate}
+%  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
+%  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
+%  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
+%\end{enumerate}
+%
+%First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness.
 Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
@@
 
-A corpus is first created to validate the use of simulated sound scenes in the development of pleasantness prediction models.
+A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models.
 It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France.
@@
 The resulting 45 second extracts are replicated to study perceptual differences between recorded and simulated scenes.
+This scale of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments.
 The recordings are first annotated in terms of the following properties:
@@
 
-Sound scenes are then simulated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources.
-This database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
+Sound scenes are then simulated using simScene~\cite{rossignol2015}, a software library in Matlab which allows the simulation of sound scenes as the additive composition of sound sources using a database of recordings associated to these sources.
+Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
 The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
-All 6 recorded and 19 replicated scenes are normalized so that their playback sound level is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\
+All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\
 
@@
     \end{subfigure}
-    \caption{Estimated time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
+    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
 \end{figure}
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
+While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.\\
 
@@
 All participants are first presented with the most quiet then most noisy of the recorded scenes (resp.
-P3 and P15) to calibrate their answers.
+P3 and P15).
 A random listening order is generated for each subject to attenuate ordering effects for the remaining of the test.
-Participants can only listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\
+Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\
 
@@
 Beyerdynamics DT-990 Pro headphones are used for all participants.
-These headphones are calibrated in a semi-anechoic chamber, resulting in a relation between playback Leq and LAeq and sound card output voltage for pink noise.
-This ensures that correctly scaled scenes will be heard at the desired sound level by listeners.\\
+These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.\\
 
@@
 
-\subsection{Validation}
+\subsection{Corpus validation}
 \label{sec:data_val}
@@
 
-\subsection{Indicators for pleasantness prediction}
+\subsection{Acoustical indicators for soundscape description}
 \label{sec:methods_inds}
@@
 
-The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity.
+The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:
+\begin{equation}
+TFSD_{f, t} = \frac{\lvert\frac{dL(f, t)}{dfdt}\rvert}{\sum_{f=31.5Hz}^{f=16kHz}\lvert\frac{dL(f, t)}{dfdt}\rvert}
+\end{equation}
 To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
@@
 
-\subsection{Indicator prediction using deep learning}
+\subsection{Source presence detection using deep learning}
 \label{sec:methods_deep}
@@
 A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
-The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$ .
+The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$.
 The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
@@
 \label{sec:results}
+
+The proposed method for predicting the time of presence of sources is evaluated in the context of pleasantness estimation.
+To this aim, baseline models are first constructed using data from the experiment in Section~\ref{sec:data_exp} and from previous work.
 
@@
 This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
-Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\
+Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.
+However, its correlation with the presence of human voices is the strongest, as expected~\cite{axelsson2010}.\\
 
@@
 The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$.
-Except for the interest (I) it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
+Except for the interest (I) the $L_{T, p}$ displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
 
@@
 \begin{align}
-\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p}\\
-\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p}
+\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p} \label{eq:pp1}\\
+\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p} \label{eq:pp2}
 \end{align}
@@
 $\hat P_{p1}$ and $\hat P_{p2}$ display similar estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness.
-The mean standard deviation of pleasantness assessments for this experiment is about 1.77 on the 11-point scale, the obtained perceptual models are thus considered accurate with about 0.6 RMSE.
+The performance of obtained perceptual models is about 0.6 RMSE, which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).
 
@@
 The analysis of physical indicators presented in Section~\ref{sec:methods_inds} is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
-Two scenes containing only one source are excluded for implementation concerns, resulting in $n=92$ scenes.\\
+The 6 recorded scenes are excluded from the study as the ground truth source composition is unknown.
+Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.\\
 
 Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments.
-First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness.
+First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
 Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source.
@@
 \begin{equation}
-\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta)
+\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta) \label{eq:pphi}
 \end{equation}
 Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}.
-This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively, and no strong contribution of voices to pleasantness is identified in this corpus.\\
+This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ (eq.\ref{eq:pp1}) and $\hat P_{p2}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of voices to pleasantness is identified in this corpus.\\
 
@@
 Though, compared to the best perceptual models its explained variance is about 9\% lower.
-Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 10-point scale.
+Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 11-point scale.
 Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{\varphi 2}$ has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\
@@
 
-\subsection{Pleasantness prediction through deep learning}
+\subsection{Time of presence prediction through deep learning}
 \label{sec:results_deep}
@@
 \centering
-\caption{Performance of the model predicting source presence with $\hat T_s(\alpha, \beta)$ labels.
+\caption{Performance of the model predicting source presence with binary ground truth labels obtained from $\hat T_s(\alpha, \beta)$.
 Presence metrics are computed for n=8600 1s frames and time of presence metrics on n=200 45s scenes.
@@
 	$\hat T_s(\alpha, \beta)$ RMSE & 0.12 & 0.13 & 0.10 & 0.11 \\
-	$\hat T_s(\alpha, \beta)$ MAE & 0.06 & 0.05 & 0.06 & 0.08 \\ \hline
 \end{tabular}
@@
 The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}.
-First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus.
+First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
 This performance of the detection model is expected given the low errors on time of presence estimates.
@@
 
-The perceptual models of pleasantness are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
+The perceptual models of pleasantness (eq.\ref{eq:pp1}-\ref{eq:pp2}) are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
 \begin{align}
@@
 
-The physical model on this corpus includes only the contribution of bird sources. This result is attributed to the overall low diversity of the studied corpus:
+The physical model on this corpus (eq.\ref{eq:pphi}) includes only the contribution of bird sources.
+This result is attributed to the overall low diversity of the studied corpus which gives more importance to this perceived dimension:
 \begin{itemize}
@@
 Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness.
-These properties result in no significant contribution of voice events to pleasantness for this corpus.
+These properties result in no significant correlation of indicators describing voice events to pleasantness for this corpus.
 \end{itemize}
@@
 
-A perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual characterization.
+A perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual soundscape characterization.
 Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.\\
 
-Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, leads to better estimates than previous state-of-the-art models.\\
+Subsequently, a deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, compares to previous state-of-the-art models.
+Though, provided enough data the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a range of applications for soundscape characterization, categorization, or perceptual mapping.\\
 

### 2019-06-03 paper/paperV1.tex
JFP proofread corrections

@@
 \date{
-$^1$ LS2N, UMR 6004, Ecole Centrale de Nantes, CNRS, F-44300, France\\
+$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
 $^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
@@
 Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}.
-While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} and traffic noise, human voices and birds calls can respectively be used as a proxy of its three types~\cite{lavandier2006, ricciardi2014, aumond2017}.
+While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}.
 In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}.
@@
 \begin{enumerate}
-  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
-  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
-  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
+  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled.
+  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
+  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
 \end{enumerate}
@@
 
-%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
-Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities.
-These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
-%
-%The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
-Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}.
-Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments.
-While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}.
-In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
-%
-%Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
-Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost.
-Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\
-%
-%Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
-Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
-Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes.
-These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
-In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\
-%
-%In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
-Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature.
-We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
-%
-%More precisely, the contributions of this paper are three-fold:
-%\begin{enumerate}
-%  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
-%  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
-%  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
-%\end{enumerate}
-%
-%First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness.
-Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
-Estimations of pleasantness using these predictions are next compared to existing models from physical indicators in Section~\ref{sec:results}.
-
-\section{Data}
+\section{Perception of simulated sound scenes}
 \label{sec:data}
@@
 A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
-For each of the 19 recordings, 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception.
-The resulting 45 second extracts are replicated to study perceptual differences between recorded and simulated scenes.
-This scale of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments.
-The recordings are first annotated in terms of the following properties:
+For each of the 19 recordings, 45 seconds of audio in a single channel are extracted.
+This duration of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments.
+The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
+The extracts are then subjectively annotated in terms of the following properties:
 
 \begin{itemize}
-\item Background sources that are present throughout the whole scene, and are associated with a sound level.
+\item Background sources that are present throughout the whole scene, and their respective sound level considered constant.
 \item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
@@
 
-Sound scenes are then simulated using simScene~\cite{rossignol2015}, a software library in Matlab which allows the simulation of sound scenes as the additive composition of sound sources using a database of recordings associated to these sources.
+The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}.
+\textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources.
 Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
 The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
-All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\
+All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\
 
@@
 Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
-These statistics include, conditionally to the ambiance:
+These statistics include, conditionally to the ambiance (resp.
+\textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):
 
@@
 \item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
-\item Event-to-background ratios in dB for all events and backgrounds besides the first, in terms of normal distributions,
-\item The inter-onset of event occurrences in seconds, also in terms of normal distributions.
+\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions with parameters identified from annotations of recordings,
+\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
 \end{itemize}
@@
 Adjustments are made on the variance of the considered properties to better cover possible scenarios.
-The EBR of voice events, mostly read English recordings, are reduced to improve realism.
-An additional \textit{place} ambiance is added with properties derived empirically from available data.
-Diverse new scenarios are then generated by sampling the background and event properties distributions.
-For each scene the time of presence for the traffic, voice and bird sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space.
+Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism.
+An additional \textit{square} ambiance is added with properties derived empirically from available data.
+Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes.
+For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}.
+Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way.
+Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled.
+As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space.
 This is shown in Figure~\ref{fig:tvb_pres}.
-The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB to 77.1~dB.\\
+The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.\\
 
@@
 
-A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point (0-10) scales.
-These scales are presented in french and translated in this report using standard terminology.
+A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
+These scales are presented in French and translated in this report using standard terminology.
 The first 5 scales relate to general properties of the scene:
@@
 The influence of traffic on pleasantness is further separated into that of background and salient events.
-Thus, 4 additional questions are presented:
+Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
 \begin{itemize}
@@
 \end{itemize}
+where $p$ denotes a perceptual evaluation.\\
 
-Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood.
+Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood.
 Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}.
@@
 P3 and P15).
-A random listening order is generated for each subject to attenuate ordering effects for the remaining of the test.
+A random listening order is generated for each subject to control ordering effects for the remaining of the test.
 Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\
@@
 
-A total of 23 students aged from 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.
+A total of 23 students aged from 22 to 23 yo including 16 males and 7 females at Ecole Centrale de Nantes completed the test.
+All gave consent prior to the experiment and reported normal hearing.
+With 23 subjects, the incomplete block design is not perfectly balanced.
 
@@
 
-Participants evaluate each scene on criteria using 11-point Likert scales.
+%Participants evaluate each scene on criteria using 11-point Likert scales.
 Many statistical tools assume that the studied data follows continuous, single mode distributions.
@@
 No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\
+%
+%\begin{figure}[!h]
+%    \centering
+%    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
+%    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
+%\end{figure}
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
-    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
-\end{figure}
-
-The perceptual responses of the 6 recorded and corresponding replicated scenes are then compared.
+The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared.
 Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios.
-Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold.
-Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices.
+Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold.
+As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}.
+On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale (catherine?).
+Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters.
+The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4.
 For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half.
@@
 Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration.
-To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly.
-Discrepancies can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy.
+To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3.
+Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy.
 For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.
-Overall, no consistent difference between the perception of recorded and replicated scenes is found for the studied points.
+Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
 Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\
@@
 \centering
-\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Kolmogorov-Smirnov test are in bold (n=23, p<0.05)}
+\caption{Mean differences of perceptual assessments between original and replicated sound scenes.
+Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
 \label{tab:ogrep}
@@
 	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
-	P1 & 0.43 & \textbf{-1.65} & -1.04 & 0.43 & 0.13 & -0.91 & 0.39 & -2.09 & 0.61 \\
+	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
 	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
-	P4 & \textbf{0.91} & 0 & \textbf{-1.83} & 0.48 & 1.30 & -0.96 & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
-	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & -0.91 & 0.09 & \textbf{-1.43} \\
-	P15 & -1.35 & 0.52 & 0.52 & -1.17 & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
-	P18 & 1.13 & -0.30 & -1.17 & -0.43 & 1.39 & -1.04 & -1.83 & 0.83 & \textbf{1.30} \\ \hline
+	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
+	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
+	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
+	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
 \end{tabular}
@@
 
-To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied.
-It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level.
-Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75).
-The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora.
-For both sets the variance explained by the first two components is similar (resp.
-79.4\% - 18.1\% and 79.6\% - 15.2\%).
+To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied.
+It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants.
+No standardization is applied to the data.
+Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively.
+The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora.
+For both sets the variance explained by the first two components is similar, resp.
+79.4\% - 18.1\% and 79.6\% - 15.2\%.
 Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}.
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace_rec} as dots.
-The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\
+Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes.
+The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.\\
 
 Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows.
-The standard deviation of assessments is represented by ellipses for one scene.
+The standard deviation of assessments is represented by ellipses for one location (P1).
 The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/pca_p1.eps}
-    \caption{Principal components analysis of the four general questions for scenes with real scenarios at the assessment level (n=575).
+    \includegraphics[width=0.9\textwidth]{figures/pca_p1.eps}
+    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).
+Red dots are projections of the assessments of each scene averaged over subjects.
 Arrows show discrepancies between corresponding recorded and replicated scenes.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/pca_sim.eps}
-    \caption{Principal components analysis of assessments for the simulated scenes (red dots) at the assessment level (n=1725), with projections of recorded and replicated scenes (blue dots).}\label{fig:pspace_sim}
+    \includegraphics[width=0.9\textwidth]{figures/pca_sim.eps}
+    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
+Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
 \end{figure}
@@
 Typically this includes indicators derived from sound level measurements.
-For this study the following are computed with 1~s (slow) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
+For this study the following variables are computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
 
@@
 \item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level.
-The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
+The $L_{10}$ is often associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
 \item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
@@
 \begin{equation}
-TFSD_{f, t} = \frac{\lvert\frac{dL(f, t)}{dfdt}\rvert}{\sum_{f=31.5Hz}^{f=16kHz}\lvert\frac{dL(f, t)}{dfdt}\rvert}
+TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
 \end{equation}
+where $L(f, t)$ is the third-octave spectrum of the signal.
 To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
 
-The simulation process outputs ground truth source contributions as separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
+The simulation process of the scenes outputs ground truth source contributions as separate channels.
+Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
 
@@
 Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed.
-$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta_s(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$.
+$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta L_{s}(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$.
 It is parametrized by $\alpha$ and $\beta$ thresholds:
@@
 \begin{align}
-\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
-\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
+\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
+\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
 \end{align}
 
-where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments.
-The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
-This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\
+where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene.
+This indicator evaluates the presence or absence of a given source in a small time frame.
+First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated.
+Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold.
+A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene.
+The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
+This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.\\
 
@@
 \begin{itemize}
-\item The development set of 5 hours containing 400 scenes of 45~s each, which is used during the training process,
-\item The evaluation set of 2.5 hours containing 200 scenes of 45~s each, which is used to compute the performance of the trained model and its generalization capabilities.
+\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
+\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
 \end{itemize}
 
-The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}.
+The generation procedure is the same for both datasets, and is equivalent to that of the 75 scenes in the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}.
 Only traffic, human voices and birds sources compose the sound scenes.
-The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\
+To generate the scenes, it is important to ensure a complete independance of the training and test set.
+For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
+As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.\\
 
@@
 Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels.
-Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha, \beta)$.
-Though, training a model for sound level regression is more difficult than for source detection as it requires the architecture to perform implicit source separation.
+Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.
+Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data.
 This will not be discussed here as it did not result in better time of presence estimations in our experiments.\\
@@
 A final sigmoid activation is used to obtain outputs in the 0-1 range.
-The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
+The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
+During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
 \begin{equation}
-L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
+BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
 \end{equation}
 where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
-This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\
+This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.
+During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher.
+The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.\\
 
@@
 \end{figure}
-
-When using the model's outputs to estimate the time of presence of sources, a threshold of 0.5 is used to obtain binary presence values.
 
@@
 We first study the relations between perceptual scales with respect to existing work.
-Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene.
+Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
 The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions.
-Indeed, pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds.
+Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
 In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
 This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
-Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.
-However, its correlation with the presence of human voices is the strongest, as expected~\cite{axelsson2010}.\\
+Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies.
+However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.\\
 
-Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds.
-The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$.
-Except for the interest (I) the $L_{T, p}$ displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
+Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42).
+The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
 
-As a baseline for pleasantness prediction, multilinear models are built as a function of both the overall loudness (OL) and source-specific parameters.
-To ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to model regression.
-Only models for which all predictors verify $VIF<5$ are considered valid.
-The best resulting models with statistically significant estimates ($p<0.05$) for all predictors are:
+As a baseline, multilinear models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100).
+All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
+Only combinations for which all predictors verify $VIF<5$ are considered valid.
+The best resulting models in terms of $R^2_{adj}$ with statistically significant coeffeicient estimates ($p<0.05$) are:
 
 \begin{align}
-\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p} \label{eq:pp1}\\
-\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p} \label{eq:pp2}
+\hat P_{1, p} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p} \label{eq:pp1}\\
+\hat P_{2, p} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p} \label{eq:pp2}
 \end{align}
 
-Both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
-A small positive contribution of the time of presence of voices is also found despite no significant correlation.\\
+Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
+A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.\\
 
 Table~\ref{tab:percm} summarizes performance metrics computed for the two models.
-$\hat P_{p1}$ and $\hat P_{p2}$ display similar estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness.
+$\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness.
 The performance of obtained perceptual models is about 0.6 RMSE, which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).
@@
 	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
-	$\hat P_{p1}$ & 0.59 & 0.91 & 0.95** \\
-	$\hat P_{p2}$ & 0.61 & 0.90 & 0.95** \\ \hline
+	$\hat P_{1, p}$ & 0.59 & 0.91 & 0.95** \\
+	$\hat P_{2, p}$ & 0.61 & 0.90 & 0.95** \\ \hline
 \end{tabular}
@@
 First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
-Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source.
-Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weaker for other sources.
-The proposed estimates of the time of presence $\hat T_s(\alpha, \beta)$ show strong correlations to their perceptual counterparts (r>0.8), along with good source discrimination properties.
-They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity.
+Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71).
+Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources.
+The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim.
+They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}.
+They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus.
 Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$.
@@
 	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
-	$\hat T_T(\alpha, \beta)$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
-	$\hat T_V(\alpha, \beta)$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
-	$\hat T_B(\alpha, \beta)$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
+	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
+	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
+	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
 \end{tabular}
@@
 Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations.
-As in Section~\ref{sec:results_perc} a VIF check is performed on predictors to ensure that no multi-collinearity exists.
-On the present corpus, the best model is:
+As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
+On the present corpus, the best model in terms of $R^2_{adj}$ is:
 \begin{equation}
-\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta) \label{eq:pphi}
+\hat P_{1, \varphi} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
 \end{equation}
+where $\varphi$ indicates a model from physical variables.
 Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}.
-This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ (eq.\ref{eq:pp1}) and $\hat P_{p2}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of voices to pleasantness is identified in this corpus.\\
+This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.\\
 
 This model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal.
-Both models' coefficients are optimized on the studied data:
+Both models' coefficients are re-optimized on the studied data for a fair comparison:
 \begin{align}
-\hat P_{\varphi 2} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
-\hat P_{\varphi 3} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} + 5.00~TFSD_{4kHz, 1/8s}
+\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
+\hat P_{3, \varphi} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} + 5.00~TFSD_{4kHz, 1/8s}
 \end{align}
@@
 Table~\ref{tab:physm} compares the performance of the three models.
-Expectedly, $\hat P_{\varphi 1}$ outperforms both baselines.
-Though, compared to the best perceptual models its explained variance is about 9\% lower.
-Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 11-point scale.
-Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{\varphi 2}$ has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\
+Expectedly, $\hat P_{1, \varphi}$ outperforms both baselines, though a validation on a different corpus would be necessary to conclude on its capabilities.
+Compared to the best perceptual models in Table~\ref{tab:percm} its $R^2_{adj}$ is about 9\% lower.
+Its RMSE is 0.83, which is also high compared to perceptual baselines, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 11-point scale in this study.
+Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.\\
 
@@
 	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
-	$\hat P_{\varphi 1}$ & 0.83 & 0.82 & 0.91** \\
-	$\hat P_{\varphi 2}$ & 0.90 & 0.79 & 0.89** \\
-	$\hat P_{\varphi 3}$ & 0.90 & 0.78 & 0.89** \\ \hline
+	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
+	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
+	$\hat P_{3, \varphi}$ & 0.90 & 0.78 & 0.89** \\ \hline
 \end{tabular}
@@
 \centering
-\caption{Performance of the model predicting source presence with binary ground truth labels obtained from $\hat T_s(\alpha, \beta)$.
+\caption{Performance of the model predicting source presence with binary ground truth labels obtained from $\hat T_s(\alpha_{opt}, \beta_{opt})$.
 Presence metrics are computed for n=8600 1s frames and time of presence metrics on n=200 45s scenes.
@@
 	Presence TN (\%) & 92.30 & 97.54 & 94.76 & 87.70 \\
-	Presenct FP (\%) & 7.70 & 2.46 & 5.24 & 12.30 \\
+	Presence FP (\%) & 7.70 & 2.46 & 5.24 & 12.30 \\
 	Presence FN (\%) & 8.04 & 7.30 & 10.17 & 6.63 \\ \hline
-	$\hat T_s(\alpha, \beta)$ RMSE & 0.12 & 0.13 & 0.10 & 0.11 \\
+	$\hat T_s(\alpha_{opt}, \beta_{opt})$ RMSE & 0.12 & 0.13 & 0.10 & 0.11 \\
 \end{tabular}
@@
 \begin{enumerate}
-\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
+\item The 6 recorded scenes, which contain additional sources not present in the datasets simulated for the optimization of the deep learning model,
+\item The 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017},
 \item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
 \end{enumerate}
-Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:results_phys}.
+Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:results_phys}.
 Thus, only the outputs corresponding to the presence of birds are used.
-The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}.
+The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred}.
 First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
-This performance of the detection model is expected given the low errors on time of presence estimates.
-Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training.
-In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\
+This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}.
+Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training.
+This indicates that the detection model generalizes well to new content and sources.
+For the simulated subcorpus of the experiments all performance metrics are comparable.
+Applying the detection model on the small corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
+This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.\\
 
@@
 \centering
-\caption{Pleasantness prediction quality using the source detection model compared to labels.
-The experiment corpus is split into respectively the 25 recorded and replicated scenes and 75 simulated scenes.}
+\caption{Pleasantness prediction quality on the perceptual experiment corpus using the source detection model compared to labels.
+The corpus is split in three parts: the 6 recorded scenes (Rec.), the 19 replicated scenes (Rep.), and the 75 scenes with simulated scenarios (Sim.).}
 \label{tab:pppred}
 %\resizebox{\columnwidth}{!}{
-\begin{tabular}{ c | c c c | c c c }
+\begin{tabular}{ c | c c c c | c c c }
 \hline
-	Presence & \multicolumn{3}{|c}{Model output} & \multicolumn{3}{|c}{$\hat T_s(\alpha, \beta)$ labels} \\ \hline
-	Sub-corpus & All & 1 & 2 & All & 1 & 2 \\ \hline
-	RMSE & 0.84 & 0.80 & 0.85 & 0.83 & 0.73 & 0.85 \\ \hline
-	r & 0.91** & 0.92** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
-	$R^2_{adj}$ & 0.82 & - & - & 0.82 & - & - \\ \hline
+	Model & \multicolumn{4}{|c}{$P_{1, \varphi}$ with model outputs} & \multicolumn{3}{|c}{$P_{1, \varphi}$ with $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels} \\ \hline
+	Sub-corpus & All & Rec. & Rep. & Sim. & All & Rep. & Sim. \\ \hline
+	RMSE & 0.84 & 1.09 & 0.68 & 0.85 & 0.83 & 0.73 & 0.85 \\ \hline
+	r & 0.91** & 0.89** & 0.93** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
+	$R^2_{adj}$ & 0.82 & 0.73 & 0.82 & 0.80 & 0.82 & 0.81 & 0.79 \\ \hline
 \end{tabular}
@@
 
-Pleasantness prediction using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-Though, errors are significantly higher than the one of perceptual models in Section~\ref{sec:results_perc}.\\
+Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
+Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.\\
 
@@
 \begin{align}
-\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
-\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
+\hat P_{3, p} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
+\hat P_{4, p} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
 \end{align}
@@
 
-Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus.
+Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels on the simulated corpus.
 This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
 Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
-Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\
+Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha_{opt}, \beta_{opt})$.\\
 
-The detection-based pleasantness predictions are of lower quality on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy.
-This is expected as the architecture is not trained on complex scenes.
-Though, pleasantness estimates are interestingly still better on average than on the corpus of simulated scenes.
-The detection model is thus quite robust to scenes with additional sources as perturbators.
+A higher pleasantness predictions quality is obtained using detection-based presence labels compared to $\hat T_s(\alpha_{opt}, \beta_{opt})$ on the corpus containing replicated scenes, which feature higher polyphonies and a richer source taxonomy.
+Although interesting, this result may be attributed to the small sample size as the architecture is not trained on complex scenes.
+The detection model is thus quite robust to scenes with additional sources as perturbators, but a study on a larger corpus would be required to validate this conclusion.
 

### 2019-06-04 paper/paperV1.tex
JFP proofread

@@
 These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.\\
+As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.
 
@@
 Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}.
-In this paper, we will focus on the first affective dimension which is pleasantness.\\
-
+In this paper, we will focus on the first affective dimension which is pleasantness.
 Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}.
@@
 In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}.
-Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.\\
+Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
 
@@
 To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels.
-We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
+We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.
 
@@
 The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
-All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\
+All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
 
@@
 This is shown in Figure~\ref{fig:tvb_pres}.
-The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.\\
+The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.
 
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.\\
+While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.
 
@@
 \end{itemize}
-where $p$ denotes a perceptual evaluation.\\
+where $p$ denotes a perceptual evaluation.
 
@@
 A random listening order is generated for each subject to control ordering effects for the remaining of the test.
-Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\
+Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.
 
@@
 These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
-From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.\\
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.
 
@@
 As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}.
-On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale (catherine?).
+On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale.
 Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters.
@@
 Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
-Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\
+Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.
 
@@
 Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes.
-The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.\\
+The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.
 
@@
 The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}.
-This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\
+This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.
 
@@
 Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters.
-These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.\\
+These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.
 
@@
 The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
-This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.\\
+This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.
 
@@
 For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.\\
+As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.
 
@@
 Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception.
-The spectral blocks of dimension 29x8 are then processed independently in this study.\\
+The spectral blocks of dimension 29x8 are then processed independently in this study.
 
@@
 Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data.
-This will not be discussed here as it did not result in better time of presence estimations in our experiments.\\
+This will not be discussed here as it did not result in better time of presence estimations in our experiments.
 
@@
 During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher.
-The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.\\
+The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.
 
@@
 Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies.
-However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.\\
+However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.
 
 Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42).
-The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\
+The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.
 
@@
 Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
-A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.\\
+A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.
 
@@
 The 6 recorded scenes are excluded from the study as the ground truth source composition is unknown.
-Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.\\
+Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.
 
@@
 Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$.
-In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.\\
+In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.
 
@@
 Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}.
-This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.\\
+This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
 
@@
 Its RMSE is 0.83, which is also high compared to perceptual baselines, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 11-point scale in this study.
-Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.\\
+Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.
 
@@
 Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model.
-The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.\\
+The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.
 
@@
 Applying the detection model on the small corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
-This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.\\
+This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.
 
@@
 Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.\\
+Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.
 
@@
 However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for \textit{in situ} models.
-This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\
+This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.
 
@@
 Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness.
-The trained deep learning architecture already displays only few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\
+The trained deep learning architecture already displays only few detection errors for these sources, and can be extended to any additional source provided enough data is available.
 
@@
 Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
-Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha_{opt}, \beta_{opt})$.\\
+Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha_{opt}, \beta_{opt})$.
 
@@
 A perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual soundscape characterization.
-Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.\\
+Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.
 
 Subsequently, a deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, compares to previous state-of-the-art models.
-Though, provided enough data the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a range of applications for soundscape characterization, categorization, or perceptual mapping.\\
+Though, provided enough data the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a range of applications for soundscape characterization, categorization, or perceptual mapping.
 

### 2019-06-04 paper/paperV1.tex
abstract 1

@@
 \begin{abstract}
+Need for characterization of soundscape perception through objective descriptors, 
+Dependance on sources, level measurements insufficient
+sensor networks + deep learning
 
+For such purposes, a corpus of simulated sound scenes is proposed that allows the use of ground truth source contributions for the development of learning-based models.
+It includes semi-randomly generated original scenarios representative of the urban environment.
+This simulated corpus is shown through a listening experiment to yield similar perceptual responses compared to recorded sound scenes.
+
+to predict the perceived source presence in 1~s time frames with labels obtained from
+
+This model is compared to both perceptual and physical baseline models obtained from the perceptual experiment's data in the context of pleasantness prediction
 \end{abstract}

### 2019-06-04 paper/paperV1.tex
abstract

--- 2019-06-04 paper/paperV1.tex
@@
 \begin{abstract}
+
+Urban areas are often plagued with noise pollution.
+Even though a great deal of this pollution can be accounted by the sound volume, the literature shows that the recognition of sources also play a great role in the way humans are impacted.
+
+There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also source-specific attributes such as the perceived time of presence.
+Combined, they allow us to predict more precisely high level concepts like the pleasantness.
+However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.
+
+This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient for the task at hand: the prediction of the pleasantness.
+
+A corpus of simulated sound scenes is first proposed.
+This corpus is shown through a listening experiment to yield similar perceptual responses compared to recorded sound scenes.
+The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.
 

### 2019-06-04 paper/paperV1.tex
fix ref

@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes.
+Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_rec} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes.
 The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.

### 2019-06-05 paper/paperV1.tex
Abstract

@@
 \begin{abstract}
-
-Urban areas are often plagued with noise pollution.
-Even though a great deal of this pollution can be accounted by the sound volume, the literature shows that the recognition of sources also play a great role in the way humans are impacted.
-
-There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also source-specific attributes such as the perceived time of presence.
-Combined, they allow us to predict more precisely high level concepts like the pleasantness.
+The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces).
+The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environnements.
+There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as the perceived time of presence.
+Combined, they allow us to predict the first affective dimension of the soundscape: its pleasantness.
 However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.
-
-This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient for the task at hand: the prediction of the pleasantness.
-
+This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient
+to predict high level concepts like the pleasantness.
 A corpus of simulated sound scenes is first proposed.

### 2019-06-11 paper/paperV1.tex
final ml proof read

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[twocolumn]{article}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Comment / uncomment for one or two column(s) fomat
+%\documentclass{article}
+
+\usepackage[modulo,switch]{lineno}
+\modulolinenumbers[1]
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Comment / uncomment for showing line numbers
+\linenumbers
+
+\usepackage[latin9]{inputenc}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+
+\makeatletter\@ifundefined{date}{}{\date{}}
+\makeatother
+
+\markright{\hfill Gontier {\em et al.}, p.\ }
+\pagestyle{myheadings}
+
+\paperheight297mm \paperwidth210mm
+\textwidth170mm  \textheight245mm  \oddsidemargin 20mm
+\evensidemargin\oddsidemargin \hoffset-22.4mm \voffset-28.4mm
+\topmargin0pt \headheight20mm \headsep4mm \topskip0mm
+\footskip17.5mm \columnsep7mm \arraycolsep2pt \parindent10pt
+\renewcommand{\abstractname}{Summary}
+
 \usepackage[english]{babel}
@@
 \usepackage{fancyhdr}
-\usepackage{a4wide} % Smaller margins = more text per page.
 \usepackage[hidelinks]{hyperref}
@@
 \newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
-\pagestyle{fancy}
 \begin{document}
@@
 The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces).
-The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environnements.
-There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as the perceived time of presence.
-Combined, they allow us to predict the first affective dimension of the soundscape: its pleasantness.
+The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments.
+There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.
+Combined, they allow us to predict important affective dimensions of the soundscape such as its pleasantness.
 However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.
+
 This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient
-to predict high level concepts like the pleasantness.
-A corpus of simulated sound scenes is first proposed.
-This corpus is shown through a listening experiment to yield similar perceptual responses compared to recorded sound scenes.
+to predict perceived pleasantness.
+To do so, a corpus of simulated sound scenes is first designed.
+Perceptual attributes corresponding to those stimuli are gathered through a listening experiment.
 The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.
+
+% to yield similar perceptual responses compared to recorded sound scenes.
+PACS numbers: 43.66.Lj (Perceptual effects of sound), 43.60.-c (Acoustic signal processing)
 
@@
 In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}.
-Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
+However, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
 
@@
 Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.
-Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level.
-Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments.
+Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using the variations of the measured sound level only.
+Because these indicators are not representative of the composition of the sound scene, the model underperforms when compared to the predictions computed using perceptual assessments.
 Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes.
@@
 \begin{enumerate}
-  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled.
-  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
+  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{} \ml{mettre sur archive}}.
+  \item Propose numerical means~\footnote{Open source code is available at \url{}.
+\ml{mettre url}} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
   \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
@@
 
-First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate the relevance of the proposed protocol for applications in \textit{in situ} sound environments.
-Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
+First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings in order to validate the relevance of the proposed protocol for \textit{in situ} sound environments.
+Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
 Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.
@@
 For each of the 19 recordings, 45 seconds of audio in a single channel are extracted.
-This duration of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments.
+This duration is chosen to be perceptually relevant for the perceptual evaluation of sound environments.
 The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
@@
 \begin{itemize}
-\item Background sources that are present throughout the whole scene, and their respective sound level considered constant.
+\item Background sources that are present throughout the whole scene, and their respective sound level considered constant,
 \item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
@@
 
-The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}.
-\textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources.
+The 45~s sound scenes are then replicated using the  \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
 Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
 The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
-All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
+The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
 
-\begin{figure}[h]
+\begin{figure*}[th]
     \centering
@@
     \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
-\end{figure}
+\end{figure*}
 
@@
 Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
-Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
+To do so, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
 These statistics include, conditionally to the ambiance (resp.
@@
 Adjustments are made on the variance of the considered properties to better cover possible scenarios.
-Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism.
+Since voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism.
 An additional \textit{square} ambiance is added with properties derived empirically from available data.
@@
 Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way.
-Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled.
-As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space.
-This is shown in Figure~\ref{fig:tvb_pres}.
+To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}.
 The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.
 
-The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.
+% Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative,
+
+The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.
 
@@
 
-A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
+In order to gather estimates of the perceputal attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted.
+Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
 These scales are presented in French and translated in this report using standard terminology.
@@
 \begin{itemize}
-\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
-\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
-\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
-\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
-\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
+\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
+\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
+\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
+\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
+\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
 \end{itemize}
@@
 The perception of source activity can be linked to the notions of dominance, time of presence, or emergence.
-We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study.
+The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study.
 The influence of traffic on pleasantness is further separated into that of background and salient events.
@@
 The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants.
-All participants are first presented with the most quiet then most noisy of the recorded scenes (resp.
+All participants are first presented with the most quiet then loudest of the recorded scenes (resp.
 P3 and P15).
 A random listening order is generated for each subject to control ordering effects for the remaining of the test.
-Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.
+Participants can listen to each scene once, and have to listen to the full extract and to answer all questions before being allowed to proceed to the next scene.
 
@@
 These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
-From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listeners.
 
-A total of 23 students aged from 22 to 23 yo including 16 males and 7 females at Ecole Centrale de Nantes completed the test.
+A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test.
 All gave consent prior to the experiment and reported normal hearing.
@@
 
-The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared.
+The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied.
 Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios.
@@
 Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
-Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.
+Thus, the proposed hypothesis that recorded and replicated scenes generate equivalent perceptual responses cannot be rejected.
 
-\begin{table}[h]
+\begin{table*}[t]
 \centering
@@
 %}
-\end{table}
+\end{table*}
 
@@
 Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows.
-The standard deviation of assessments is represented by ellipses for one location (P1).
-The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}.
+The standard deviation of assessments is represented using ellipses for one location (P1).
+The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{} \ml{mettre url}}.
 This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.
 
-\begin{figure}[h]
+\ml{On vise 12 pages. Tu dois raccourcir les legendes des figures.}
+
+\begin{figure}[th]
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/pca_p1.eps}
+    \includegraphics[width=\columnwidth]{figures/pca_p1.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).
@@
 Arrows show discrepancies between corresponding recorded and replicated scenes.
-For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
+For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.\ml{Pas de couleur, Remplacer par des croix et des points !!
+enlever le blanc autour de la figure, soit a la main, soit en controlant la bounding box}}\label{fig:pspace_rec}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/pca_sim.eps}
+    \includegraphics[width=\columnwidth]{figures/pca_sim.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
@@
 where $L(f, t)$ is the third-octave spectrum of the signal.
-To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
+To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) as an estimate the contribution of birds.
+The 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) is chosen as an estimate the contribution of voices.\\
 
@@
 \begin{itemize}
-\item $L_{eq, s}$: Equivalent sound level for source $s$
-\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined
+\item $L_{eq, s}$: Equivalent sound level for source $s$,
+\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
 \end{itemize}
@@
 
-To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
+Considering machine learning techniques to detect the presence of source requires the availability of annotated data.
+Thus, in order to train deep learning architectures for source recognition, a larger corpus is constructed.
+It is composed of two datasets of simulated scenes:
 \begin{itemize}
@@
 Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.
-Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data.
+% Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data.
 This will not be discussed here as it did not result in better time of presence estimations in our experiments.
@@
 The output of the last block is flattened then goes through a fully connected layer with output size 3.
-A final sigmoid activation is used to obtain outputs in the 0-1 range.
+A final sigmoid activation is used in order to obtain outputs in the 0-1 range.
 The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
@@
 
-\begin{figure}[!h]
+\begin{figure}[t]
     \centering
     \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
-    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
+    \caption{Architecture of the deep learning model used for source presence prediction.\ml{refaire la figure en vertical}}\label{fig:deep_arch}
 \end{figure}
@@
 
-\begin{table}[h]
+\begin{table*}[t]
 \centering
@@
 
-\begin{table}
+\begin{table*}
 \centering
@@
 \label{tab:physc}
-\resizebox{\columnwidth}{!}{
+\resizebox{\textwidth}{!}{
 \begin{tabular}{ c | c c c c c | c c c c }
@@
 }
-\end{table}
+\end{table*}
 
@@
 \hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
-\hat P_{3, \varphi} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} + 5.00~TFSD_{4kHz, 1/8s}
+\hat P_{3, \varphi} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} \\
+& + 5.00~TFSD_{4kHz, 1/8s}
 \end{align}
@@
 
-\begin{table}[t]
+\begin{table*}[t]
 \centering
@@
 For the simulated subcorpus of the experiments all performance metrics are comparable.
-Applying the detection model on the small corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
+Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
 This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.
 
-\begin{table}[t]
+\begin{table*}[t]
 \centering
@@
 %}
-\end{table}
+\end{table*}
 
 Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.
+%Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.
 
@@
 \begin{align}
-\hat P_{3, p} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
-\hat P_{4, p} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
+\hat P_{3, p} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
+& + 0.15~T_{B, p}\\
+\hat P_{4, p} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
+& + 0.09~T_{B, p}
 \end{align}
@@
 
-A perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual soundscape characterization.
+This paper demonstrated the potential of considering deep learning architectures for source presence detection.
+When applied to pleasantness prediction from acoustic indicators in urban environments, such approach leads to state of the art results without the need to design specific estimators for each source that is considered in the prediction model.
+
+Provided enough data is available, the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a wide range of applications for soundscape characterization, categorization, or perceptual mapping.
+
+Also, a perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual soundscape characterization.
 Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.
-
-Subsequently, a deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, compares to previous state-of-the-art models.
-Though, provided enough data the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a range of applications for soundscape characterization, categorization, or perceptual mapping.
 

### 2019-06-11 paper/paperV1.tex
ML Corrections

@@
 }
-\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
+\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
 \maketitle
@@
 
-This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient
-to predict perceived pleasantness.
+This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient to predict perceived pleasantness.
 To do so, a corpus of simulated sound scenes is first designed.
@@
 To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels.
-We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.
+We propose in this paper to study the potential of the use of simulated corpora~\cite{lafay2016, salamon2017} for such purpose.
 
@@
 
-In order to gather estimates of the perceputal attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted.
+In order to gather estimates of the perceptual attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted.
 Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
@@
 These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
-From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listeners.
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.
 
@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_rec} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes.
+Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes.
 The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/pca_p1.eps}
+    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).
-Red dots are projections of the assessments of each scene averaged over subjects.
 Arrows show discrepancies between corresponding recorded and replicated scenes.
-For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.\ml{Pas de couleur, Remplacer par des croix et des points !!
-enlever le blanc autour de la figure, soit a la main, soit en controlant la bounding box}}\label{fig:pspace_rec}
+For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.}\label{fig:pspace_rec}
 \end{figure}
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/pca_sim.eps}
+    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
-Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
+Assessments of simulated scenes are projected as dots, and recorded and replicated scenes are projected as crosses.}\label{fig:pspace_sim}
 \end{figure}
@@
 The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:
-\begin{equation}
+
+\begin{align}
 TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
-\end{equation}
+\end{align}
+
 where $L(f, t)$ is the third-octave spectrum of the signal.
@@
 Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.
-% Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data.
-This will not be discussed here as it did not result in better time of presence estimations in our experiments.
 
@@
 During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
-\begin{equation}
+
+\begin{align}
 BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
-\end{equation}
+\end{align}
+
 where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
-    \caption{Architecture of the deep learning model used for source presence prediction.\ml{refaire la figure en vertical}}\label{fig:deep_arch}
+    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
+    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
 \end{figure}
@@
 	$TFSD_{500Hz, 1s}$ & NS & 0.41** & NS & 0.28** & NS & -0.24* & -0.39** & 0.74** & NS \\
-	$TFSD_{4kHz, 1/8s}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
+	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.45** & -0.54** & NS & 0.63** \\ \hline
 	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
@@
 	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
-	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
-	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
-	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
+	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & NS & 0.35** & -0.57** & -0.29** & 0.56** & 0.81** & -0.39** & -0.37** \\
+	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.35** & NS & NS & -0.39** & 0.81** & NS \\
+	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.46** & -0.57** & NS & 0.91** \\ \hline
 \end{tabular}
@@
 On the present corpus, the best model in terms of $R^2_{adj}$ is:
-\begin{equation}
-\hat P_{1, \varphi} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
-\end{equation}
+
+\begin{align}
+\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
+\end{align}
+
 where $\varphi$ indicates a model from physical variables.
@@
 Both models' coefficients are re-optimized on the studied data for a fair comparison:
+
 \begin{align}
 \hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
-\hat P_{3, \varphi} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} \\
-& + 5.00~TFSD_{4kHz, 1/8s}
+\begin{split}
+\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
+& + 2.99~TFSD_{4kHz, 1/8s}
+\end{split}
 \end{align}
@@
 	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
-	$\hat P_{3, \varphi}$ & 0.90 & 0.78 & 0.89** \\ \hline
+	$\hat P_{3, \varphi}$ & 0.91 & 0.78 & 0.89** \\ \hline
 \end{tabular}
@@
 	Sub-corpus & All & Rec. & Rep. & Sim. & All & Rep. & Sim. \\ \hline
-	RMSE & 0.84 & 1.09 & 0.68 & 0.85 & 0.83 & 0.73 & 0.85 \\ \hline
+	RMSE & 0.84 & 1.09 & 0.68 & 0.85 & 0.83 & 0.72 & 0.86 \\ \hline
 	r & 0.91** & 0.89** & 0.93** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
-	$R^2_{adj}$ & 0.82 & 0.73 & 0.82 & 0.80 & 0.82 & 0.81 & 0.79 \\ \hline
+	$R^2_{adj}$ & 0.82 & 0.73 & 0.82 & 0.80 & 0.82 & 0.79 & 0.79 \\ \hline
 \end{tabular}
@@
 The perceptual models of pleasantness (eq.\ref{eq:pp1}-\ref{eq:pp2}) are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
+
 \begin{align}
-\hat P_{3, p} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
-& + 0.15~T_{B, p}\\
-\hat P_{4, p} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
+\begin{split}
+\hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
+& + 0.15~T_{B, p}
+\end{split}\\
+\begin{split}
+\hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
 & + 0.09~T_{B, p}
+\end{split}
 \end{align}
+
 Several hypotheses can be drawn from this result.

### 2019-06-18 paper/paperV1.tex
Added links

@@
 \date{
-$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
-$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
+$^1$ LS2N, UMR CNRS 6004, Ecole Centrale de Nantes, F-44321, France\\
+$^2$ ETIS, UMR CNRS 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
 $^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
@@
 There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.
-Combined, they allow us to predict important affective dimensions of the soundscape such as its pleasantness.
+Combined, they allow the prediction of important affective dimensions of the soundscape such as its pleasantness.
 However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.
@@
 Perceptual attributes corresponding to those stimuli are gathered through a listening experiment.
-The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.
+The availability of the contributions of the individual sound sources within the simulated sound scenes then allows the training and validate the proposed estimation procedure.
 
-% to yield similar perceptual responses compared to recorded sound scenes.
 PACS numbers: 43.66.Lj (Perceptual effects of sound), 43.60.-c (Acoustic signal processing)
-
 \end{abstract}
@@
 \begin{enumerate}
-  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{} \ml{mettre sur archive}}.
-  \item Propose numerical means~\footnote{Open source code is available at \url{}.
-\ml{mettre url}} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
-  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
+  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
+  \item Propose numerical means~\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
+  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree of accuracy.
 \end{enumerate}
@@
 Adjustments are made on the variance of the considered properties to better cover possible scenarios.
-Since voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism.
+Since voice events consist mostly of read English recordings, their EBR is reduced to improve realism.
 An additional \textit{square} ambiance is added with properties derived empirically from available data.
@@
 The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.
-
-% Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative,
 
@@
 \item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
-\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
+\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
 \end{itemize}
@@
 The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations.
-Beyerdynamics DT-990 Pro headphones are used for all participants.
+Beyerdynamics DT-990 Pro headphones are used by all participants.
 These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
@@
 
-A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test.
+A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test (note that with 23 subjects, the incomplete block design is not perfectly balanced).
 All gave consent prior to the experiment and reported normal hearing.
-With 23 subjects, the incomplete block design is not perfectly balanced.
 
@@
 \label{sec:data_val}
-
-%Participants evaluate each scene on criteria using 11-point Likert scales.
-Many statistical tools assume that the studied data follows continuous, single mode distributions.
-It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments.
-A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus.
-Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal.
-and shown in Figure~\ref{fig:hclusters}.
-No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\
-%
-%\begin{figure}[!h]
-%    \centering
-%    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
-%    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
-%\end{figure}
 
@@
 As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}.
-On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale.
+On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale.
 Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters.
@@
 For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.
-Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
-Thus, the proposed hypothesis that recorded and replicated scenes generate equivalent perceptual responses cannot be rejected.
+Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
 
@@
 \centering
-\caption{Mean differences of perceptual assessments between original and replicated sound scenes.
-Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
+\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
 \label{tab:ogrep}
@@
 
-To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied.
+Next, the perceptual space generated by the experiment's five general scales (P, L, OL, I and C) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity.
 It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants.
@@
 Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes.
-The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.
+These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments.
+This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.
 
@@
 The standard deviation of assessments is represented using ellipses for one location (P1).
-The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{} \ml{mettre url}}.
+The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/index.html}}.
 This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.
-
-\ml{On vise 12 pages. Tu dois raccourcir les legendes des figures.}
 
@@
 
-\begin{align}
-TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
-\end{align}
-
+\begin{equation}
+TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f1, t)}{df1dt}\rvert}
+\end{equation}
 where $L(f, t)$ is the third-octave spectrum of the signal.
-To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) as an estimate the contribution of birds.
-The 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) is chosen as an estimate the contribution of voices.\\
+To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.
 
 The simulation process of the scenes outputs ground truth source contributions as separate channels.
-Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:
-
-\begin{itemize}
-\item $L_{eq, s}$: Equivalent sound level for source $s$,
-\item $\Delta L_{s}$: Source emergence, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
-\end{itemize}
+Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources: The equivalent sound level $L_{eq, s}$ for source $s$ and the source emergence $\Delta L_{s}$, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
 
 Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed.
-$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence $\Delta L_{s}(t, f) = L_s(t, f) - L_{\bar{s}}(t, f)$.
+$\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence spectrum $\Delta L_{s}(t, f)$.
 It is parametrized by $\alpha$ and $\beta$ thresholds:
@@
 The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
-This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.
 
@@
 The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$.
-The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5.
+The convolutional layers have respectively 128, 64, 32 and 32 output channels, and a common kernel size of 5x5.
 The output of the last block is flattened then goes through a fully connected layer with output size 3.
-A final sigmoid activation is used in order to obtain outputs in the 0-1 range.
-The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
+A final sigmoid activation is used in order to obtain outputs in the 0-1 range, which correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
 During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
 
-\begin{align}
+\begin{equation}
 BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
-\end{align}
-
+\end{equation}
 where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
@@
 Only combinations for which all predictors verify $VIF<5$ are considered valid.
-The best resulting models in terms of $R^2_{adj}$ with statistically significant coeffeicient estimates ($p<0.05$) are:
+The best resulting models in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) are:
 
 \begin{align}
-\hat P_{1, p} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p} \label{eq:pp1}\\
-\hat P_{2, p} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p} \label{eq:pp2}
+\begin{split}
+\hat P_{1, p} & = 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p}\\
+&\qquad + 0.14~T_{B, p} \label{eq:pp1}
+\end{split}\\
+\begin{split}
+\hat P_{2, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
+&\qquad + 0.12~T_{B, p} \label{eq:pp2}
+\end{split}
 \end{align}
@@
 $\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness.
-The performance of obtained perceptual models is about 0.6 RMSE, which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).
+The performance of obtained perceptual models is about 0.6 (RMSE), which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).
 
@@
 Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$.
-In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.
+In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a relevant predictor without the need for a source separation and level estimation algorithm.
 
@@
 
-\begin{align}
+\begin{equation}
 \hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
-\end{align}
+\end{equation}
 
@@
 \hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
-& + 2.99~TFSD_{4kHz, 1/8s}
+&\qquad + 2.99~TFSD_{4kHz, 1/8s}
 \end{split}
@@
 For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset.
-To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
-\begin{enumerate}
-\item The 6 recorded scenes, which contain additional sources not present in the datasets simulated for the optimization of the deep learning model,
-\item The 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017},
-\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
-\end{enumerate}
+To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in three parts: the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
+
 Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:results_phys}.
@@
 Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
-%Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.
 
@@
 \hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
-& + 0.15~T_{B, p}
+&\qquad + 0.15~T_{B, p}
 \end{split}\\
@@
 \hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
-& + 0.09~T_{B, p}
+&\qquad + 0.09~T_{B, p}
 \end{split}
@@
 The physical model on this corpus (eq.\ref{eq:pphi}) includes only the contribution of bird sources.
-This result is attributed to the overall low diversity of the studied corpus which gives more importance to this perceived dimension:
-\begin{itemize}
-\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
+This result is attributed to the overall low diversity of the studied corpus which gives more importance to this perceived dimension.
+Indeed, only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
 Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic.
 It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
-\item The diversity of available isolated source extracts for scene simulation is rather low.
+Additionally, the diversity of available isolated source extracts for scene simulation is rather low.
 Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness.
 These properties result in no significant correlation of indicators describing voice events to pleasantness for this corpus.
-\end{itemize}
 Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness.
@@
 Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
-Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha_{opt}, \beta_{opt})$.
+Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels.
 
-A higher pleasantness predictions quality is obtained using detection-based presence labels compared to $\hat T_s(\alpha_{opt}, \beta_{opt})$ on the corpus containing replicated scenes, which feature higher polyphonies and a richer source taxonomy.
+A higher pleasantness prediction quality is obtained using detection-based presence labels compared to $\hat T_s(\alpha_{opt}, \beta_{opt})$ on the corpus containing replicated scenes, which feature higher polyphonies and a richer source taxonomy.
 Although interesting, this result may be attributed to the small sample size as the architecture is not trained on complex scenes.
@@
 When applied to pleasantness prediction from acoustic indicators in urban environments, such approach leads to state of the art results without the need to design specific estimators for each source that is considered in the prediction model.
-
 Provided enough data is available, the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a wide range of applications for soundscape characterization, categorization, or perceptual mapping.

### 2019-06-18 paper/paperV1.tex
Added links

@@
 The standard deviation of assessments is represented using ellipses for one location (P1).
-The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/index.html}}.
+The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/web/index.html}}.
 This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

### 2019-06-18 paper/paperV1.tex
Added links

@@
 \begin{enumerate}
-  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
-  \item Propose numerical means~\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
+  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
+  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
   \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree of accuracy.

### 2019-06-19 paper/submitted/gontierAa2019.tex
submitted version

--- 2019-06-18 paper/paperV1.tex
@@
 \usepackage{graphicx}
+\graphicspath{{figures/}{./}}
 
@@
 
-\bibliographystyle{unsrt}
-\bibliography{refs}
+\begin{thebibliography}{10}
+
+\bibitem{ec2002}
+EC.
+\newblock Directive 2002/49/{EC} of the {E}uropean {P}arliament and of the
+  {C}ouncil of 25 june 2002 relating to the assessment and management of
+  environmental noise.
+\newblock {\em Off. J. Eur. Communities}, 189:12, 2002.
+
+\bibitem{iso2014}
+{ISO 12913-1:2014}.
+\newblock Acoustics - soundscape - part 1: definition and conceptual framework.
+\newblock Standard, International Organization for Standardization, Geneva, CH,
+  2014.
+
+\bibitem{viollon2000}
+S.~Viollon and C.~Lavandier.
+\newblock Multidimensional assessment of the acoustic quality of urban
+  environments.
+\newblock In {\em INTER-NOISE 2000}, 2000.
+
+\bibitem{axelsson2010}
+O.~Axelsson, M.E. Nilsson, and B.~Berglund.
+\newblock A principal components model of soundscape perception.
+\newblock {\em J. Ac. Soc. Am.}, 128:2836, 2010.
+
+\bibitem{cain2013}
+R.~Cain, P.~Jennings, and J.~Poxon.
+\newblock The development and application of the emotional dimensions of a
+  soundscape.
+\newblock {\em Applied Acoustics}, 74:232--239, 2013.
+
+\bibitem{jeon2018}
+J.Y. Jeon, J.Y. Hong, C.~Lavandier, J.~Lafon, O.~Axelsson, and M.~Hurtig.
+\newblock A cross-national comparison in assessment of urban park soundscapes
+  in {F}rance, {K}orea, and {S}weden through laboratory experiments.
+\newblock {\em Applied Acoustics}, 133:107--117, 2018.
+
+\bibitem{aletta2016}
+F.~Aletta, J.~Kang, and O.~Axelsson.
+\newblock Soundscape descriptors and a conceptual framework for developing
+  predictive soundscape models.
+\newblock {\em Landsc. Urban Plan.}, 149:65--74, 2016.
+
+\bibitem{aumond2017}
+P.~Aumond, A.~Can, B.~De Coensel, D.~Botteldooren, C.~Ribeiro, and
+  C.~Lavandier.
+\newblock Modeling soundscape pleasantness using perceptive assessments and
+  acoustic measurements along paths in urban context.
+\newblock {\em Acta Acust. unit. Acust.}, 103:430--443, 2017.
+
+\bibitem{delaitre2014}
+P.~Delaitre, C.~Lavandier, C.~Ribeiro, M.~Quoy, E.~D'Hondt, E.~Gonzalez Boix,
+  and K.~Kambona.
+\newblock Influence of loudness of noise events on perceived sound quality in
+  urban context.
+\newblock In {\em Inter Noise}, 2014.
+
+\bibitem{blauert1997}
+J.~Blauert and U.~Jekosch.
+\newblock Sound quality evaluation: a multi layered problem.
+\newblock {\em Acta Acust. unit. Acust.}, 83:747--53, 1997.
+
+\bibitem{jekosch2004}
+U.~Jekosch.
+\newblock Basic concepts and terms quality, reconsidered in the context of
+  product-sound quality.
+\newblock {\em Acta Acust. unit. Acust.}, 90:999--1006, 2004.
+
+\bibitem{nilsson2007}
+M.E. Nilsson, D.~Botteldooren, and B.~De Coensel.
+\newblock Acoustic indicators of soundscape quality and noise annoyance in
+  outdoor urban areas.
+\newblock In {\em 19th International Congress on Acoustics}, 2007.
+
+\bibitem{perez2012}
+I.~Garc\'ia P\'erez, I.~Aspuru Soloaga, K.~Herranz-Pascual, and
+  I.~Garc\'ia-Borreguero.
+\newblock Validation of an indicator for the assessment of the environmental
+  sound in urban places.
+\newblock In {\em Euronoise}, 2012.
+
+\bibitem{guastavino2007}
+C.~Guastavino.
+\newblock Categorization of environmental sounds.
+\newblock {\em Can. J. Exp. Psychol.}, 61:54--63, 2007.
+
+\bibitem{gygi2007}
+B.~Gygi, G.~R. Kidd, and C.~S. Watson.
+\newblock Similarity and categorization of environmental sounds.
+\newblock {\em Perception \& Psychophysics}, 69:839--55, 2007.
+
+\bibitem{salamon2014}
+J.~Salamon, C.~Jacoby, and J.~P. Bello.
+\newblock A dataset and taxonomy for urban sound research.
+\newblock In {\em 22nd ACM International Conference on Multimedia}, 2014.
+
+\bibitem{brown2011}
+A.L. Brown.
+\newblock Towards standardization in soundscape preference assessment.
+\newblock {\em Applied Acoustics}, 72:387--392, 2011.
+
+\bibitem{lavandier2006}
+C.~Lavandier and B.~Defreville.
+\newblock The contribution of sound source characteristics in the assessment of
+  urban soundscapes.
+\newblock {\em Acta Acust. unit. Acust.}, 92:912--921, 2006.
+
+\bibitem{ricciardi2014}
+P.~Ricciardi, P.~Delaitre, C.~Lavandier, F.~Torchia, and P.~Aumond.
+\newblock Sound quality indicators for urban places in paris cross-validated by
+  {M}ilan data.
+\newblock {\em J. Ac. Soc. Am.}, 138:2337--2348, 2014.
+
+\bibitem{hong2017}
+J.Y. Hong and J.Y. Jeon.
+\newblock Relationship between spatiotemporal variability of soundscape and
+  urban morphology in a multifunctional urban area: A case study in {S}eoul,
+  {K}orea.
+\newblock {\em Building and Environment}, 126:382--95, 2017.
+
+\bibitem{guastavino2006}
+C.~Guastavino.
+\newblock The ideal urban soundscape: Investigating the sound quality of
+  {F}rench cities.
+\newblock {\em Acta Acust. unit. Acust.}, 92:945--51, 2006.
+
+\bibitem{ardouin2018}
+J.~Ardouin, L.~Charpentier, M.~Lagrange, F.~Gontier, N.~Fortin, J.~Picaut,
+  D.~Ecoti\`ere, G.~Guillaume, and C.~Mietlicky.
+\newblock An innovative low-cost sensor for urban sound monitoring.
+\newblock In {\em INTER-NOISE 2018}, 2018.
+
+\bibitem{mydlarz2017}
+C.~Mydlarz, J.~Salamon, and J.P. Bello.
+\newblock The implementation of low-cost urban acoustic monitoring devices.
+\newblock {\em Applied Acoustics}, 117:207--218, 2017.
+
+\bibitem{cakir2015}
+E.~Cakir, T.~Heittola, H.~Huttunen, and T.~Virtanen.
+\newblock Polyphonic sound event detection using multi label deep neural
+  networks.
+\newblock In {\em 2015 International Joint Conference on Neural Networks
+  (IJCNN)}, 2015.
+
+\bibitem{salamon2017-2}
+J.~Salamon and J.~P. Bello.
+\newblock Deep convolutional neural networks and data augmentation for
+  environmental sound classification.
+\newblock {\em IEEE Signal Processing Letters}, 24, 2017.
+
+\bibitem{lafay2016}
+Gr{\'e}goire Lafay, Mathieu Lagrange, Mathias Rossignol, Emmanouil Benetos, and
+  Axel Roebel.
+\newblock {A morphological model for simulating acoustic scenes and its
+  application to sound event detection}.
+\newblock {\em {IEEE/ACM Transactions on Audio, Speech and Language
+  Processing}}, 24(10):1854--1864, 2016.
+
+\bibitem{salamon2017}
+J.~Salamon, D.~MacConnell, M.~Cartwright, P.~Li, and J.~P. Bello.
+\newblock Scaper: A library for soundscape synthesis and augmentation.
+\newblock In {\em 2017 IEEE Workshop on Applications of Signal Processing to
+  Audio and Acoustics (WASPAA)}, 2017.
+
+\bibitem{gloaguen2017}
+J.R. Gloaguen, A.~Can, M.~Lagrange, and J.F. Petiot.
+\newblock Creation of a corpus of realistic urban sound scenes with controlled
+  acoustic properties.
+\newblock In {\em Proceedings of Meetings on Acoustics}, 2017.
+
+\bibitem{panayotov2015}
+V.~Panayotov, G.~Chen, D.~Povey, and S.~Khudanpur.
+\newblock Librispeech: an {ASR} corpus based on public domain audio books.
+\newblock In {\em Proceedings of the 2015 IEEE International Conference on
+  Acoustics, Speech and Signal Processing (ICASSP)}, 2015.
+
+\bibitem{gontier2018}
+F.~Gontier, P.~Aumond, M.~Lagrange, C.~Lavandier, and J.-F. Petiot.
+\newblock Towards perceptual soundscape characterization using event detection
+  algorithms.
+\newblock In {\em Proceedings of the Detection and Classification of Acoustic
+  Scenes and Events 2018 workshop (DCASE)}, 2018.
+
+\bibitem{dagnelie2003}
+P.~Dagnelie.
+\newblock {\em Principes d'exp\'erimentation: planification des exp\'eriences
+  et analyse de leurs r\'esultats}.
+\newblock Presses Agronomiques de Gembloux, 2003.
+
+\bibitem{wilcoxon1945}
+F.~Wilcoxon.
+\newblock Individual comparisons by ranking methods.
+\newblock {\em Biometrics Bulletin}, 1:80--3, 1945.
+
+\bibitem{pratt1959}
+J.~W. Pratt.
+\newblock Remarks on zeros and ties in the {W}ilcoxon signed rank procedures.
+\newblock {\em Journal of the American Statistical Association}, 54:655--67,
+  1959.
+
+\bibitem{itatoolbox2017}
+M.~Berzborn, R.~Bomhardt, J.~Klein, J.G. Richter, and M.~Vorlander.
+\newblock The ita-toolbox: an open source matlab toolbox for acoustic
+  measurements and signal processing.
+\newblock In {\em 43th Annual German Congress on Acoustics}, 2017.
+
+\bibitem{gontier2017}
+F.~Gontier, M.~Lagrange, P.~Aumond, A.~Can, and C.~Lavandier.
+\newblock An efficient audio coding scheme for quantitative and qualitative
+  large scale acoustic monitoring using the sensor grid approach.
+\newblock {\em Sensors}, 17, 2017.
+
+\bibitem{kingma2015}
+D.~P. Kingma and J.~Lei Ba.
+\newblock Adam: a method for stochastic optimization.
+\newblock In {\em Proceedings of the 2015 International Conference on Learning
+  Representations (ICLR)}, 2015.
+
+\end{thebibliography}
 

### 2019-09-18 paper/paperV2.tex
Added V2

--- 2019-06-18 paper/paperV1.tex
@@
 \begin{abstract}
+
 The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces).
@@
 There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.
-Combined, they allow the prediction of important affective dimensions of the soundscape such as its pleasantness.
-However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.
 
-This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient to predict perceived pleasantness.
+This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources.
 To do so, a corpus of simulated sound scenes is first designed.
 Perceptual attributes corresponding to those stimuli are gathered through a listening experiment.
-The availability of the contributions of the individual sound sources within the simulated sound scenes then allows the training and validate the proposed estimation procedure.
+The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure.
+The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.
 
@@
 \label{sec:intro}
-In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
-Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities.
-These maps aim at providing useful tools for urban planning and noise reduction purposes.
-As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.
 
-The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
-Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}.
-In this paper, we will focus on the first affective dimension which is pleasantness.
-Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}.
-While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}.
-In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}.
-However, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.
+%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
+Consequently, the 2002/49/EC European directive~\cite{ec2002} requires the development of noise maps in large cities.
+These maps aim at providing useful tools for noise reduction planning purposes.
+In this context, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as the "acoustic environment as perceived or experienced and/or understood by people, in context" provides a shift from the noise-only paradigm by considering the possible positive influence of some the sound environment's components.
 
-Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost.
-Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.
-Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using the variations of the measured sound level only.
-Because these indicators are not representative of the composition of the sound scene, the model underperforms when compared to the predictions computed using perceptual assessments.
-Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes.
-These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.
-In order to tackle this issue, the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.
+Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}, by identifying its composition in terms of sources of interest from acoustic measurements at a reasonable cost.
+Classical monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment.
+Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}.
+While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline.
+Indeed, it requires large transmission bandwitdh and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved.
+Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information.
+This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}.
+In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes.
+Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomiesn it demonstrates that the information content of third ocrave spectra is sufficient to identify sound sources.
 
-In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}.
-Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings.
-To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels.
-We propose in this paper to study the potential of the use of simulated corpora~\cite{lafay2016, salamon2017} for such purpose.
+The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}.
+A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}.
+Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}.
+Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}.
+In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}.
+The perceived time of presence presents the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.
+
+In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{mesaros2017}.
+In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}.
+Several architectures may be used to such aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}.
+Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}.
+Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}.
+To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity.
+It is possible to manually annotate each recording in the corpus.
+However, this process can be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources.
+As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes.
+By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs.
 
@@
   \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
-  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree of accuracy.
+  \item Demonstrate through a comparative study with state of the art approaches that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy.
 \end{enumerate}
 
-First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings in order to validate the relevance of the proposed protocol for \textit{in situ} sound environments.
-Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
-Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.
+First, a procedure for simulating sound scenes corpora is designed in Section~\ref{sec:data} and a listening test is conducted to justify its relevance in the context of a study on perceived quality.
+A large corpus is constructed and used in Section~\ref{sec:pred} to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
+The resulting time of presence are further applied in Section~\ref{sec:app} to the prediction of pleasantness in urban environments, in comparison to existing models from perceptual and physical indicators.
 
-\section{Perception of simulated sound scenes}
+\section{Corpus}
 \label{sec:data}
 
-\subsection{Corpus generation}
+\subsection{Generation procedure}
 \label{sec:data_corp}
 
-A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models.
-It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France.
+Considering machine learning techniques to detect the presence of sources requires the availability of annotated data.
+Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed.
+This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
+Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality~\cite{lavandier2006}.
+The corresponding database of isolated is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
+\textit{simScene} generates original scenarios using the following information:
+
+\begin{itemize}
+\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
+\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
+\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
+\end{itemize}
+
+Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
+To do so, a corpus of 74 recordings obtained in~\cite{aumond2017} is taken as reference.
+The recordings include 19 locations in the 13th district of Paris, France.
 A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
-For each of the 19 recordings, 45 seconds of audio in a single channel are extracted.
-This duration is chosen to be perceptually relevant for the perceptual evaluation of sound environments.
-The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
-The extracts are then subjectively annotated in terms of the following properties:
+The statistics on background and event properties needed for \textit{simScene} generation are extracted conditionally on the ambiance from annotations of the corpus presented in~\cite{gloaguen2017}, which include:
 
@@
 
-The 45~s sound scenes are then replicated using the  \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
-Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
-The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization.
-The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
+Adjustments are made on the variance of the considered properties to better cover possible scenarios.
+Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism.
+An additional \textit{square} ambiance is added with properties derived empirically from available data.
+Diverse new scenarios are then generated by sampling the background and event properties distributions and used with random samples from the database of isolated occurences to simulate sound scenes.
+At the end of the simulation process each sound scene is associated to a random sound level, which is conditioned to the ambiance according to typical sound levels in urban context.
+
+The simulated deep learning corpus is composed of two subsets:
+\begin{itemize}
+\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
+\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
+\end{itemize}
+
+The generation procedure is the same for both datasets.
+Though, it is important in deep learning applications to ensure a complete independance of the training and test set.
+To do so, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
+As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.
+
+\subsection{Perceptual experiment}
+\label{sec:data_exp}
+
+An experiment in the form a listening test is conducted in order to validate the perceptual correspondance of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.
+
+Given the aim of this test, a separate corpus is constructed.
+First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences.
+These scenes are equally distributed among the five considered ambiances (resp.
+\textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}).
+For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}.
+Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way.
+To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected.
+The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}.
+Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.
 
@@
 
-To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios.
-Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}.
-We choose to restrict the scene composition to these 3 sources.
-Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
-To do so, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}.
-These statistics include, conditionally to the ambiance (resp.
-\textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):
+In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes.
+Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments.
+For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted.
+The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
+The annotations obtained in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}.
+Furthermore, the original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
+The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
 
-\begin{itemize}
-\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
-\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions with parameters identified from annotations of recordings,
-\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
-\end{itemize}
+The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.\\
 
-Adjustments are made on the variance of the considered properties to better cover possible scenarios.
-Since voice events consist mostly of read English recordings, their EBR is reduced to improve realism.
-An additional \textit{square} ambiance is added with properties derived empirically from available data.
-Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes.
-For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}.
-Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way.
-To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}.
-The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.
-
-The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.
-
-\subsection{Perceptual experiment}
-\label{sec:data_exp}
-
-In order to gather estimates of the perceptual attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted.
-Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
+During the test, participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
 These scales are presented in French and translated in this report using standard terminology.
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.
-
-The perception of source activity can be linked to the notions of dominance, time of presence, or emergence.
-The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study.
-The influence of traffic on pleasantness is further separated into that of background and salient events.
+The perceived time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study.
+The influence of traffic on perception is further separated into that of background and salient events.
 Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
@@
 Beyerdynamics DT-990 Pro headphones are used by all participants.
-These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise.
+The calibration of the headphone was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the sound pressure at the entrance of a blocked ear canal.
+To do so, small DPA 4060 microphones have been set at the entrance of the ear canals of a human subject~\cite{moller1992}.
 From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.
@@
 
-\subsection{Corpus validation}
+\subsection{Simulated corpus validation}
 \label{sec:data_val}
@@
 
-\section{Methods}
-\label{sec:methods}
-
 \subsection{Acoustical indicators for soundscape description}
-\label{sec:methods_inds}
+\label{sec:data_inds}
 
 Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters.
-These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.
+These indicators may be used to automatically annotate sound scenes in the design of a deep learning architecture for sound source recognition.
 
@@
 \begin{equation}
-TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f1, t)}{df1dt}\rvert}
+TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f_1=31.5Hz}^{f_1=16kHz}\lvert\frac{d^2L(f_1, t)}{df_1dt}\rvert}
 \end{equation}
 where $L(f, t)$ is the third-octave spectrum of the signal.
-To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.
+This indicator represents the variations in both the time and frequency dimensions to highlight sources of interest.
+For example, bird activity is characterized by narrow-band energy with fast paced variations in time, which translates into high $TFSD$ values in the corresponding frequency range.
+As a baseline indicator for source activity, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.
 
@@
 
-\begin{align}
+\begin{align}\label{eq:ts}
 \hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
@@
 
-where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene.
+where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene.
 This indicator evaluates the presence or absence of a given source in a small time frame.
@@
 
-\subsection{Source presence detection using deep learning}
-\label{sec:methods_deep}
-
-Considering machine learning techniques to detect the presence of source requires the availability of annotated data.
-Thus, in order to train deep learning architectures for source recognition, a larger corpus is constructed.
-It is composed of two datasets of simulated scenes:
-\begin{itemize}
-\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
-\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
-\end{itemize}
-
-The generation procedure is the same for both datasets, and is equivalent to that of the 75 scenes in the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}.
-Only traffic, human voices and birds sources compose the sound scenes.
-To generate the scenes, it is important to ensure a complete independance of the training and test set.
-For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.
-
-The developped model should extract relevant source information from a representation of the audio signal.
-Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal.
-Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}.
-Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range.
-Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception.
-The spectral blocks of dimension 29x8 are then processed independently in this study.
-
-Target outputs are associated to each input frame to train the model.
-In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator.
-Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels.
-Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.
-
-A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}.
-The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$.
-The convolutional layers have respectively 128, 64, 32 and 32 output channels, and a common kernel size of 5x5.
-The output of the last block is flattened then goes through a fully connected layer with output size 3.
-A final sigmoid activation is used in order to obtain outputs in the 0-1 range, which correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
-During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
-
-\begin{equation}
-BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
-\end{equation}
-where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
-This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.
-During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher.
-The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
-    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
-\end{figure}
-
-\section{Results}
-\label{sec:results}
-
-The proposed method for predicting the time of presence of sources is evaluated in the context of pleasantness estimation.
-To this aim, baseline models are first constructed using data from the experiment in Section~\ref{sec:data_exp} and from previous work.
-
-\subsection{Perceptual pleasantness model}
-\label{sec:results_perc}
-
-\begin{table*}[t]
-\centering
-\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
-\label{tab:percc}
-%\resizebox{\columnwidth}{!}{
-\begin{tabular}{ c | c c c c c c c c c }
-\hline
-	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
-	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
-	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
-	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
-	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
-	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
-	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
-	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
-	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
-	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
-\end{tabular}
-%}
-\end{table*}
-
-We first study the relations between perceptual scales with respect to existing work.
-Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
-The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions.
-Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
-In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
-This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
-Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies.
-However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.
-
-Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42).
-The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.
-
-As a baseline, multilinear models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100).
-All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
-Only combinations for which all predictors verify $VIF<5$ are considered valid.
-The best resulting models in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) are:
-
-\begin{align}
-\begin{split}
-\hat P_{1, p} & = 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p}\\
-&\qquad + 0.14~T_{B, p} \label{eq:pp1}
-\end{split}\\
-\begin{split}
-\hat P_{2, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
-&\qquad + 0.12~T_{B, p} \label{eq:pp2}
-\end{split}
-\end{align}
-
-Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
-A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.
-
-Table~\ref{tab:percm} summarizes performance metrics computed for the two models.
-$\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness.
-The performance of obtained perceptual models is about 0.6 (RMSE), which is below the mean standard deviation of pleasantness assessments for this experiment (1.77).
-
-\begin{table}[t]
-\centering
-\caption{Performance of perceptual models for pleasantness prediction (**: p<0.01).}
-\label{tab:percm}
-%\resizebox{\columnwidth}{!}{
-\begin{tabular}{ c | c | c | c }
-\hline
-	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
-	$\hat P_{1, p}$ & 0.59 & 0.91 & 0.95** \\
-	$\hat P_{2, p}$ & 0.61 & 0.90 & 0.95** \\ \hline
-\end{tabular}
-%}
-\end{table}
-
-\subsection{Physical pleasantness models}
-\label{sec:results_phys}
-
-The analysis of physical indicators presented in Section~\ref{sec:methods_inds} is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
-The 6 recorded scenes are excluded from the study as the ground truth source composition is unknown.
-Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.
+The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception.
+The analysis is performed using arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
+The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown.
+Thus, these 6 scenes are excluded of the study.
+Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.
 
@@
 Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources.
-The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim.
+The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceived counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim.
 They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}.
 They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus.
-Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$.
-In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a relevant predictor without the need for a source separation and level estimation algorithm.
+Thus, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicator can be used to annotate simulated sound scenes in terms of estimated perceived time of presence for the three sources.
 
@@
 
-Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations.
-As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
-On the present corpus, the best model in terms of $R^2_{adj}$ is:
+\section{Source time of presence prediction}
+\label{sec:pred}
+
+\subsection{Deep learning model for presence prediction}
+\label{sec:pred_deep}
+
+A deep learning model is implemented using the Python Pytorch~\cite{paszke2017} framework and trained on the corpus presented in Section~\ref{sec:data_corp} for source time of presence prediction.
+
+The developped model should extract relevant source information from a representation of the audio signal.
+Typically, spectral representations are preferred to the raw audio waveform because of the regularities they underline in the signal.
+Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}.
+Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range as the input signal representation.
+Instead of a regression task where the output is directly the time of presence, a multiple label classification task on 1~s texture frames is preferred as the resulting training procedure is easier.
+Individual inputs are thus obtained by splitting the resulting spectrograms into texture windows of 1~s duration.
+The spectral blocks of dimension 29x8 are then processed independently by the model.
+
+Ground truth target outputs are associated to each input frame to train the model.
+Considering the results discussed in Section~\ref{sec:data_inds}, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ time of presence estimation seems well suited to automatically label the dataset for this task.
+Binary presence values on 1~s frames obtained during the $\hat T_s(\alpha_{opt}, \beta_{opt})$ computation before averaging in eq.\ref{eq:ts} are thus used as individual weak labels.
+
+The architecture of the model is shown in Figure~\ref{fig:deep_arch}.
+The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$.
+The convolutional layers have respectively 128, 64, 32 and 32 output channels, and a common kernel size of 5x5.
+The output of the last block is flattened then goes through a fully connected layer with output size 3.
+A final sigmoid activation is used in order to obtain outputs in the 0-1 range, which correspond to the presence of traffic, voices and birds in the 1~s frame respectively.
+During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
 
 \begin{equation}
-\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
+BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
 \end{equation}
+where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range.
+This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.
+During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher.
+The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.
 
-where $\varphi$ indicates a model from physical variables.
-Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}.
-This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
+    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
+\end{figure}
 
-This model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal.
-Both models' coefficients are re-optimized on the studied data for a fair comparison:
+\subsection{Results}
+\label{sec:pred_res}
 
-\begin{align}
-\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
-\begin{split}
-\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
-&\qquad + 2.99~TFSD_{4kHz, 1/8s}
-\end{split}
-\end{align}
-
-Table~\ref{tab:physm} compares the performance of the three models.
-Expectedly, $\hat P_{1, \varphi}$ outperforms both baselines, though a validation on a different corpus would be necessary to conclude on its capabilities.
-Compared to the best perceptual models in Table~\ref{tab:percm} its $R^2_{adj}$ is about 9\% lower.
-Its RMSE is 0.83, which is also high compared to perceptual baselines, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 11-point scale in this study.
-Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.
-
-\begin{table}[t]
-\centering
-\caption{Performance of physical models for pleasantness prediction (**: p<0.01).}
-\label{tab:physm}
-%\resizebox{\columnwidth}{!}{
-\begin{tabular}{ c | c | c | c }
-\hline
-	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
-	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
-	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
-	$\hat P_{3, \varphi}$ & 0.91 & 0.78 & 0.89** \\ \hline
-\end{tabular}
-%}
-\end{table}
-
-\subsection{Time of presence prediction through deep learning}
-\label{sec:results_deep}
-
-The performance of the source detection deep learning architecture presented in Section~\ref{sec:methods_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}.
+The performance of the source detection deep learning architecture presented in Section~\ref{sec:pred_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}.
 The overall presence detection accuracy is 92.11\%.
@@
 
-\begin{table*}[t]
+\begin{table*}[h!]
 \centering
@@
 
-Next, the capacity of these predictions to give relevant pleasantness estimates is studied.
-For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset.
-To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in three parts: the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
+\section{Application to pleasantness prediction}
+\label{sec:app}
 
-Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:results_phys}.
+In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}.
+Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
+In this context, the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.
+
+\subsection{Models of pleasantness}
+\label{sec:app_mdls}
+
+\begin{table*}[t]
+\centering
+\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
+\label{tab:percc}
+%\resizebox{\columnwidth}{!}{
+\begin{tabular}{ c | c c c c c c c c c }
+\hline
+	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
+	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
+	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
+	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
+	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
+	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
+	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
+	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
+	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
+	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
+\end{tabular}
+%}
+\end{table*}
+
+Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:data_exp}.
+We first study the relations between perceptual scales with respect to existing work.
+Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
+The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions.
+Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
+In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
+This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
+Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies.
+However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.
+Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42).
+The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.
+
+Multilinear regression models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100).
+All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
+Only combinations for which all predictors verify $VIF<5$ are considered valid.
+The best resulting models in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) are:
+
+\begin{align}
+\begin{split}
+\hat P_{1, p} & = 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p}\\
+&\qquad + 0.14~T_{B, p} \label{eq:pp1}
+\end{split}\\
+\begin{split}
+\hat P_{2, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
+&\qquad + 0.12~T_{B, p} \label{eq:pp2}
+\end{split}
+\end{align}
+
+Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
+A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.
+
+Similarly, multilinear regression models of pleasantness are constructed using acoustical indicators described in Section~\ref{sec:data_inds}.
+Namely the $L_{50}$ and $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicators are used as predictors, as they display the highest correlation values with the overall loudness and time of presence of sources respectively.
+Again, all combinations of the proposed physical variables are considered, and a VIF check ($VIF<5$) is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
+On the present corpus, the best model in terms of $R^2_{adj}$ is:
+
+\begin{equation}
+\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
+\end{equation}
+
+where $\varphi$ indicates a model from physical variables.
+Indicators related to traffic and voices activity are absent compared to the perceptual models.
+This is consistent with the results underlined in Table~\ref{tab:physc}: the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
+These results can be attributed to the diversity of the studied corpus.
+Indeed, only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
+Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic.
+It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
+Additionally, the diversity of available isolated source extracts for scene simulation is rather low.
+Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness.
+These properties result in no significant correlation of indicators describing voice events to pleasantness for this corpus.
+
+The $\hat P_{1, \varphi}$ model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal.
+Both models' coefficients are re-optimized on the studied data for a fair comparison:
+
+\begin{align}
+\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
+\begin{split}
+\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
+&\qquad + 2.99~TFSD_{4kHz, 1/8s}
+\end{split}
+\end{align}
+
+Table~\ref{tab:pleasm} summarizes the performance of the two perceptual and three physical models.
+$\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness.
+The performance of obtained perceptual models is about 0.6 (RMSE), which is below the average standard deviation of pleasantness assessments for this experiment: 1.77 on a 11-point scale.
+$\hat P_{1, \varphi}$ outperforms both $\hat P_{2, \varphi}$ and $\hat P_{3, \varphi}$, though a validation on a different corpus would be necessary to conclude on its capabilities.
+Compared to the models from perceptual parameters its $R^2_{adj}$ is about 9\% lower.
+Its RMSE is 0.83, which is also high compared to perceptual baselines, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 in this study.
+Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.
+
+\begin{table}[h]
+\centering
+\caption{Performance of baseline models for pleasantness prediction (**: p<0.01).}
+\label{tab:pleasm}
+%\resizebox{\columnwidth}{!}{
+\begin{tabular}{ c | c | c | c }
+\hline
+	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
+	$\hat P_{1, p}$ & 0.59 & 0.91 & 0.95** \\
+	$\hat P_{2, p}$ & 0.61 & 0.90 & 0.95** \\ \hline
+	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
+	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
+	$\hat P_{3, \varphi}$ & 0.91 & 0.78 & 0.89** \\ \hline
+\end{tabular}
+%}
+\end{table}
+
+\subsection{Prediction using deep learning}
+\label{sec:app_deep}
+
+The proposed deep learning method is applied to the perceptual experiment corpus to obtain estimations for the time of presence of traffic, voices and birds.
+These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}.
+To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
+
+Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}.
 Thus, only the outputs corresponding to the presence of birds are used.
@@
 Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training.
-This indicates that the detection model generalizes well to new content and sources.
+This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation.
 For the simulated subcorpus of the experiments all performance metrics are comparable.
@@
 
-Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal.
+Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:app_mdls}.
+This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
+Since the labels are extracted from a masking model approximating the perceived time of presence, they can be considered as "weak" labels.
+Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels.
 
@@
 
-The perceptual models of pleasantness (eq.\ref{eq:pp1}-\ref{eq:pp2}) are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
+This study shows the potential of deep learning architectures in combination with corpora of simulated sound scenes for the perceptual characterization of sound environments.
+With the rich additional information about the source composition available for such corpora, new indicators are computed that outperform existing ones in their relation to the perceived time of presence of sources.
+Training a deep learning architecture on a large corpus of simulated scenes automatically annotated using these indicators then allows for the prediction of the time of presence in new recordings, where information about the contribution of each source is not available.
+The resulting predictions can be applied to the estimation of descriptors of soundscape perception such as the pleasantness.
 
-\begin{align}
-\begin{split}
-\hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
-&\qquad + 0.15~T_{B, p}
-\end{split}\\
-\begin{split}
-\hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
-&\qquad + 0.09~T_{B, p}
-\end{split}
-\end{align}
+Though, the performance of the trained model and its capacity to generalize to recorded data rely on several conditions.
+First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments.
+According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy.
+This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment.
+The perceptual models on the listening test corpus are similar to those found in previous studies with both textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}.
+Second, the quality of predictions is bound by that of annotations.
+Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test.
+Furthermore, the indicator is computed using two parameters optimized on available subjective data, though it was found to be quite robust by using a cross-validation scheme during its optimization.
+While the quality of predictions from the model are encouraging, testing on a larger corpus is required to fully assess its generalization capabilities.
 
-Several hypotheses can be drawn from this result.
-First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the \textit{ex situ} setting of the experiment.
-However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for \textit{in situ} models.
-This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.
+Provided enough excerpts of isolated source occurences are available, the methodology proposed in this paper can be extended to include additional sound sources as the scene simulation process, the time of presence indicator and the deep learning model are all independant from the considered taxonomy.
+Of course, the complexity of the learning process scales with the number of considered simultaneously active sources.
+Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra.
+To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.
 
-The physical model on this corpus (eq.\ref{eq:pphi}) includes only the contribution of bird sources.
-This result is attributed to the overall low diversity of the studied corpus which gives more importance to this perceived dimension.
-Indeed, only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
-Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic.
-It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
-Additionally, the diversity of available isolated source extracts for scene simulation is rather low.
-Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness.
-These properties result in no significant correlation of indicators describing voice events to pleasantness for this corpus.
-Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness.
-The trained deep learning architecture already displays only few detection errors for these sources, and can be extended to any additional source provided enough data is available.
-
-Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels on the simulated corpus.
-This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
-Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels.
-Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels.
-
-A higher pleasantness prediction quality is obtained using detection-based presence labels compared to $\hat T_s(\alpha_{opt}, \beta_{opt})$ on the corpus containing replicated scenes, which feature higher polyphonies and a richer source taxonomy.
-Although interesting, this result may be attributed to the small sample size as the architecture is not trained on complex scenes.
-The detection model is thus quite robust to scenes with additional sources as perturbators, but a study on a larger corpus would be required to validate this conclusion.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-This paper demonstrated the potential of considering deep learning architectures for source presence detection.
-When applied to pleasantness prediction from acoustic indicators in urban environments, such approach leads to state of the art results without the need to design specific estimators for each source that is considered in the prediction model.
-Provided enough data is available, the proposed protocol can be extended to a larger taxonomy to refine perceptual models, and included in a wide range of applications for soundscape characterization, categorization, or perceptual mapping.
-
-Also, a perceptual experiment was conducted that validates the use of corpora of simulated sound scenes in the context of perceptual soundscape characterization.
-Even with newly generated scenarios, a reduced source taxonomy does not significantly affect the relations between abstract or content-related subjective descriptors.
-
-Future work will focus on the robustness of the source detection model, which heavily depends on the quantity and quality of data available for scene simulation.
-Together with additional recordings, data augmentation techniques for machine learning will be studied to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.
-Furthermore, this methodology can be adapted to specific environments by extending or refining the source taxonomy in both the scene simulation process and detection model.
-For example, additional transport-related sources such as train and plane activity could be introduced, and a difference could be made depending on voice expressivity or bird species.
+Future work will focus on studying the robustness of the proposed prediction scheme to a refined sound sources taxonomy as well as its application to a large scale sensor network.
 

### 2019-09-18 paper/paperV2.tex
ml proof v2

@@
 
-The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces).
+The impact of urban sound on human beings has often been extensively studied from a negative point of view (sound pollution).
+In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces).
 The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments.
@@
 
-This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources.
-To do so, a corpus of simulated sound scenes is first designed.
+This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources at a sufficient accuracy.
+To validate this assertion, a corpus of simulated sound scenes is first designed.
 Perceptual attributes corresponding to those stimuli are gathered through a listening experiment.
@@
 
-Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}, by identifying its composition in terms of sources of interest from acoustic measurements at a reasonable cost.
-Classical monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment.
+Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}.
+This requires the identification of its composition in terms of sources of interest from acoustic measurements at a reasonable cost.
+Most monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment.
 Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}.
 While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline.
-Indeed, it requires large transmission bandwitdh and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved.
+Indeed, it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved.
 Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information.
@@
 In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes.
-Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomiesn it demonstrates that the information content of third ocrave spectra is sufficient to identify sound sources.
+Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomies, it demonstrates that the information content of third octave spectra is sufficient to identify sound sources.
 
@@
 In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}.
-The perceived time of presence presents the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.
+The perceived time of presence has the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.
 
-In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{mesaros2017}.
+In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which has demonstrated state-of-the-art performance in many tasks studied in the machine listening community~\cite{mesaros2017}.
 In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}.
-Several architectures may be used to such aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}.
+Several architectures may be used to this aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}.
 Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}.
@@
 It is possible to manually annotate each recording in the corpus.
-However, this process can be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources.
+However, this process would be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources.
 As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes.
 By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs.
+Thanks to this data generation procedure, we are able to validate the proposed approach.
 
@@
   \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
-  \item Demonstrate through a comparative study with state of the art approaches that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy.
+  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy through a comparative study with state of the art approaches.
 \end{enumerate}
 
-First, a procedure for simulating sound scenes corpora is designed in Section~\ref{sec:data} and a listening test is conducted to justify its relevance in the context of a study on perceived quality.
-A large corpus is constructed and used in Section~\ref{sec:pred} to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
-The resulting time of presence are further applied in Section~\ref{sec:app} to the prediction of pleasantness in urban environments, in comparison to existing models from perceptual and physical indicators.
+The procedure for simulating sound scenes corpora is presented in Section~\ref{sec:data} with the outcomes of the listening test conducted to justify the relevance of the sound scenes corpora in the context of a study on perceived quality.
+A large corpus is constructed and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales, as presented in Section~\ref{sec:pred}.
+The application of the resulting time of presence to the prediction of pleasantness in urban environments is then presented in Section~\ref{sec:app}.%The, in comparison to existing models from perceptual and physical indicators.
 
@@
 An additional \textit{square} ambiance is added with properties derived empirically from available data.
-Diverse new scenarios are then generated by sampling the background and event properties distributions and used with random samples from the database of isolated occurences to simulate sound scenes.
-At the end of the simulation process each sound scene is associated to a random sound level, which is conditioned to the ambiance according to typical sound levels in urban context.
+Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes.
+At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.
 
@@
 
-An experiment in the form a listening test is conducted in order to validate the perceptual correspondance of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.
+An experiment in the form a listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.
 
 Given the aim of this test, a separate corpus is constructed.
-First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences.
+First, 200 simulated scenes of 45s each with diverse new scenarios are generated with the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences.
 These scenes are equally distributed among the five considered ambiances (resp.
@@
 The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception.
-The analysis is performed using arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
+The analysis is performed using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
 The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown.
@@
 Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.
+TODO implementation : more details needed
 
@@
 
-\section{Source time of presence prediction}
+\section{Prediction of time of presence of sources}
 \label{sec:pred}
 
-\subsection{Deep learning model for presence prediction}
+\subsection{Deep learning for presence prediction}
 \label{sec:pred_deep}
@@
 This is expected given the spectral content related to these sources.
-Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model.
+Traffic is usually located in lower frequency bands than voice or bird events, but may contain high frequency components mistaken for birds by the model.
 The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.
@@
 Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
-In this context, the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.
+To demonstrate the potential of the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.
 
@@
 Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:data_exp}.
-We first study the relations between perceptual scales with respect to existing work.
+We first study the relationships between perceptual scales with respect to existing work.
 Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
-The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions.
+The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both for the relations to source contributions and between general scales, as discussed in Section~\ref{sec:data_val}.
 Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
-In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered.
+In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered in this study.
 This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
@@
 These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}.
-To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
+To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
 
@@
 Thus, only the outputs corresponding to the presence of birds are used.
-The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred}.
+TODO The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred} TODO reformuler.
 First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
@@
 This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment.
-The perceptual models on the listening test corpus are similar to those found in previous studies with both textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}.
+The perceptual models on the listening test corpus are similar to those found in previous studies with both \textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}.
 Second, the quality of predictions is bound by that of annotations.
@@
 
-Provided enough excerpts of isolated source occurences are available, the methodology proposed in this paper can be extended to include additional sound sources as the scene simulation process, the time of presence indicator and the deep learning model are all independant from the considered taxonomy.
+The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available.
+Indeed, the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy.
 Of course, the complexity of the learning process scales with the number of considered simultaneously active sources.
@@
 To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.
+TODO cite may be paper of dcase on data augmentation ?
 
-Future work will focus on studying the robustness of the proposed prediction scheme to a refined sound sources taxonomy as well as its application to a large scale sensor network.
+Future work will thus focus on studying the robustness of the proposed prediction scheme to a refined sound sources taxonomy as well as its application to a large scale sensor network.
 

### 2019-09-19 paper/paperV2.tex
FG proof v2

@@
 
-The impact of urban sound on human beings has often been extensively studied from a negative point of view (sound pollution).
+The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution).
 In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces).
@@
 Thus, these 6 scenes are excluded of the study.
-Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.
-TODO implementation : more details needed
+Two simulated scenes contained only one source, leading to infinite source emergences.
+Statistics including these scenes cannot be computed and they are thus excluded in this analysis, which as a result includes $n=92$ scenes.
 
@@
 Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
-To demonstrate the potential of the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.
+To demonstrate the potential of the proposed method for predicting the time of presence of sources, it is applied to pleasantness estimation.
 
@@
 Thus, only the outputs corresponding to the presence of birds are used.
-TODO The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred} TODO reformuler.
+Table~\ref{tab:pppred} presents the performance of pleasantness estimations using outputs from the deep learning architecture compared to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels computed with eq.~\ref{eq:ts} on separated source-specific channels.
 First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
@@
 Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra.
-To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching.
-TODO cite may be paper of dcase on data augmentation ?
+To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.
 

### 2019-10-02 paper/coverLetter.tex
CL/PA proof + Cover

\documentclass[10pt]{article}

\title{Reply to reviewers concerning submission AAA-D-19-00080: "Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques"}

\begin{document}

\maketitle

As a preamble, we would like to thank the editor and the reviewer for their comments and suggestions. Following these comments, we made several changes to the article including a modification of the structure of the document, which are summarized here. The next sections list our answers to each of the reviewer’s comments, with references to the revised manuscript (page, column, and paragraph) where appropriate.

\section{Answers to Reviewer I}

\begin{enumerate}

\item \emph{My biggest concern with this work is its generalization. By training on a dataset derived from only 19 recordings, and evaluating on only 6 recordings that weren't replicated in the synthetic data, it's unconvincing that this model would generalize to new data.}

$\rightarrow$ Statistics used to simulate sound scenes are based on 74 recordings of which the 19 used for replication are a subset. Section 2.1 has been rewritten to describe more extensively this reference database. To generate more diverse scenarios the variance of all statistics is further increased. However, we acknowledge that the model needs to be evaluated on a more complete and representative set of recordings to assess fully its generalization capabilities.

\item \emph{It's even less convincing considering that in the final model, only the presence of birds was shown to be predictive of pleasantness, when in large cities like NYC, construction, loud music, dogs, and overnight delivery trucks are some of the most complained about sources of noise, none of which are even included in this model. Therefore, while I don't believe that the authors approaches and analyses are flawed, I think the datasets they are basing this research on are very limited and not representative of urban acoustics. [...]\\
Section 4.2 - Seems pretty odd that only birds were used in the model, no?}

$\rightarrow$ In this paper, we chose to focus on the three main categories of sources that are known to impact soundscape quality. When applying the proposed approach to a specific environment, the taxonomy of sound sources needs to be extended accordingly. This is now discussed in Section 5. Also, as is now emphasized in the Introduction and Section 2 the present study considers the point of view of pedestrians, which is vastly different from that of residents in terms of sources of interest.\\
The quality of the pleasantness model obtained from time of presence approximations is bound by the listening test corpus, and thus mostly by the contents of simulated scenes. Section 4.1 discusses possible causes to the absence of other sources in the model. However, this model is not intended as a contribution but rather as a intermediary to show that presence predictions from the deep learning architecture are accurate enough to estimate more abstract perceptual notions.

\item \emph{Title - The perceived time of presence of sources is an intermediate goal, no? Much of this paper is focused on investigating pleasantness. Possibly change the title to match this goal}

$\rightarrow$ The main objective of this work is the prediction of the perceived time of presence of sources of interest for the perceptual characterization of soundscapes. We study the estimation of pleasantness as an application, but the developped method could also be used to predict other quality descriptors or in mapping applications (eg.~\cite{lavandier2016}). The structure of the paper has been rearranged to emphasize this aim.

\item \emph{Section 2.1 - Each of the recordings from the 19 locations are 45 seconds long? Including the 6 recordings that weren't replicated? This could be made clearer.}

$\rightarrow$ All 100 sound scenes (75 simulated, 19 replicated and 6 recorded) in the listening test corpus are 45~s long. For the replicated and recorded scenes, the 45~s extracts are parts of longer recordings. This is now more clearly stated in Section 2.2.

\item \emph{Section 3.1 - The motivation and description of the computation of the time of presence approximation from the synthetic soundscapes could have been clearer. I think you should emphasize that this computation requires having separated tracks like in the synthetic soundscapes. And then also emphasizing the optimization of alpha and beta such that these measures match the perceptual measures as closely as possible. This all didn't quite make sense until the next section in which I realized they were going to be used as the training data for the ML model.}

$\rightarrow$ These points are now more clearly explained in Section 2.4, which is dedicated to the evaluation of indicators to be used in the automatic annotation of the simulated dataset for deep learning.

\item \emph{Section 3.1 - Was there any overall background noise, e.g. brown noise or "urban rumble", added to the recordings to lower the SNR to something more realistic?}

$\rightarrow$ In this study no background noise was added in a controlled manner. However, some background noise is present in the isolated samples used to simulate sound scenes. Particularly, each scene contains at least one background source for which recordings obtained on Freesound are used. These recordings were selected to be representative of the source but were not cleaned through pre-processing, leading to (uncontrolled) background noise being present to an extent in each simulated scene.

\item \emph{Section 3.2 - 125 ms is quite a large hop size. Since you are using synthetic data anyway (and thus are not very data limited), consider a smaller hop size, as is typical in most machine listening literature (e.g. 5 - 20 ms). Especially considering that you are detecting birds which are quite short sound events.}

$\rightarrow$ This work is placed in the context of a monitoring application through sensor networks in urban areas, for which using features computed on fine time scales may not be viable in the long term due to storage and privacy concerns. Though information on smaller time scales would be more appropriate to detect sound events (from a few milliseconds for birds to tens of milliseconds for speech), this motivates our choice to operate on the third-octave spectrum with 125ms hop size, which is typical in monitoring studies. This is now discussed in the introduction.

\end{enumerate}

\section{Answers to Reviewer II}

\begin{enumerate}

\item \emph{The abstract should be slightly revised; there is too much "background" info, while the reader might want to see already something about (quantifiable) results here.}

$\rightarrow$ The abstract has been revised to reduce background information in favor of the description and results of the current study.

\item \emph{The first two paragraphs of the Intro might be re-thought, they are not very well connected and do not add much to the narrative to frame the issue under investigation. I would expect a more extensive literature review about sound sources recognition algorithms and auditory scenes characterization methods which is currently missing.}

$\rightarrow$ The introduction has been rewritten to better introduce the objectives of this study, as well as related work in machine learning, monitoring applications, and soundscape perception.

\item \emph{The claim that the 45-second excerpt is perceptually relevant should be supported by some evidence/reference somehow.}

$\rightarrow$ Many studies dealing with the environmental acoustic quality are associated with duration of acoustic measurements ranging from few seconds for the shorter~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}. Due to the limited duration of perceptual tests because of the fatigue, stimuli between 30s and 1mn are often prefered for laboratory tests.

\item \emph{The process of manual annotation of the excerpts should be described in more detail to make it replicable.}

$\rightarrow$ The 74 scenes in the reference corpus were obtained as part of the GRAFIC project~\cite{aumond2017} and manually annotated in~\cite{gloaguen2017}. These annotations are reused in this paper, as is now stated more clearly in Section 2.1.

\item \emph{In general, references should be added when claims are made, it is not suitable to leave to the reader the task of tracking back such info. E.g., lines 176-180}

$\rightarrow$ References have been added throughout the paper when applicable.

\item \emph{Some more details could be provided about the procedure and technical equipment used for the perceptual experiment.}

$\rightarrow$ The description of the equipment used in the listening test and its calibration has been extended in Section 2.2.

\item \emph{When applicable, details of the ethical review process should be reported.}

$\rightarrow$ Following ethical procedures, the participants were informed on this study via written information, and written consent was obtained for all of the participants. The present study does not require research ethics approval according to French law on ethical review of research involving humans. It does not concern specific health problem, any vulnerable groups or sensitive issues, it does not involve any method that aims to affect the participants physically or mentally, nor is it associated with any risks beyond those of everyday life.

\item \emph{Some concepts are introduced but not always clearly defined. For instance - prominence (is it proximity/intensity?), emergence (is it signal-to-noise ratio?), etc. More aspects are mentioned in section 3, but the reader might want to know more already at this stage, and it would be desirable to discuss them more extensively.}

$\rightarrow$ Concepts such as source dominance discussed in this paper are subjective descriptors of source activity, and as such have no clearly defined relation to the acoustical properties of the audio signal.

\item \emph{In Section 5 I was longing for more discussion about possible challenges and issues one might face when the complexity of the auditory scenes increases (e.g., more sources), and when other environmental non-acoustic factors (e.g., match between auditory and visual settings) come into play, which could not be detected within the proposed framework.}

$\rightarrow$ Section 5 now discusses the generalization and extension perspectives of the proposed approach in more detail. Its adaptation to new sound sources or environments is straightforward on the condition that enough data (both isolated samples of new sources and reference scenarios) can be gathered. As non-acoustic factors, visual settings are known to have an influence on soundscape perception and quality (see~\cite{viollon2002, ricciardi2014, pheasant2008}). Though, these interactions go beyond the range of this study in the context of acoustic sensor networks where only sound is recorded.

\end{enumerate}

\section{Answers to Reviewer III}

\begin{enumerate}

\item \emph{I am not sure what to make of this study. I simply do not get the message, if there is one. The article seems to lack a clear scientific question, and consequently there is no answer - no message. [...]\\
The arguments the authors present in the introduction for conducting this study do not convince me. Surely, future smart cities could have all sorts of sensors all over for monitoring various types of physical entities, including sound. But, would that motivate the present study and its approach? [...]\\
In general, I often find it hard to follow the presentation of the study and its results, and often get confused. Perhaps this is because there is no red thread in the story or no clear scientific question to answer. [...]\\
Overall, for this article to be successful, the authors must agree on one clear scientific question and one clear message. Then they must write a clear story, guiding the reader to the conclusions. Remember to answer your question.}

$\rightarrow$ The aim of this work is to determine wether a deep learning architecture can identify active sound sources in polyphonic urban sound environments from typical monitoring measurements (third-octave spectra). It is motivated by the fact that the composition of sound environments has a known influence on the perception of pedestrians, but is currently not explicitely studied in monitoring applications based on sensor networks. We achieve this goal by learning on labels approximating the perceived time of presence, which can be obtained in an automatic way if the ground truth source contributions in the training set is known. Pleasantess prediction is studied here as an application to assess the usefulness of the proposed approach.\\
The structure of the manuscript has been revised to clearly expose the process of this work:
\begin{itemize}
\item Section 2 details the deep learning dataset construction, its validation through a listening test and the choice of acoustic indicator for its automatic annotation,
\item Section 3 presents the training and results of the source detection deep learning model,
\item Section 4 applies this model to pleasantness estimation.
\end{itemize}

\item \emph{At Inter-Noise in Hamburg in 2016 Lund\'en, Axelsson and Hurtig presented the results from a similar study, which seems to have been driven by the curiosity of learning whether machine learning is useful at all and what acoustical features would predict the outcome of soundscape assessments. This does not appear to be the motivation for the present study. Yet, the authors place a lot of energy on reporting on which regression model provides for the best predictions. Why is that?\\
Scientifically, the most interesting question related to this study is: what acoustical properties would be most relevant to monitor in the future if we are to replace the current sound level approach with one that is more holistic? It seems that the authors have some insight into this. I recommend refocusing the article on this question. It would make the article a lot more worthwhile. Leave the potential practical applications of the results for the discussion, rather than trying to use them as a motivation for the study in the introduction.}

$\rightarrow$ The original structure of the manuscript was misleadingly focused on pleasantness estimation, though it is in fact only an application of the proposed approach studied here to evaluate the quality of source presence predictions. As such, the motivation of this study is entirely on the identification of active components of the soundscape, which could have several fields of application from quality assessment to information-rich mapping.

\item \emph{On Rows 39-40 in the introduction the authors claim that noise maps stipulated by the European Environmental Nose Directive (2002/49/EC; please correct the typo on Row 37) would provide "useful tools for urban planning and noise reduction purposes." I must say that urban planning is not at all on the EC agenda in this respect. The purpose of the noise maps is to provide intelligence for noise actions plans only. Urban planning would require a completely different approach than what the present noise abatement framework provides.}

$\rightarrow$ The claim that the European Comission 2002/49/EC directive was aimed at urban planning has been removed from the introduction.

\item \emph{The quote related to ISO 12913-1 on Rows 43-45 is incorrect. Please correct.}

$\rightarrow$ Following the changes in the introduction structure, the quote related to ISO 12913-1 no longer appears.

\item \emph{In particular, it is hard to understand the presentation of figures, tables and equations. They must be more fully explained in the text.}

$\rightarrow$ Figures, tables and equations are now more clearly explained in the text.

\item \emph{I recommend the authors to occasionally remind the reader about the meaning of the different abbreviations used. Now I must move back and forth in the text to try to remember what they mean, which causes me to lose track of where I am going. This makes it very hard to interpret and to understand the results. It is necessary to support the reader in understanding the text.}

$\rightarrow$ Several abbreviations have been removed or explained throughout the paper to improve its overall readability.

\item \emph{Also keep in mind that the reader may not be an engineer, physicist or mathematician and may not understand the equations. Please explain the equations in plain language, and help the reader to understand what they mean. For example, what does the second derivative of a function mean (velocity or acceleration)?}

$\rightarrow$ Explications for the main equations in the manuscript have been added (eq. 1-3). Specifically, in eq. 1 the derivation is applied once on each dimension of the third-octave spectrum (hence the term second derivative), which underlines variations in time and frequency.

\item \emph{I fail to understand how the different parts of the study relate to each other. How does the perception study relate to the machine learning part? Are they connected, or are they completely separate? I also got confused when I learned that there were some 400, or was it 200, simulated acoustic scenes used for machine learning, while there first were some 75 such scenes.}

$\rightarrow$ The structure of the paper has been changed to better link different parts of the study. Section 2 presents the generation of a large simulated dataset with 400 sound scenes for training a source detection model and an additional (independant) 200 scenes for its evaluation. A third corpus containing 100 sound scenes including 75 simulated, 19 replicated and 6 recordrd, is constructed in Section 2.2 as part of a listening test to assess the change in perception in simulated scenes and to obtain subjective evaluations useful in the rest of the paper. In Section 2.3, the listening test results motivate the choice of indicator used to automatically annotate the perceived time of presence on the 600 total scenes used in the machine learning approach.

\item \emph{The Equations 5-11 must be explained in more detail. Please also inform the reader that you created multivariate linear regression models and how this was done. It was not until I saw the equations that I understood that they are regression models. This is not explained in the text. In general, you must present to the reader how statistical analyses are conducted and with what statistical package. At present, there is no such information in the text.}

$\rightarrow$ Additional explications were added to equations 5-11 (now 5-8), and the procedure used to obtain baseline models from perceptual and acoustical parameters through multivariate linear regressions has been detailed in Section 4.1. All statistical analyses are conducted using Matlab and its statistical toolbox (code is available) and the machine learning part is implemented in the Pytorch framework. This is now stated where applicable.

\end{enumerate}

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2019-10-02 paper/paperV2.tex
CL/PA proof + Cover

@@
 
-The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution).
+The impact of urban sound on human beings has often been studied from a negative point of view (noise pollution).
 In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces).
 The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments.
-There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.
+There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence, dominance or volume.
 
@@
 Perceptual attributes corresponding to those stimuli are gathered through a listening experiment.
-The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure.
-The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.
+From the contributions of the individual sound sources available for the simulated corpus, a physical indicator approximating the perceived time of presence of sources is computed and used to train and evaluate a multi-label source detection model.
+This model predicts the presence of simultaneously active sources from fast third octave spectra, allowing the estimation of perceptual attributes such as pleasantness in urban sound environments at a sufficient degree of precision.
 
@@
 
-%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor.
-Consequently, the 2002/49/EC European directive~\cite{ec2002} requires the development of noise maps in large cities.
-These maps aim at providing useful tools for noise reduction planning purposes.
-In this context, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as the "acoustic environment as perceived or experienced and/or understood by people, in context" provides a shift from the noise-only paradigm by considering the possible positive influence of some the sound environment's components.
-
 Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}.
 This requires the identification of its composition in terms of sources of interest from acoustic measurements at a reasonable cost.
-Most monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment.
-Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}.
-While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline.
-Indeed, it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved.
+Most current monitoring applications rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013, nilsson2006}, which offers limited information regarding the content of the sound environment.
+Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio such as Mel spectrograms or Mel frequency cepstrum coefficients, on much finer time scales in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}.
+Using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators.
+Though, it may not be desirable in long-term monitoring applications where analysis is performed offlinen as it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved~\cite{gontier2017}.
 Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information.
@@
 A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}.
-Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}.
+In the case of pedestrians, three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}.
 Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}.
-In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}.
-The perceived time of presence has the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.
+In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010}, the time of presence~\cite{ricciardi2014}, or the volume.
+The dominance is built on coupled notions of time and volume, while the time of presence is more easily linked to the physical activity of sources.
+In a more holistic approach of soundscape quality evaluation, the time of presence should thus be easier to identify from acoustic monitoring measurements and would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.
 
 In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which has demonstrated state-of-the-art performance in many tasks studied in the machine listening community~\cite{mesaros2017}.
-In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}.
-Several architectures may be used to this aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}.
-Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}.
+In the context of urban sound environments, machine listening has been successfully applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017, boes2018}, sound scene classification~\cite{torija2013, barchiesi2015, sun2019, valenti2016, salamon2017-2} and soundscape quality evaluation~\cite{lunden2016, yu2009, boes2018}.
+A wide range of architectures exist, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}.
 Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}.
@@
   \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
-  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy through a comparative study with state of the art approaches.
+  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes (pleasantness) to a sufficient degree of accuracy through a comparative study with state of the art approaches.
 \end{enumerate}
 
-The procedure for simulating sound scenes corpora is presented in Section~\ref{sec:data} with the outcomes of the listening test conducted to justify the relevance of the sound scenes corpora in the context of a study on perceived quality.
-A large corpus is constructed and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales, as presented in Section~\ref{sec:pred}.
-The application of the resulting time of presence to the prediction of pleasantness in urban environments is then presented in Section~\ref{sec:app}.%The, in comparison to existing models from perceptual and physical indicators.
+Section~\ref{sec:data} presents the generation procedure for a large corpus of simulated sound scenes, its validation through a listening test in the context of a study on perceived soundscape quality, and a study of available acoustical indicators for the automatic annotation of source activity.
+In Section~\ref{sec:pred} a deep learning architecture that performs multi-label sound source recognition at relevant time scales is trained and evaluated.
+Section~\ref{sec:app} then presents an application of this model to the prediction of pleasantness in urban environments.
 
@@
 This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
-Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality~\cite{lavandier2006}.
-The corresponding database of isolated is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
+Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}.
+The corresponding isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
 \textit{simScene} generates original scenarios using the following information:
@@
 \item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
-\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
+\item Event-to-background ratios in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
 \item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
@@
 
-Newly generated scenarios should cover most real-life situations while remaining perceptually plausible.
-To do so, a corpus of 74 recordings obtained in~\cite{aumond2017} is taken as reference.
-The recordings include 19 locations in the 13th district of Paris, France.
-A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}.
-The statistics on background and event properties needed for \textit{simScene} generation are extracted conditionally on the ambiance from annotations of the corpus presented in~\cite{gloaguen2017}, which include:
+Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible.
+To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}.
+These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration.
+In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}).
+The available annotations include:
 
@@
 \item Background sources that are present throughout the whole scene, and their respective sound level considered constant,
-\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
+\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio in dB.
 \end{itemize}
 
-Adjustments are made on the variance of the considered properties to better cover possible scenarios.
-Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism.
-An additional \textit{square} ambiance is added with properties derived empirically from available data.
+This information is reused to extract statistics on background and event sources activity, which are then used as inputs by \textit{simScene}.
+Though, adjustments are made on the variance of the considered properties to extend the range of covered scenarios.
+Furthermore, since voice events in the isolated samples database consist mostly of read English recordings, the voice event mean event-to-background ratios are reduced for all ambiances to improve realism.
+An additional \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity.
 Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes.
@@
 The generation procedure is the same for both datasets.
-Though, it is important in deep learning applications to ensure a complete independance of the training and test set.
-To do so, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.
+Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets.
+To do so, the isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
+As a result, the two corpora differ by the scenarios and isolated samples used in their generation.
 
@@
 
-An experiment in the form a listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.
+An listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model.
+Given the aim of this test, a separate corpus is constructed.
+This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.
 
-Given the aim of this test, a separate corpus is constructed.
-First, 200 simulated scenes of 45s each with diverse new scenarios are generated with the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences.
+First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a third isolated samples database to ensure independance from the deep learning datasets.
 These scenes are equally distributed among the five considered ambiances (resp.
@@
 To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected.
-The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}.
-Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.
+Figure~\ref{fig:tvb_pres} shows the distribution of the 75 selected extracts.
+Playback sound levels are realistically drawn for each scene as described in Section~\ref{sec:data_corp}, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.
 
@@
     \end{subfigure}
-    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
+    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with scenarios generated by \textit{simScene}.}\label{fig:tvb_pres}
 \end{figure*}
@@
 The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
-The annotations obtained in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}.
-Furthermore, the original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
-The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.
+The manual annotations available in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}.
+Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
+The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\
 
-The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.\\
-
-During the test, participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10).
+During the test, participants are asked to evaluate sound scenes in terms of 8 criteria represented by 11 point scales (0-10).
 These scales are presented in French and translated in this report using standard terminology.
@@
 \item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
-\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
+\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
 \end{itemize}
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-The perceived time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study.
-The influence of traffic on perception is further separated into that of background and salient events.
-Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
+Additionally, to assess the perceived source activity 3 additional questions are presented to the participants and evaluated on the same 11 point scales:
 \begin{itemize}
-\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
-\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
+\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
 \end{itemize}
@@
 
-A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test (note that with 23 subjects, the incomplete block design is not perfectly balanced).
-All gave consent prior to the experiment and reported normal hearing.
+A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 subjects, the incomplete block design is not perfectly balanced).
+All participants gave written consent prior to the experiment, and evaluations were further anonymized.
 
@@
 
-The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied.
+The effect of the \textit{simScene} generation procedure on perception is first investigated.
+To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scenarios.
 Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios.
@@
 
-Next, the perceptual space generated by the experiment's five general scales (P, L, OL, I and C) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity.
-It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants.
+Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity.
+It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants.
 No standardization is applied to the data.
@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes.
+Additionally, the assessments averaged on all subjects are projected onto the principal components space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes.
 These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments.
@@
 
-Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters.
-These indicators may be used to automatically annotate sound scenes in the design of a deep learning architecture for sound source recognition.
+\begin{figure*}[t!]
+    \centering
+    \includegraphics[width=1.5\columnwidth]{figures/pred_ex.png}
+    \caption{Example of simulated sound scene with ground truth source contributions and resulting presence predictions by the deep learning architecture (greyed out areas)}\label{fig:pred_ex}
+\end{figure*}
 
-First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level.
-Typically this includes indicators derived from sound level measurements.
-For this study the following variables are computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
+To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:data_corp}, the activity of sound sources present in each scene must be known.
+Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source.
+Several indicators are identified in the literature that that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}.
+These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.
 
+First, some indicators can be computed directly from the mixed audio.
+Typically this includes indicators derived from sound level measurements used in monitoring applications.
+For this study the following variables are considered, and computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:
 \begin{itemize}
@@
 \end{itemize}
-
-The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:
-
+The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of perceived source activity. Its expression is:
 \begin{equation}
@@
 For example, bird activity is characterized by narrow-band energy with fast paced variations in time, which translates into high $TFSD$ values in the corresponding frequency range.
-As a baseline indicator for source activity, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.
+The $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the perceived activity of birds and voices respectively.
 
-The simulation process of the scenes outputs ground truth source contributions as separate channels.
-Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources: The equivalent sound level $L_{eq, s}$ for source $s$ and the source emergence $\Delta L_{s}$, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
-
-Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is computed.
+In the case of simulated scenes, the generation process outputs ground truth source contributions as separate channels.
+This information could be used to compute additional indicators that describe source activity perception with better accuracy.
+The following are computed for traffic, voice and bird sources: the equivalent sound level $L_{eq, s}$ for source $s$ and the source emergence $\Delta L_{s}$, taken as the difference between the equivalent sound level of source $s$ and that of all other sources combined.
+Events and background occurences of the same source are added in this study.
+Next, the $\hat T_s(\alpha, \beta)$ time of presence approximation proposed in~\cite{gontier2018} is considered.
 $\hat T_s(\alpha, \beta)$ is based on a binary source emergence model computed on the third-octave band emergence spectrum $\Delta L_{s}(t, f)$.
@@
 A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene.
-The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.
+The optimal threshold values are optimized via grid search so that the time of presence estimation $\hat T_s(\alpha_{opt}, \beta_{opt})$ is best correlated with the average perceived time of presence $T_{s, p}$ as obtained by human evaluation.
+Both thresholds are optimized once in this study using data from the listening test in Section~\ref{sec:data_exp} and equal $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$, though several other pairs of values yielded close performance as per the optimization metric defined in eq.\ref{eq:ts}.
 
@@
 
-Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments.
-First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
-Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71).
-Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources.
-The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceived counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim.
-They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}.
-They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus.
-Thus, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicator can be used to annotate simulated sound scenes in terms of estimated perceived time of presence for the three sources.
-
-\begin{table*}
+\begin{table*}[ht!]
 \centering
@@
 \end{table*}
+
+Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments.
+First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
+Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71).
+Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources.
+The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceived counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim.
+They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were already correlated with those of other sources in Table~\ref{tab:percc}.
+They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus.
+Thus, because source contributions are available in simulated scenes they can be automatically annotated in terms of estimated perceived time of presence using the $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicator for the three sources.
 
@@
 Traffic is usually located in lower frequency bands than voice or bird events, but may contain high frequency components mistaken for birds by the model.
-The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.
+The resulting root mean squared error on the time of presence is 12\% overall, and is close for the three sources.
 
@@
 
+Figure~\ref{fig:pred_ex} presents an example of predictions for a sound scene in the listening test corpus.
+The model takes the mixed audio (top) as its input, and outputs binary presence predictions for traffic, voice and bird sources on 1s time frames.
+Here the ground truth contributions are shown as reference, though they are never available to the deep learning architecture.
+A traffic event masks voice and bird activity from 23s to 37s, and an obvious false positive for traffic presence is visible at 7-8s.
+
 \section{Application to pleasantness prediction}
@@
 
-In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}.
+In the context of urban soundscape quality assessment for pedestrians, pleasantness emerges as the first affective dimension~\cite{axelsson2010}.
 Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
@@
 
-\subsection{Models of pleasantness}
+\subsection{Baseline models of pleasantness}
 \label{sec:app_mdls}
 
-\begin{table*}[t]
+\begin{table}[t]
 \centering
@@
 \hline
-	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
-	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
-	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
-	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
-	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
-	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
-	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
-	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
-	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
-	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
+	 & P & OL & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
+	P & 1 & -0.89** & -0.76** & 0.05 & 0.57** \\
+	OL &  & 1 & 0.59** & 0.17 & -0.45** \\
+	$T_{T, p}$ &  &  & 1 & -0.35** & -0.42** \\
+	$T_{V, p}$ &  &  &  & 1 & -0.21* \\
+	$T_{B, p}$ &  &  &  &  & 1 \\ \hline
 \end{tabular}
 %}
-\end{table*}
+\end{table}
 
@@
 Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
-The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both for the relations to source contributions and between general scales, as discussed in Section~\ref{sec:data_val}.
-Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
+The correlations between perceived source activity and pleasantness are consistent with the literature~\cite{aumond2017, gontier2018}.
+Pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds.
 In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered in this study.
 This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments.
-Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies.
-However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.
-Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42).
-The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.
+Relations between source-specific parameters are weak with the exception of traffic being negatively correlated with birds (r=-0.42).
 
-Multilinear regression models of pleasantness are built as a function of the overall loudness (OL) and source-specific parameters ($L_{T, p}$, $T_{T, p}$, $T_{V, p}$, $T_{B, p}$) with assessments averaged for each scene (n=100).
-All 31 combinations of the 5 predictors are considered, but to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
+Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged for each scene (n=100).
+All 15 combinations of the 4 predictors are considered, and to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
 Only combinations for which all predictors verify $VIF<5$ are considered valid.
-The best resulting models in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) are:
+The best resulting model in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) is:
 
-\begin{align}
+\begin{equation}
 \begin{split}
-\hat P_{1, p} & = 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p}\\
-&\qquad + 0.14~T_{B, p} \label{eq:pp1}
-\end{split}\\
-\begin{split}
-\hat P_{2, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
-&\qquad + 0.12~T_{B, p} \label{eq:pp2}
+\hat P_{1, p} & = 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p}\\
+&\qquad + 0.12~T_{B, p} \label{eq:pp1}
 \end{split}
-\end{align}
+\end{equation}
 
-Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity.
-A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.
+where $OL$, $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$ represent the perceived overall loudness and time of presence of traffic, voices and birds respectively.
+This expression is close to existing models of pleasantness in the literature~\cite{ricciardi2014, aumond2017}, with both the overall loudness and traffic activity negatively contributing and bird activity positively contributing to pleasantness.
+A small positive contribution of the time of presence of voices is also found on this corpus despite no significant correlation underlined in Table~\ref{tab:percc}.
 
@@
 Namely the $L_{50}$ and $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicators are used as predictors, as they display the highest correlation values with the overall loudness and time of presence of sources respectively.
-Again, all combinations of the proposed physical variables are considered, and a VIF check ($VIF<5$) is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
+Again, all combinations of the proposed physical variables are considered, and a variance inflation factor check ($VIF<5$) is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
 On the present corpus, the best model in terms of $R^2_{adj}$ is:
@@
 where $\varphi$ indicates a model from physical variables.
-Indicators related to traffic and voices activity are absent compared to the perceptual models.
-This is consistent with the results underlined in Table~\ref{tab:physc}: the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
-These results can be attributed to the diversity of the studied corpus.
-Indeed, only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
+Indicators related to traffic and voices activity are absent compared to the perceptual model.
+This is consistent with the results underlined in Table~\ref{tab:physc}: the $L_{50}$ used as a predictor is correlated to the traffic time of presence $T_{T, p}$ that appeared in the perceptual model $\hat P_{1, p}$ (eq.\ref{eq:pp1}), and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
+These results can be attributed to the diversity of the studied corpus as only traffic, voice and bird sources are present in simulated sound scenes.
+As a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process.
 Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic.
@@
 
-The $\hat P_{1, \varphi}$ model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal.
-Both models' coefficients are re-optimized on the studied data for a fair comparison:
+The $\hat P_{1, \varphi}$ model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, noted $\hat P_{2, \varphi}$ and $\hat P_{3, \varphi}$ respectively, for which predictor variables are directly computed from the audio signal.
+Coefficients for both models are re-optimized on the studied data for a fair comparison.
 
@@
 
-Table~\ref{tab:pleasm} summarizes the performance of the two perceptual and three physical models.
-$\hat P_{1, p}$ and $\hat P_{2, p}$ display similar estimation errors, adjusted $R^2$ and correlation to assessed pleasantness.
-The performance of obtained perceptual models is about 0.6 (RMSE), which is below the average standard deviation of pleasantness assessments for this experiment: 1.77 on a 11-point scale.
+The $\hat P_{2, \varphi}$ model does not explicitely involve the contribution to pleasantness of specific sources, but rather of emerging sound events with the $L_{10}-L_{90}$ indicator.
+Conversely, the $\hat P_{3, \varphi}$ model includes the time and frequency second derivative in the $500Hz$ and $4kHz$ bands at relevant time scales to underline the activity of voice and bird sources.
+
+Table~\ref{tab:pleasm} summarizes the performance of the perceptual and physical models.
+The perceptual model $\hat P_{1, p}$ yields a root mean squared error of 0.61, which is below the average standard deviation of pleasantness assessments for this experiment: 1.77 on a 11-point scale.
 $\hat P_{1, \varphi}$ outperforms both $\hat P_{2, \varphi}$ and $\hat P_{3, \varphi}$, though a validation on a different corpus would be necessary to conclude on its capabilities.
-Compared to the models from perceptual parameters its $R^2_{adj}$ is about 9\% lower.
-Its RMSE is 0.83, which is also high compared to perceptual baselines, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 in this study.
-Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were obtained using only the $L_{50}$.
+Compared to the model from perceptual parameters its $R^2_{adj}$ is about 9\% lower.
+Its root mean squared error is 0.83, which is also high compared to the perceptual baseline, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 in this study.
+Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{2, \varphi}$ has no impact for the considered corpus, as the same overall performance metrics were observed using only the $L_{50}$.
 
@@
 	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
-	$\hat P_{1, p}$ & 0.59 & 0.91 & 0.95** \\
-	$\hat P_{2, p}$ & 0.61 & 0.90 & 0.95** \\ \hline
+	$\hat P_{1, p}$ & 0.61 & 0.90 & 0.95** \\ \hline
 	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
@@
 To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
-
-Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}.
-Thus, only the outputs corresponding to the presence of birds are used.
-Table~\ref{tab:pppred} presents the performance of pleasantness estimations using outputs from the deep learning architecture compared to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels computed with eq.~\ref{eq:ts} on separated source-specific channels.
-First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
-This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}.
-Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training.
-This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation.
-For the simulated subcorpus of the experiments all performance metrics are comparable.
-Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
-This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.
 
@@
 
-Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:app_mdls}.
+Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}.
+Thus, only the outputs corresponding to the presence of birds are used.
+Table~\ref{tab:pppred} presents the performance of pleasantness estimations using outputs from the deep learning architecture compared to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels computed with eq.~\ref{eq:ts} on separated source-specific channels.
+First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
+This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}.
+Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training.
+This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation.
+For the simulated subcorpus of the experiments all performance metrics are comparable.
+Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$.
+This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.
+
+Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from acoustical indicators in table~\ref{tab:pleasm}.
 This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average.
@@
 The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available.
-Indeed, the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy.
+This is because the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy.
 Of course, the complexity of the learning process scales with the number of considered simultaneously active sources.
@@
 
-This research is funded by the French National Agency for Research, under the CENSE project (convention ANR-16-CE22-0012)
+This research is funded by the French National Agency for Research, under the CENSE project (convention ANR-16-CE22-0012).
+Part of the data used was collected in the framework of the GRAFIC project, supported by the French Environment and Energy Management Agency (ADEME) under contract No.
+1317C0028.
 

### 2019-10-07 paper/paperV2.tex
JFP Proof

@@
 These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration.
-In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}).
+In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}).
 The available annotations include:
@@
 
-An listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model.
+A listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model.
 Given the aim of this test, a separate corpus is constructed.
@@
 The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
-The manual annotations available in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}.
+The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene}.
 Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
@@
 
-During the test, participants are asked to evaluate sound scenes in terms of 8 criteria represented by 11 point scales (0-10).
+During the test, participants are asked to evaluate sound scenes using 8 criteria represented by 11 point scales (0-10).
 These scales are presented in French and translated in this report using standard terminology.
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-Additionally, to assess the perceived source activity 3 additional questions are presented to the participants and evaluated on the same 11 point scales:
+Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11 point scales:
 \begin{itemize}
@@
 Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, the assessments averaged on all subjects are projected onto the principal components space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes.
+Additionally, the assessments averaged on all subjects for active individuals (simulated scenes) are projected onto the principal components space for and represented in Figure~\ref{fig:pspace_sim} as dots.
+The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space and represented as crosses.
 These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments.
-This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.
+This further demonstrates the diversity of scenarios created by the scene generation procedure.
 
@@
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
-Assessments of simulated scenes are projected as dots, and recorded and replicated scenes are projected as crosses.}\label{fig:pspace_sim}
+Assessments of simulated scenes (active individuals) are projected as dots, and recorded and replicated scenes (supplementary individuals) are projected as crosses.}\label{fig:pspace_sim}
 \end{figure}
@@
 Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source.
-Several indicators are identified in the literature that that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}.
+Several indicators are identified in the literature that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}.
 These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.
@@
 Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged for each scene (n=100).
-All 15 combinations of the 4 predictors are considered, and to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression.
+All possible combinations of variables among the 4 predictors (15 combinations) are considered, and to ensure that no multi-collinearity exists between predictors, a variance inflation factor (VIF) test is performed prior to each regression.
 Only combinations for which all predictors verify $VIF<5$ are considered valid.
@@
 
-The proposed deep learning method is applied to the perceptual experiment corpus to obtain estimations for the time of presence of traffic, voices and birds.
-These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}.
-To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
-
 \begin{table*}[t]
@@
 \end{table*}
+
+The proposed deep learning method is applied to the perceptual experiment corpus to obtain estimations for the time of presence of traffic, voices and birds.
+These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}.
+To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
 
@@
 The perceptual models on the listening test corpus are similar to those found in previous studies with both \textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}.
-Second, the quality of predictions is bound by that of annotations.
+Second, the quality of predictions is bounded by that of annotations.
 Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test.
@@
 
-The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available.
+The methodology proposed in this paper can be extended to include the identification of additional sound sources, though it requires the availablity of isolated samples to simulate enough diverse sound scenes to train the detection architecture.
 This is because the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy.

### 2019-10-07 paper/submitted/coverLetter.tex
v2 ready to submit

--- 2019-10-02 paper/coverLetter.tex
@@
 
-As a preamble, we would like to thank the editor and the reviewer for their comments and suggestions.
+As a preamble, we would like to thank the editor and the reviewers for their comments and suggestions.
 Following these comments, we made several changes to the article including a modification of the structure of the document, which are summarized here.
@@
 Section 2.1 has been rewritten to describe more extensively this reference database.
-To generate more diverse scenarios the variance of all statistics is further increased.
+To generate more diverse scenarios the variance of all statistics is next further increased.
 However, we acknowledge that the model needs to be evaluated on a more complete and representative set of recordings to assess fully its generalization capabilities.
@@
 
-$\rightarrow$ In this paper, we chose to focus on the three main categories of sources that are known to impact soundscape quality.
+$\rightarrow$ In this paper, we chose to focus on the three main canonical categories of sources that are known to impact soundscape quality.
 When applying the proposed approach to a specific environment, the taxonomy of sound sources needs to be extended accordingly.
 This is now discussed in Section 5.
-Also, as is now emphasized in the Introduction and Section 2 the present study considers the point of view of pedestrians, which is vastly different from that of residents in terms of sources of interest.\\
+Also, as now emphasized in the Introduction and Section 2, the present study considers the point of view of pedestrians, which is vastly different from that of residents in terms of sources of interest.\\
 The quality of the pleasantness model obtained from time of presence approximations is bound by the listening test corpus, and thus mostly by the contents of simulated scenes.
 Section 4.1 discusses possible causes to the absence of other sources in the model.
-However, this model is not intended as a contribution but rather as a intermediary to show that presence predictions from the deep learning architecture are accurate enough to estimate more abstract perceptual notions.
+However, this model is not intended as a contribution but rather as a mean to show that presence predictions from the deep learning architecture are accurate enough to estimate useful and more abstract perceptual indicators.
 
@@
 $\rightarrow$ The main objective of this work is the prediction of the perceived time of presence of sources of interest for the perceptual characterization of soundscapes.
-We study the estimation of pleasantness as an application, but the developped method could also be used to predict other quality descriptors or in mapping applications (eg.~\cite{lavandier2016}).
+We study the estimation of pleasantness as an application, but the developed method could also be used to predict other quality descriptors or in mapping applications (eg.~\cite{lavandier2016}).
 The structure of the paper has been rearranged to emphasize this aim.
@@
 Particularly, each scene contains at least one background source for which recordings obtained on Freesound are used.
-These recordings were selected to be representative of the source but were not cleaned through pre-processing, leading to (uncontrolled) background noise being present to an extent in each simulated scene.
+These recordings were selected to be representative of the source but were not cleaned through pre-processing, leading to (uncontrolled) background noise being present to some extent in each simulated scene.
 
@@
 
-$\rightarrow$ This work is placed in the context of a monitoring application through sensor networks in urban areas, for which using features computed on fine time scales may not be viable in the long term due to storage and privacy concerns.
-Though information on smaller time scales would be more appropriate to detect sound events (from a few milliseconds for birds to tens of milliseconds for speech), this motivates our choice to operate on the third-octave spectrum with 125ms hop size, which is typical in monitoring studies.
+$\rightarrow$ Information on smaller time scales would indeed be more appropriate to detect sound events (from a few milliseconds for birds to tens of milliseconds for speech).
+Though, this work is placed in the context of a monitoring application through sensor networks in urban areas, for which using features computed on fine time scales may not be viable in the long term due to storage and privacy concerns.
+This motivates our choice to operate on the third-octave spectrum with 125ms hop size, which is typical in monitoring studies.
 This is now discussed in the introduction.
@@
 
-$\rightarrow$ The abstract has been revised to reduce background information in favor of the description and results of the current study.
+$\rightarrow$ The abstract has been revised to reduce background information in favor of the description of the contributions and results of the current study.
 
@@
 $\rightarrow$ Many studies dealing with the environmental acoustic quality are associated with duration of acoustic measurements ranging from few seconds for the shorter~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}.
-Due to the limited duration of perceptual tests because of the fatigue, stimuli between 30s and 1mn are often prefered for laboratory tests.
+Due to the limited duration of perceptual tests because of the fatigue, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests.
 
@@
 $\rightarrow$ The aim of this work is to determine wether a deep learning architecture can identify active sound sources in polyphonic urban sound environments from typical monitoring measurements (third-octave spectra).
-It is motivated by the fact that the composition of sound environments has a known influence on the perception of pedestrians, but is currently not explicitely studied in monitoring applications based on sensor networks.
+It is motivated by the fact that the composition of sound environments has a known influence on the perception of pedestrians, but is currently not explicitly studied in monitoring applications based on sensor networks.
 We achieve this goal by learning on labels approximating the perceived time of presence, which can be obtained in an automatic way if the ground truth source contributions in the training set is known.
-Pleasantess prediction is studied here as an application to assess the usefulness of the proposed approach.\\
-The structure of the manuscript has been revised to clearly expose the process of this work:
+Pleasantness prediction is studied here as an application to assess the usefulness of the proposed approach.\\
+The structure of the manuscript has been revised to clearly expose the contributions of this work:
 \begin{itemize}
@@
 
-$\rightarrow$ The claim that the European Comission 2002/49/EC directive was aimed at urban planning has been removed from the introduction.
+$\rightarrow$ The claim that the European Commission 2002/49/EC directive was aimed at urban planning has been removed from the introduction.
 
@@
 
-$\rightarrow$ Explications for the main equations in the manuscript have been added (eq.
+$\rightarrow$ Explanations for the main equations in the manuscript have been added (eq.
 1-3).
@@
 $\rightarrow$ The structure of the paper has been changed to better link different parts of the study.
-Section 2 presents the generation of a large simulated dataset with 400 sound scenes for training a source detection model and an additional (independant) 200 scenes for its evaluation.
-A third corpus containing 100 sound scenes including 75 simulated, 19 replicated and 6 recordrd, is constructed in Section 2.2 as part of a listening test to assess the change in perception in simulated scenes and to obtain subjective evaluations useful in the rest of the paper.
-In Section 2.3, the listening test results motivate the choice of indicator used to automatically annotate the perceived time of presence on the 600 total scenes used in the machine learning approach.
+Section 2 presents the generation of a large simulated dataset with 400 sound scenes for training a source detection model and an additional (independent) 200 scenes for its evaluation.
+A third corpus containing 100 sound scenes including 75 simulated, 19 replicated and 6 recorded, is constructed in Section 2.2 as part of a listening test to assess the change in perception in simulated scenes and to obtain subjective evaluations useful in the rest of the paper.
+In Section 2.3, the listening test results motivate the choice of the indicator used to automatically annotate the perceived time of presence on the 600 total scenes used in the machine learning approach.
 
@@
 $\rightarrow$ Additional explications were added to equations 5-11 (now 5-8), and the procedure used to obtain baseline models from perceptual and acoustical parameters through multivariate linear regressions has been detailed in Section 4.1.
-All statistical analyses are conducted using Matlab and its statistical toolbox (code is available) and the machine learning part is implemented in the Pytorch framework.
+All statistical analyses are conducted using Matlab and its statistical toolbox (code is made available on the companion website) and the machine learning part is implemented in the Pytorch framework (also available).
 This is now stated where applicable.

### 2019-11-26 paper/coverLetter2.tex
Second revision

\documentclass[10pt]{article}

\title{Reply to reviewers concerning submission AAA-D-19-00080R1: "Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques"}

\begin{document}

\maketitle

As a preamble, we would like to thank the editor and the reviewers for their comments and suggestions. Following these comments, we made several changes to the article, which are summarized here. The next section list our answers to each of the reviewer’s comments, with references to the revised manuscript where appropriate.

\section{Answers to Reviewer III}

\begin{enumerate}

\item \emph{The revised version of this manuscript reads somewhat better than the original version. However, it still would benefit from further restructuring and the provision of further details. In the response letter, I find a lot of detailed information in the authors responses to the reviewers that are not implemented in the manuscript. If one think that some information is valuable to the reviewers, it is equality relevant to the readers of the published work. In general, all information provided to reviewers in the response letter must be implemented in the manuscript. For many journals, failure in doing so is reason enough for rejecting a manuscript from being published. Please, revise the manuscript accordingly.}

$\rightarrow$ All elements provided to the reviewers as part of the previous response letter are now presented in the manuscript. In particular, the choice of 45~s long extracts for the listening test and absence of controlled background noise in simulated sound scenes are now included in Section 2.1.

\item \emph{As I understand this manuscript, it contains three parts. Consequently, I propose that each part of the study should be allocated its own separate section.\\
Section 2 should report on the listening experiment, only. I propose that it is structured into subsections: 2.1 Participants, 2.2 Stimuli, 2.3 Data collection instrument, 2.4 Equipment, 2.5 Procedure and experimental design, and 2.6 Results. Add or remove subsections as appropriate.}

$\rightarrow$ Section 2 has been rewritten according to the following structure: 2.1 Stimuli, 2.2 Equipment, 2.3 Participants, 2.4 Procedure, 2.5 Results.

\item \emph{Section 3 should report on the machine learning study, including the corpus used and how it was generated. I still do not understand how the latter was done. How many different sounds for each source (traffic, voices, birds) were used to generate the scenes?}

$\rightarrow$ The machine learning corpus is now described in Section 3.1 (instead of Section 2.1). As is now emphasized, its generation procedure is the same as for the simulated sub-corpus of the listening test. However, the isolated samples database is different. Details of the isolated samples database have been added in Section 2.1 and 3.1  for the listening test corpus and the deep learning corpus respectively.

\item \emph{For the reader to be able to evaluate the validity of the study, it is necessary to know some more about how simScene works and how it utilized the information from the 74 recordings from the GRAFIC project to generate the scenes used.}

$\rightarrow$ The description of the generation procedure from distributions extracted from the GRAFIC project recordings using simScene has been extended in Section 2.1.

\item \emph{Section 4 should report on the application of the machine learning model to the prediction of pleasantness. The purpose and objective of this section must be clearly and explicitly communicated to the reader at the beginning of the section, providing a clear focus. Currently, the purpose of the section is not clear enough.}

$\rightarrow$ The introduction of Section 4 has been extended to better motivate the objective of this application.

\item \emph{Probably the presentation of the acoustic indices (now in Section 2.4) should be moved to Section 4, where they are used.}

$\rightarrow$ Acoustic indices are first used to identify a potential automatic annotation of the deep learning corpus in terms of perceived source presence. The presentation of these indices has thus been moved to Section 3.2.

\item \emph{In describing what the authors did, they must be precise and consistent in the terminology. For each phenomenon described, they must use exactly the same terms throughout the manuscript without any variation, so that the reader may understand what they are referring to. At present, there is too much variation in expressions, which is confusing. For example, on Row 261 the authors mention 75 selected extracts. What are those? Are they the same as the 75 simulated scenes mentioned previously?}

$\rightarrow$ The terminology used throughout the manuscript has been revised and is now more consistent.

\item \emph{For the listening experiment, I understand it as if the authors used 100 sound scenes, out of which 75 were simulated, 6 were recordings and 19 were replicated scenes from a previous project. What does it mean that the 19 scenes from a previous project were replicated? Where they simulations that match previously recorded scenes, or were they new recordings that are similar to previous recordings?}

$\rightarrow$ As is now more clearly described in Section 2.1, replicated sound scenes are generated using simScene, with scene composition (scenarios) constrained by annotations of the reference recordings. This differs from the 75 simulated sound scenes for which the scene composition is sampled from gaussian distributions.

\item \emph{On Row 247, the authors mention that 200 scenes were generated. Does it mean that the 75 simulated scenes used were selected among the 200?}

$\rightarrow$ As the generation procedure for simulated scenes includes random sampling from distributions, simScene may output similar sound scenes. To avoid this phenomenon in the listening test corpus, 200 scenes are first generated, among which 75 scenes are selected to maximize content diversity.

\item \emph{On Rows 256-260: The authors write that "Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected."\\
I am not sure that I understand what 3D space the authors are referring to. Do they mean the 3D space of Traffic, Voices and Birds presented in Figure 1? If so, then please spell this out explicitly in the text, including what this 3D space represents and how it was created. Please also explain how to "maximize the minimum pairwise distance."\\
I find it very hard to understand Figure 1 from the limited information provided. It must be improved (or removed).}

$\rightarrow$ The selection procedure description has been simplified and Figure 1 has been removed to enhance the reader's comprehension, as the considered indicator for the selection of the 75 simulated scenes was not presented until Section 2.4 (now 3.2).

\item \emph{On Row 285, I wonder what it means that the authors "normalized" the sounds. Does it mean that the authors calibrated the sounds to the authentic sound levels?}

$\rightarrow$ The sound scenes are scaled so that they are heard at the desired sound level (as measured in the case of recorded and replicated scenes) when using the specific configuration (headphones, sound card) presented in Section 2.2. The term "normalized" has been replaced with "scaled" to avoid confusion.

\item \emph{On Rows 290-291, the authors write that the participants used 8 criteria to evaluate the sounds. I suppose that it means that they used 8 attributes. It also seems as if the authors used 11-point bipolar scales, delimited by opposite terms. Then the term 'Likert scale' used on Row 370 is incorrect. A Likert scale is a 5-point category scale asking to what extent a person agrees or disagrees with a statement. As far as I can tell, the authors did not use any Likert scales.}

$\rightarrow$ The term "Likert scale" has been removed throughout the manuscript.

\item \emph{Rows 329, 345, 352 and 414: Replace 'subjects,' with 'participants.'}

$\rightarrow$ All mentions of "subjects" are now replaced with "participants".

\item \emph{I do not understand the information provided in the paragraph about the sound level calibration on Rows 334-348. Particularly, I do not understand the rationale for the calibration procedure. It must be explained in some more detail. Perhaps a conceptual diagram showing how the different equipment were connected would be helpful? The reader should be able to replicate what was done based on the information provided in the manuscript.}

$\rightarrow$ The steps undertaken during the calibration procedure are now detailed in Section 2.2. From this procedure, a simple (linear) relation is obtained between the voltage at the output of the sound card and the sound level produced by the headphones. Sound scenes can be scaled using this information so that they are played at the desired sound level using the specific headphones and sound card configuration.

\item \emph{On Rows 401-403 it seems as if the 6 recorded and 19 replicated scenes (n=25) all were recordings. Is that so?}

$\rightarrow$ The 19 replicated scenes are generated by simScene, and thus are not recordings. This part has been rewritten to avoid any confusion in this regard.

\item \emph{Figures 2 and 3: Please observe that these two figures present Principal Components Loadings for the five attribute scales. Because the scales are bipolar, it is confusing to see lines extending in two directions. There should only be five vectors extending from the origin of the space to the points in the 2D space that represent the coordinates of the loadings on the two Principal Components. Moreover, I am not sure that it is necessary to include the ellipses, short arrows and datapoints in these figures. I just find it messy.}

$\rightarrow$ Both figures displaying the principal components analyses have been modified so that loadings are in one direction only. Furthermore, the ellipses and arrows have been removed along with the corresponding text. They are available as supplementary material on the companion website.

\item \emph{How were the PCA conducted and with what statistical package? [...] What statistical package was used to calculate the correlation coefficients? [...] What statistical package was used for conduction the regression analyses?}

$\rightarrow$ All statistical analyses are conducted using Matlab. This is emphasized in the introduction as well as in the beginning of Section 2.5. The code is also made available for further information about the specific functions used.

\item \emph{Equations 2 and 3 must be better explained, and presented separately. Please use plain language, to explain what the equations mean.}

$\rightarrow$ The description of both equations has been extended to enhance the reader's comprehension.

\item \emph{Table 2: I expect to see * p<0.05 and ** p<0.001 as foot notes below the table and not in the caption.}

$\rightarrow$ Significance thresholds have been moved to table footnotes where applicable.

\item \emph{Are they 1- or 2-tailed probabilities?}

$\rightarrow$ The Matlab \textit{corrcoef} function was used to compute all correlation coefficients. The corresponding significance test is two-tailed. This is now stated in Section 3.2 and 4.1, where correlation coefficients are presented.

\item \emph{The authors must also explain all abbreviations in the table. It must be possible to read tables and figures independently of the text body. Please, include all correlation coefficients in the table, also those that are not statistically significant.}

$\rightarrow$ Abbreviations of perceptual attributes have been added to the respective table captions. Non-significant correlations have also been added to Table 2.

\item \emph{In Section 4, I miss the relevant F- and t- statistics for the regression models. For each model, one must report the overall F-statistics and the t-statistics for each factor included in the models. This is typically presented in the text body.}

$\rightarrow$ F- and t- statistics have been added to the text following equations 5 and 6 on the baseline models of pleasantness.

\item \emph{For each model, the authors must also explain what data was used (which corpus).}

$\rightarrow$ Baseline models of pleasantness are computed on the listening test corpus, as the considered perceptual attributes are available on this corpus only. The baseline model from acoustic indices is computed on a subset of 92 sound scenes as the computation of time of presence estimations requires separated source contributions, which are not available in the 6 recorded scenes. These elements are now more clearly described in Section 4.1.

\end{enumerate}

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2019-11-26 paper/paperV3.tex
Second revision

--- 2019-10-07 paper/paperV2.tex
@@
 \usepackage{subcaption}
+%\usepackage{tablefootnote}
+\usepackage{threeparttable}
 \usepackage{bbm}
@@
   \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
-  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
+  \item Propose numerical means\footnote{All statistical analyses are done with Matlab.
+Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
   \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes (pleasantness) to a sufficient degree of accuracy through a comparative study with state of the art approaches.
@@
 
-Section~\ref{sec:data} presents the generation procedure for a large corpus of simulated sound scenes, its validation through a listening test in the context of a study on perceived soundscape quality, and a study of available acoustical indicators for the automatic annotation of source activity.
-In Section~\ref{sec:pred} a deep learning architecture that performs multi-label sound source recognition at relevant time scales is trained and evaluated.
+Section~\ref{sec:exp} presents the generation procedure for a corpus of simulated sound scenes and its validation through a listening test in the context of a study on perceived soundscape quality.
+In Section~\ref{sec:pred} a larger dataset is constructed and automatically annotated, and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales.
 Section~\ref{sec:app} then presents an application of this model to the prediction of pleasantness in urban environments.
 
-\section{Corpus}
-\label{sec:data}
+\section{Listening experiment}
+\label{sec:exp}
 
-\subsection{Generation procedure}
-\label{sec:data_corp}
+\subsection{Stimuli}
+\label{sec:exp_stim}
 
-Considering machine learning techniques to detect the presence of sources requires the availability of annotated data.
-Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed.
-This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
-Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}.
-The corresponding isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
-\textit{simScene} generates original scenarios using the following information:
+\begin{figure*}[t!]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/parcours.png}
+    \caption{Noise map of the path and 19 locations of the soundwalks presented in~\cite{aumond2017}.
+Sound levels (Leq in dB SPL) were measured at each of the 19 locations and interpolated to obtain a sound level estimations along the path.}\label{fig:path}
+\end{figure*}
+
+A listening test is conducted in order to validate the perceptual correspondence of simulated scenes with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model.
+To do so, a corpus is constructed that contains 100 sound scenes, including 75 simulated scenes, 6 recorded scenes for reference, and 19 replicated scenes that correspond to simulated reproductions of recorded scenarios.
+
+The 75 simulated and 19 replicated scenes are generated using the \textit{simScene} software, an open source library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}.
+\textit{simScene} can be used in two main modes.
+First, the \textit{generate} mode is used to create new scenarios.
+It is controlled by the following information for each source in the considered taxonomy:
 
@@
 
+\textit{simScene} samples these distributions to generate original scenarios with background and event annotations, then simulates sound scenes as the composition of examples in the isolated samples database, with source type, event-to-background ratios and event onsets given by the generated scenarios.
+
+The second mode is the \textit{replicate} mode, which directly takes as input a scenario with background and event source type, event-to-background ratio and onset-offset information.
+In this mode, \textit{simScene} only samples the available isolated samples database to simulate a sound scene that replicates the given scenario with different acoustic content.
+
+In this study we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}.
+The isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources.
+The isolated samples database contains 8mn, 17mn and 37mn of background traffic, voice and bird extracts respectively, as well as 30mn, 37mn, and 5mn of event traffic, voice and bird extracts respectively.
+No background noise is added to these three sources, though some uncontrolled noise is present in the uncleaned extracts composing the isolated samples database.
+A simulated sound scene is thus a mix of up to six separate channels, each one corresponding to either the background or event activity of one of the sources in the chosen taxonomy.
+
 Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible.
 To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}.
-These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration.
+These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~mn in duration.
 In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}).
@@
 
-This information is reused to extract statistics on background and event sources activity, which are then used as inputs by \textit{simScene}.
-Though, adjustments are made on the variance of the considered properties to extend the range of covered scenarios.
+This information is reused to estimate the mean and variance of the distributions on background and event sources activity used by \textit{simScene} to generate new scenarios.
+These distributions are conditional to the four considered ambiances and three source types.
+The variance of all event inter-onset distributions is manually increased to extend the range of covered scenarios.
 Furthermore, since voice events in the isolated samples database consist mostly of read English recordings, the voice event mean event-to-background ratios are reduced for all ambiances to improve realism.
-An additional \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity.
-Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes.
+A fifth \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity.
 At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.
 
-The simulated deep learning corpus is composed of two subsets:
-\begin{itemize}
-\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
-\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
-\end{itemize}
-
-The generation procedure is the same for both datasets.
-Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets.
-To do so, the isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set.
-As a result, the two corpora differ by the scenarios and isolated samples used in their generation.
-
-\subsection{Perceptual experiment}
-\label{sec:data_exp}
-
-A listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model.
-Given the aim of this test, a separate corpus is constructed.
-This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.
-
-First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a third isolated samples database to ensure independance from the deep learning datasets.
-These scenes are equally distributed among the five considered ambiances (resp.
+The 75 simulated scenes of the listening test corpus are generated using \textit{simScene} in the \textit{generate} mode with the extracted distributions as input.
+To improve the diversity of scenarios considered in the listening test, 200 sound scenes are first simulated, equally distributed among the five ambiances (resp.
 \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}).
-For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}.
-Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way.
-To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected.
-Figure~\ref{fig:tvb_pres} shows the distribution of the 75 selected extracts.
-Playback sound levels are realistically drawn for each scene as described in Section~\ref{sec:data_corp}, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.
-
-\begin{figure*}[th]
-    \centering
-     \begin{subfigure}[t]{0.33\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
-    \end{subfigure}%
-    \begin{subfigure}[t]{0.33\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
-    \end{subfigure}
-    \begin{subfigure}[t]{0.33\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
-    \end{subfigure}
-    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with scenarios generated by \textit{simScene}.}\label{fig:tvb_pres}
-\end{figure*}
+A common duration of 45 seconds is chosen due to the limited duration of listening tests because of the fatigue.
+Though many studies dealing with the environmental acoustic quality are associated with acoustic measurements ranging from a few seconds~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests.
+To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence for each type of sources (ranging from no presence to 100\% of presence) without redundancy.
+Playback sound levels are realistically drawn for each scene, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.
 
@@
 Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments.
+The path undertaken during these soundwalks is shown in Figure~\ref{fig:path}, with sound levels in dB SPL interpolated from measurements at the 19 locations.
 For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted.
 The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
-The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene}.
+The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode.
+The 19 resulting sound scenes are thus simulated, but their scenarios follow those of the reference recordings to the extent of annotation precision.
 Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
-The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\
+The 6 recorded and 19 replicated scenes are scaled using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.
 
-During the test, participants are asked to evaluate sound scenes using 8 criteria represented by 11 point scales (0-10).
-These scales are presented in French and translated in this report using standard terminology.
-The first 5 scales relate to general properties of the scene:
+\subsection{Equipment}
+\label{sec:exp_equi}
+
+During the test the scenes are played at a given sound level as discussed in Section~\ref{sec:exp_stim}, through the same computer and sound card configurations.
+Beyerdynamics DT-990 Pro headphones are used by all participants.
+The calibration of the headphones was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the corresponding binaural sound pressure.
+To do so, the following procedure was conducted:
+\begin{enumerate}
+\item A pink noise generator is set at an arbitrary level and its output voltage is measured.
+\item Small DPA 4060 microphones are set at the entrance of the ear canals of a human participant~\cite{moller1992}.
+The generator is used as input of the headphones, placed over the head of the participant.
+The voltage at the output of the binaural microphones is measured.
+\item The headphones are removed, and the generator is used as input of Genelec 1031A loudspeaker placed at a distance of 1 meter from the participant's head.
+The loudspeaker's amplification is tuned until the same voltage is outputted by the binaural microphones.
+\item The head is replaced by a sound level meter to measure the sound level in dB of the loudspeaker.
+This corresponds to the binaural sound level produced by the headphones for the considered output voltage of the pink noise generator.
+\end{enumerate}
+By repeating this procedure for different settings of the generator level, the relation between the logarithm of the generator output voltage and headphones playback sound level in dB is approximated as a linear function.
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener, when using the calibrated headphones and a specific sound card configuration for which the relation between the numeric level and output voltage is known.
+
+\subsection{Participants}
+\label{sec:exp_part}
+
+A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 participants, the incomplete block design is not perfectly balanced).
+All participants gave written consent prior to the experiment, and evaluations were further anonymized.
+
+\subsection{Procedure}
+\label{sec:exp_proc}
+
+\begin{table*}[t!]
+\centering
+\caption{Mean differences of perceptual assessments (resp.
+pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) between original and replicated sound scenes.
+Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
+\label{tab:ogrep}
+%\resizebox{\columnwidth}{!}{
+\begin{tabular}{ c | c c c c c c c c c }
+\hline
+	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
+	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & 0.39 & \textbf{-2.09} & 0.61 \\
+	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 1.04 & \textbf{-4} & 0.22 \\
+	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
+	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
+	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
+	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
+\end{tabular}
+%}
+\end{table*}
+
+Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point scales (0-10).
+These attributes are presented in French and translated in this report using standard terminology.
+The first 5 semantic differential scales relate to general properties of the scene:
 \begin{itemize}
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11 point scales:
+Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point scales:
 \begin{itemize}
@@
 P3 and P15).
-A random listening order is generated for each subject to control ordering effects for the remaining of the test.
+A random listening order is generated for each participant to control ordering effects for the remaining of the test.
 Participants can listen to each scene once, and have to listen to the full extract and to answer all questions before being allowed to proceed to the next scene.
 
-The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations.
-Beyerdynamics DT-990 Pro headphones are used by all participants.
-The calibration of the headphone was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the sound pressure at the entrance of a blocked ear canal.
-To do so, small DPA 4060 microphones have been set at the entrance of the ear canals of a human subject~\cite{moller1992}.
-From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.
+\subsection{Results}
+\label{sec:exp_res}
 
-A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 subjects, the incomplete block design is not perfectly balanced).
-All participants gave written consent prior to the experiment, and evaluations were further anonymized.
-
-\subsection{Simulated corpus validation}
-\label{sec:data_val}
-
+The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab.
 The effect of the \textit{simScene} generation procedure on perception is first investigated.
-To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scenarios.
+To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scene compositions.
 Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios.
@@
 As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}.
-On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale.
+On the first five scales, all mean differences are lower than 2 points on the 11-point discrete scale.
 Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters.
@@
 
-\begin{table*}[t]
-\centering
-\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
-\label{tab:ogrep}
-%\resizebox{\columnwidth}{!}{
-\begin{tabular}{ c | c c c c c c c c c }
-\hline
-	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
-	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
-	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
-	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
-	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
-	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
-	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
-\end{tabular}
-%}
-\end{table*}
-
 Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity.
@@
 No standardization is applied to the data.
-Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively.
+Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for recorded and replicated scenes (n=25) and simulated scenes (n=75) respectively.
 The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora.
@@
 Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}.
-Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities.
-Additionally, the assessments averaged on all subjects for active individuals (simulated scenes) are projected onto the principal components space for and represented in Figure~\ref{fig:pspace_sim} as dots.
-The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space and represented as crosses.
-These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments.
+Thus, the perceptual spaces based on simulated sounds or on recordings and replicated scenes, do not reveal major differences.
+Additionally, the assessments averaged on all participants for active individuals (simulated scenes) are projected as dots onto the principal components space in Figure~\ref{fig:pspace_sim}.
+The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space as crosses.
+These projections show that the space covered by simulated scenes based on new compositions covers that of the studied real-life environments (recorded and replicated sounds).
 This further demonstrates the diversity of scenarios created by the scene generation procedure.
+Additional information on the relation between pairs of recorded and replicated scenes is available on the companion website\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/web/index.html}}.
 
-Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows.
+%Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows.
 The standard deviation of assessments is represented using ellipses for one location (P1).
-The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{http://felixgontier.github.io/soundSourcePresenceEstimation/web/index.html}}.
+The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes.
 This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.
@@
     \centering
-    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
-    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).
-Arrows show discrepancies between corresponding recorded and replicated scenes.
-For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.}\label{fig:pspace_rec}
+    \includegraphics[width=0.8\columnwidth]{figures/pca_recrep.eps}
+    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).}\label{fig:pspace_rec}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
+    \includegraphics[width=0.8\columnwidth]{figures/pca_sim2.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
@@
 
+\section{Prediction of time of presence of sources}
+\label{sec:pred}
+
+\subsection{Corpus}
+\label{sec:pred_corp}
+
+Considering machine learning techniques to detect the presence of sources requires the availability of annotated data.
+Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed.
+This corpus is composed of sound scenes simulated using \textit{simScene} in the \textit{generate} mode with the same procedure as for the 75 simulated scenes in the listening test corpus described in Section~\ref{sec:exp_stim}.
+
+The simulated deep learning corpus is composed of two subsets:
+\begin{itemize}
+\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
+\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
+\end{itemize}
+
+The generation procedure is the same for both subsets, and reuses the distributions extracted from recordings of the GRAFIC project in Section~\ref{sec:exp_stim}.
+Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets.
+To do so, an isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is constructed that is independant from the one used in the generation of simulated scenes in the listening test corpus.
+This database contains 12mn, 49mn, and 73mn of background traffic, voice and bird extracts respectively, and 32mn, 12mn, and 5mn of event traffic, voice and bird extracts respectively.
+The isolated samples database is split in the same proportions as the two subsets: two-thirds for the development set and the remaining one-third for the evaluation set.
+As a result, the two subsets and the listening test simulated scenes are simulated using three separate, independant isolated samples databases.
+
 \subsection{Acoustical indicators for soundscape description}
-\label{sec:data_inds}
+\label{sec:pred_inds}
 
-\begin{figure*}[t!]
-    \centering
-    \includegraphics[width=1.5\columnwidth]{figures/pred_ex.png}
-    \caption{Example of simulated sound scene with ground truth source contributions and resulting presence predictions by the deep learning architecture (greyed out areas)}\label{fig:pred_ex}
-\end{figure*}
-
-To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:data_corp}, the activity of sound sources present in each scene must be known.
+To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:pred_corp}, the activity of sound sources present in each scene must be known.
 Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source.
@@
 
-\begin{align}\label{eq:ts}
-\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
-\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
-\end{align}
-
-where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene.
-This indicator evaluates the presence or absence of a given source in a small time frame.
-First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated.
-Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold.
-A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene.
-The optimal threshold values are optimized via grid search so that the time of presence estimation $\hat T_s(\alpha_{opt}, \beta_{opt})$ is best correlated with the average perceived time of presence $T_{s, p}$ as obtained by human evaluation.
-Both thresholds are optimized once in this study using data from the listening test in Section~\ref{sec:data_exp} and equal $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$, though several other pairs of values yielded close performance as per the optimization metric defined in eq.\ref{eq:ts}.
+\begin{equation}\label{eq:ts}
+\hat T_s(\alpha, \beta) = \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
+\end{equation}
+where $s$ denotes the sound source, $N_t$ is the number of considered time frames, $N_f$ is the number of frequency bands.
+This indicator evaluates the of percentage of presence of a given source with regard to other sources.
+First, for each time frame, frequency bands for which the source is emergent, that is the difference between its sound level and that of all other sources $\Delta L_{s}(t, f)$ is greater than the $\alpha$ threshold value, are isolated.
+Then, the source is considered present in this time frame if the emergence for these bands is on average greater than the $\beta$ threshold.
+A time of presence estimation is obtained for each source by averaging the resulting binary presence values along the $N_t$ time frames composing the scene.
+The optimal threshold values are found via grid search so that the time of presence estimation $\hat T_s(\alpha_{opt}, \beta_{opt})$ is best correlated with the average perceived time of presence $T_{s, p}$ as obtained by human evaluation:
+\begin{equation}\label{eq:ts_opt}
+\alpha_{opt},\beta_{opt} = \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
+\end{equation}
+where $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $r$ denotes the Pearson correlation coefficient and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene.
+Both thresholds are optimized once in this study using data from the listening test in Section~\ref{sec:exp}, and equal $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$, though several other pairs of values yielded close performance as per the optimization metric defined in eq.\ref{eq:ts}.
 
@@
 
-The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception.
+The considered indicators are compared to subjective assessments obtained during the listening test in Section~\ref{sec:exp} to identify their capacity to explain dimensions of soundscape perception and annotate the deep learning corpus.
 The analysis is performed using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes.
@@
 \centering
-\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
+%\resizebox{\textwidth}{!}{
+\caption{Pearson's correlation coefficients between physical and perceptual (resp.
+pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) indicators (n = 92).}
 \label{tab:physc}
-\resizebox{\textwidth}{!}{
+\begin{threeparttable}[t]
 \begin{tabular}{ c | c c c c c | c c c c }
 \hline
-	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
-	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
-	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
-	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
-	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
-	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
-	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
-	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\
-	$L_{10}-L_{90}$ & NS & NS & -0.24* & NS & -0.22* & NS & NS & NS & NS \\ \hline
-	$TFSD_{500Hz, 1s}$ & NS & 0.41** & NS & 0.28** & NS & -0.24* & -0.39** & 0.74** & NS \\
-	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.45** & -0.54** & NS & 0.63** \\ \hline
-	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
-	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
-	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
-	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
-	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
-	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
-	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & NS & 0.35** & -0.57** & -0.29** & 0.56** & 0.81** & -0.39** & -0.37** \\
-	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.35** & NS & NS & -0.39** & 0.81** & NS \\
-	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.46** & -0.57** & NS & 0.91** \\ \hline
+	 & P & L & OL & I & C & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
+	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.66** & 0.07 & -0.41** \\
+	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.63** & 0.06 & -0.35** \\
+	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.71** & 0.06 & -0.46** \\
+	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.71** & 0.06 & -0.47** \\
+	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.71** & 0.03 & -0.44** \\
+	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.67** & 0.07 & -0.40** \\
+	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.73** & 0.08 & -0.50** \\
+	$L_{10}-L_{90}$ & 0.13 & -0.18 & -0.24* & -0.06 & -0.22* & -0.01 & -0.01 & -0.09 \\ \hline
+	$TFSD_{500Hz, 1s}$ & 0.07 & 0.41** & 0.11 & 0.28** & -0.15 & -0.39** & 0.74** & -0.17 \\
+	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.54** & -0.18 & 0.63** \\ \hline
+	$L_{eq, T}$ & -0.58** & 0.20 & 0.46** & -0.46** & -0.42** & 0.71** & -0.16 & -0.36** \\
+	$L_{eq, V}$ & -0.17 & 0.50** & 0.31** & 0.08 & -0.37** & -0.04 & 0.71** & -0.40** \\
+	$L_{eq, B}$ & 0.27* & -0.04 & -0.11 & 0.35** & 0.18 & -0.24* & -0.04 & 0.71** \\ \hline
+	$\Delta L_T$ & -0.45** & -0.11 & 0.26* & -0.59** & -0.22* & 0.66** & -0.51** & -0.26* \\
+	$\Delta L_V$ & 0.04 & 0.50** & 0.17 & 0.35** & -0.20 & -0.38** & 0.59** & -0.01 \\
+	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & 0.08 & 0.25* & -0.25* & -0.10 & -0.03 \\ \hline
+	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & -0.05 & 0.35** & -0.57** & -0.29** & 0.81** & -0.39** & -0.37** \\
+	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & 0.12 & 0.44** & 0.05 & 0.35** & -0.11 & -0.39** & 0.81** & -0.16 \\
+	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.57** & -0.08 & 0.91** \\ \hline
 \end{tabular}
-}
+\begin{tablenotes}
+\item *: p<0.05, **: p<0.01
+\end{tablenotes}
+%}
+\end{threeparttable}
 \end{table*}
 
-Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments.
+Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments, with two-tailed significance tests.
 First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
-Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71).
+Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source $s$ (r=0.71).
 Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources.
@@
 
-\section{Prediction of time of presence of sources}
-\label{sec:pred}
-
 \subsection{Deep learning for presence prediction}
@@
 
-A deep learning model is implemented using the Python Pytorch~\cite{paszke2017} framework and trained on the corpus presented in Section~\ref{sec:data_corp} for source time of presence prediction.
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
+    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
+\end{figure}
+
+A deep learning model is implemented using the Python Pytorch~\cite{paszke2017} framework and trained on the corpus presented in Section~\ref{sec:pred_corp} for source time of presence prediction.
 
@@
 Ground truth target outputs are associated to each input frame to train the model.
-Considering the results discussed in Section~\ref{sec:data_inds}, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ time of presence estimation seems well suited to automatically label the dataset for this task.
+Considering the results discussed in Section~\ref{sec:pred_inds}, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ time of presence estimation seems well suited to automatically label the deep learning corpus for this task.
 Binary presence values on 1~s frames obtained during the $\hat T_s(\alpha_{opt}, \beta_{opt})$ computation before averaging in eq.\ref{eq:ts} are thus used as individual weak labels.
@@
 
-\begin{figure}[h!]
+\begin{figure*}[t!]
     \centering
-    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
-    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
-\end{figure}
+    \includegraphics[width=1.5\columnwidth]{figures/pred_ex.png}
+    \caption{Example of simulated sound scene with ground truth source contributions and resulting presence predictions by the deep learning architecture (greyed out areas)}\label{fig:pred_ex}
+\end{figure*}
 
@@
 
-The performance of the source detection deep learning architecture presented in Section~\ref{sec:pred_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}.
+The performance of the source detection deep learning architecture presented in Section~\ref{sec:pred_deep} on the evaluation subset of the deep learning corpus is summarized in Table~\ref{tab:perf_cmp}.
 The overall presence detection accuracy is 92.11\%.
@@
 \centering
-\caption{Performance of the model predicting source presence with binary ground truth labels obtained from $\hat T_s(\alpha_{opt}, \beta_{opt})$.
+\caption{Performance of the detection model predicting source presence with binary ground truth labels obtained from $\hat T_s(\alpha_{opt}, \beta_{opt})$.
 Presence metrics are computed for n=8600 1s frames and time of presence metrics on n=200 45s scenes.
@@
 Figure~\ref{fig:pred_ex} presents an example of predictions for a sound scene in the listening test corpus.
-The model takes the mixed audio (top) as its input, and outputs binary presence predictions for traffic, voice and bird sources on 1s time frames.
+The detection model takes the mixed audio (top) as its input, and outputs binary presence predictions for traffic, voice and bird sources on 1s time frames.
 Here the ground truth contributions are shown as reference, though they are never available to the deep learning architecture.
@@
 In the context of urban soundscape quality assessment for pedestrians, pleasantness emerges as the first affective dimension~\cite{axelsson2010}.
-Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
-To demonstrate the potential of the proposed method for predicting the time of presence of sources, it is applied to pleasantness estimation.
+Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources.
+Technological, human and natural sources are the most influent on urban soundscape quality, and are sometimes specified as traffic, voices and birds respectively~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}.
+One of the source activity descriptors used in existing models is the perceived time of presence.
+Thus, the proposed method for predicting the time of presence of sources can be applied to pleasantness prediction in an urban context.
+In this section the proposed approach is compared to baseline models from perceptual or acoustical variables to study its potential in the prediction of high-level perceptual descriptors such as the pleasantness.
 
@@
 \centering
-\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
+\begin{threeparttable}
+\caption{Pearson's correlation coefficients between perceptual scales (resp. pleasantness, overall loudness, time of presence of traffic, voices and birds) at the scene level (n=100)}
 \label{tab:percc}
@@
 \end{tabular}
+\begin{tablenotes}
+\item *: p<0.05, **: p<0.01
+\end{tablenotes}
 %}
+\end{threeparttable}
 \end{table}
 
-Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:data_exp}.
+Baseline models of pleasantness are constructed using data from the experiment in Section~\ref{sec:exp}.
 We first study the relationships between perceptual scales with respect to existing work.
-Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100).
+Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100), with two-tailed significance tests.
 The correlations between perceived source activity and pleasantness are consistent with the literature~\cite{aumond2017, gontier2018}.
@@
 
-Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged for each scene (n=100).
+Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged over participants for each scene of the listening test corpus (n=100).
 All possible combinations of variables among the 4 predictors (15 combinations) are considered, and to ensure that no multi-collinearity exists between predictors, a variance inflation factor (VIF) test is performed prior to each regression.
@@
 where $OL$, $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$ represent the perceived overall loudness and time of presence of traffic, voices and birds respectively.
+The F-statistic of $\hat P_{1, p}$ model is 229 (p<0.0001) and the t-statistics are 26.6, -15.5, -4.5, 2.9 and 4.8 for the intercept, $OL$, $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$ respectively.
 This expression is close to existing models of pleasantness in the literature~\cite{ricciardi2014, aumond2017}, with both the overall loudness and traffic activity negatively contributing and bird activity positively contributing to pleasantness.
@@
 
-Similarly, multilinear regression models of pleasantness are constructed using acoustical indicators described in Section~\ref{sec:data_inds}.
-Namely the $L_{50}$ and $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicators are used as predictors, as they display the highest correlation values with the overall loudness and time of presence of sources respectively.
+Similarly, multilinear regression models of pleasantness are constructed using acoustical indicators described in Section~\ref{sec:pred_inds}.
+As no ground truth source activity is available for the recorded scenes they are discarded from this operation.
+Furthermore, two simulated scenes containing only one source type are discarded for computational stability concerns.
+Multilinear models are thus constructed on the remaining 92 scenes of the listening test corpus.
+The $L_{50}$ and $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicators are used as predictors, as they display the highest correlation values with the overall loudness and time of presence of sources respectively.
 Again, all combinations of the proposed physical variables are considered, and a variance inflation factor check ($VIF<5$) is performed on predictors to ensure that no multi-collinearity between predictors exists in a model.
@@
 where $\varphi$ indicates a model from physical variables.
-Indicators related to traffic and voices activity are absent compared to the perceptual model.
+The F-statistic of $\hat P_{1, \varphi}$ model is 210 (p<0.0001) and the t-statistics are 21.6, -15.8, and 3.8 for the intercept, $L_{50}$ and $\hat T_B(\alpha_{opt}, \beta_{opt})$ respectively.
+Indicators related to traffic and voices activity are absent compared to the perceptual model $\hat P_{1, p}$.
 This is consistent with the results underlined in Table~\ref{tab:physc}: the $L_{50}$ used as a predictor is correlated to the traffic time of presence $T_{T, p}$ that appeared in the perceptual model $\hat P_{1, p}$ (eq.\ref{eq:pp1}), and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.
@@
 \centering
-\caption{Performance of baseline models for pleasantness prediction (**: p<0.01).}
+\caption{Performance of baseline models for pleasantness prediction.}
 \label{tab:pleasm}
+\begin{threeparttable}
 %\resizebox{\columnwidth}{!}{
@@
 \end{tabular}
+\begin{tablenotes}
+\item **: p<0.01
+\end{tablenotes}
 %}
+\end{threeparttable}
 \end{table}
@@
 \label{tab:pppred}
+\begin{threeparttable}
 %\resizebox{\columnwidth}{!}{
@@
 \end{tabular}
+\begin{tablenotes}
+\item **: p<0.01
+\end{tablenotes}
 %}
+\end{threeparttable}
 \end{table*}
@@
 These estimations are then applied to pleasantness estimation and compared to models presented in Section~\ref{sec:app_mdls}.
-To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the datasets simulated for the optimization of the deep learning model, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
+To evaluate the detection model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the listening test corpus is split in three parts: 1) the 6 recorded scenes which contain additional sources not present in the deep learning corpus, 2) the 19 replicated scenes that also include additional sources as annotated in~\cite{gloaguen2017}, and 3) the 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation subsets of the deep learning corpus.
 
@@
 Table~\ref{tab:pppred} presents the performance of pleasantness estimations using outputs from the deep learning architecture compared to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels computed with eq.~\ref{eq:ts} on separated source-specific channels.
-First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
+First, pleasantness estimations are equally effective using the detection model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus.
 This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}.
@@
 First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments.
-According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy.
+According to the results of the listening test presented in Section~\ref{sec:exp_res}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy.
 This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment.
@@
 Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra.
-To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.
+To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training subset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.
 

### 2019-11-26 paper/submitted/gontierAa2019v2.tex

--- 2019-10-07 paper/paperV2.tex
@@
 \usepackage{color}
+
+\graphicspath{{../figures/}{./}}
+
 \DeclareMathOperator*{\argmax}{arg\,max}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
+        \includegraphics[width=\textwidth]{tv_pres.eps}
     \end{subfigure}%
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
+        \includegraphics[width=\textwidth]{tb_pres.eps}
     \end{subfigure}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
+        \includegraphics[width=\textwidth]{vb_pres.eps}
     \end{subfigure}
@@
     \centering
-    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
+    \includegraphics[width=0.8\columnwidth]{pca_p1.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25).
@@
     \centering
-    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
+    \includegraphics[width=0.8\columnwidth]{pca_sim.eps}
     \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75).
@@
     \centering
-    \includegraphics[width=1.5\columnwidth]{figures/pred_ex.png}
+    \includegraphics[width=1.5\columnwidth]{pred_ex.png}
     \caption{Example of simulated sound scene with ground truth source contributions and resulting presence predictions by the deep learning architecture (greyed out areas)}\label{fig:pred_ex}
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
+    \includegraphics[width=\columnwidth]{deep_arch.pdf}
     \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
@@
 
-\bibliographystyle{unsrt}
-\bibliography{refs}
+\begin{thebibliography}{10}
+
+\bibitem{ardouin2018}
+J.~Ardouin, L.~Charpentier, M.~Lagrange, F.~Gontier, N.~Fortin, J.~Picaut,
+  D.~Ecoti\`ere, G.~Guillaume, and C.~Mietlicky.
+\newblock An innovative low-cost sensor for urban sound monitoring.
+\newblock In {\em INTER-NOISE 2018}, 2018.
+
+\bibitem{mydlarz2017}
+C.~Mydlarz, J.~Salamon, and J.P. Bello.
+\newblock The implementation of low-cost urban acoustic monitoring devices.
+\newblock {\em Applied Acoustics}, 117:207--218, 2017.
+
+\bibitem{iso2014}
+{ISO 12913-1:2014}.
+\newblock Acoustics - soundscape - part 1: definition and conceptual framework.
+\newblock Standard, International Organization for Standardization, Geneva, CH,
+  2014.
+
+\bibitem{can2008}
+A.~Can, L.~Leclerq, J.~Lelong, and J.~Defrance.
+\newblock Capturing urban traffic noise dynamics through relevant descriptors.
+\newblock {\em Applied Acoustics}, 69:1270--1280, 2008.
+
+\bibitem{brocolini2013}
+L.~Brocolini, C.~Lavandier, M.~Quoy, and C.~Ribeiro.
+\newblock Measurement of acoustic environments for urban soundscapes: choice of
+  homogeneous periods, optimization of durations, and selection of indicators.
+\newblock {\em J. Ac. Soc. Am.}, 134:813--821, 2013.
+
+\bibitem{nilsson2006}
+M.E. Nilsson and B.~Berglund.
+\newblock Soundscape quality in suburban green areas and city parks.
+\newblock {\em Acta Acust. unit. Acust.}, 92:903--911, 2006.
+
+\bibitem{lunden2016}
+P.~Lund\'en, O.~Axelsson, and M.~Hurtig.
+\newblock On urban soundscape mapping: A computer can predict the outcome of
+  soundscape assessments.
+\newblock In {\em INTER-NOISE 2016}, 2016.
+
+\bibitem{aucouturier2007}
+J.-J. Aucouturier, B.~Defreville, and F.~Pachet.
+\newblock The bag-of-frames approach to audio pattern recognition: a sufficient
+  model for urban soundscapes but not for polyphonic music.
+\newblock {\em J. Ac. Soc. Am.}, 122:881--91, 2007.
+
+\bibitem{cakir2015}
+E.~Cakir, T.~Heittola, H.~Huttunen, and T.~Virtanen.
+\newblock Polyphonic sound event detection using multi label deep neural
+  networks.
+\newblock In {\em 2015 International Joint Conference on Neural Networks
+  (IJCNN)}, 2015.
+
+\bibitem{gontier2017}
+F.~Gontier, M.~Lagrange, P.~Aumond, A.~Can, and C.~Lavandier.
+\newblock An efficient audio coding scheme for quantitative and qualitative
+  large scale acoustic monitoring using the sensor grid approach.
+\newblock {\em Sensors}, 17, 2017.
+
+\bibitem{aumond2017}
+P.~Aumond, A.~Can, B.~De Coensel, D.~Botteldooren, C.~Ribeiro, and
+  C.~Lavandier.
+\newblock Modeling soundscape pleasantness using perceptive assessments and
+  acoustic measurements along paths in urban context.
+\newblock {\em Acta Acust. unit. Acust.}, 103:430--443, 2017.
+
+\bibitem{torija2013}
+A.~J. Torija, D.~P. Ruiz, and A.~F. Ramos-Ridao.
+\newblock Application of a methodology for categorizing and differentiating
+  urban soundscapes using acoustical descriptors and semantic-differential
+  attributes.
+\newblock {\em J. Ac. Soc. Am.}, 134:791--802, 2013.
+
+\bibitem{nilsson2007}
+M.E. Nilsson, D.~Botteldooren, and B.~De Coensel.
+\newblock Acoustic indicators of soundscape quality and noise annoyance in
+  outdoor urban areas.
+\newblock In {\em 19th International Congress on Acoustics}, 2007.
+
+\bibitem{viollon2000}
+S.~Viollon and C.~Lavandier.
+\newblock Multidimensional assessment of the acoustic quality of urban
+  environments.
+\newblock In {\em INTER-NOISE 2000}, 2000.
+
+\bibitem{axelsson2010}
+O.~Axelsson, M.E. Nilsson, and B.~Berglund.
+\newblock A principal components model of soundscape perception.
+\newblock {\em J. Ac. Soc. Am.}, 128:2836, 2010.
+
+\bibitem{cain2013}
+R.~Cain, P.~Jennings, and J.~Poxon.
+\newblock The development and application of the emotional dimensions of a
+  soundscape.
+\newblock {\em Applied Acoustics}, 74:232--239, 2013.
+
+\bibitem{jeon2018}
+J.Y. Jeon, J.Y. Hong, C.~Lavandier, J.~Lafon, O.~Axelsson, and M.~Hurtig.
+\newblock A cross-national comparison in assessment of urban park soundscapes
+  in {F}rance, {K}orea, and {S}weden through laboratory experiments.
+\newblock {\em Applied Acoustics}, 133:107--117, 2018.
+
+\bibitem{aletta2016}
+F.~Aletta, J.~Kang, and O.~Axelsson.
+\newblock Soundscape descriptors and a conceptual framework for developing
+  predictive soundscape models.
+\newblock {\em Landsc. Urban Plan.}, 149:65--74, 2016.
+
+\bibitem{delaitre2014}
+P.~Delaitre, C.~Lavandier, C.~Ribeiro, M.~Quoy, E.~D'Hondt, E.~Gonzalez Boix,
+  and K.~Kambona.
+\newblock Influence of loudness of noise events on perceived sound quality in
+  urban context.
+\newblock In {\em Inter Noise}, 2014.
+
+\bibitem{guastavino2007}
+C.~Guastavino.
+\newblock Categorization of environmental sounds.
+\newblock {\em Can. J. Exp. Psychol.}, 61:54--63, 2007.
+
+\bibitem{gygi2007}
+B.~Gygi, G.~R. Kidd, and C.~S. Watson.
+\newblock Similarity and categorization of environmental sounds.
+\newblock {\em Perception \& Psychophysics}, 69:839--55, 2007.
+
+\bibitem{brown2011}
+A.L. Brown.
+\newblock Towards standardization in soundscape preference assessment.
+\newblock {\em Applied Acoustics}, 72:387--392, 2011.
+
+\bibitem{lavandier2006}
+C.~Lavandier and B.~Defreville.
+\newblock The contribution of sound source characteristics in the assessment of
+  urban soundscapes.
+\newblock {\em Acta Acust. unit. Acust.}, 92:912--921, 2006.
+
+\bibitem{ricciardi2014}
+P.~Ricciardi, P.~Delaitre, C.~Lavandier, F.~Torchia, and P.~Aumond.
+\newblock Sound quality indicators for urban places in paris cross-validated by
+  {M}ilan data.
+\newblock {\em J. Ac. Soc. Am.}, 138:2337--2348, 2014.
+
+\bibitem{hong2017}
+J.Y. Hong and J.Y. Jeon.
+\newblock Relationship between spatiotemporal variability of soundscape and
+  urban morphology in a multifunctional urban area: A case study in {S}eoul,
+  {K}orea.
+\newblock {\em Building and Environment}, 126:382--95, 2017.
+
+\bibitem{lavandier2016}
+C.~Lavandier, P.~Aumond, S.~Gomez, and C.~Domingu\`es.
+\newblock Urban soundscape maps modelles with geo-referenced data.
+\newblock {\em Noise Mapping}, 3:278--94, 2016.
+
+\bibitem{mesaros2017}
+A.~Mesaros, T.~Heitolla, A.~Diment, B.~Elizalde, A.~Shah, E.~Vincent, B.~Raj,
+  and T.~Virtanen.
+\newblock {DCASE} 2017 challenge setup: tasks, datasets and baseline system.
+\newblock In {\em Detection and Classification of Acoustic Scenes and Events
+  (DCASE) 2017 workshop}, 2017.
+
+\bibitem{foggia2015}
+P.~Foggia, N.~Petkov, A.~Saggese, N.~Strisciuglio, and M.~Vento.
+\newblock Reliable detection of audio events in highly noisy environments.
+\newblock {\em Pattern Recognition Letters}, 65:22--8, 2015.
+
+\bibitem{mcfee2018}
+B.~Mc{F}ee, J.~Salamon, and J.~P. Bello.
+\newblock Adaptive pooling operators for weakly labeled sound event detection.
+\newblock {\em IEEE Transactions on Audio, Speech and Language Processing},
+  26:2180--93, 2018.
+
+\bibitem{adavanne2017}
+S.~Adavanne, P.~Pertila, and T.~Virtanen.
+\newblock Sound event detection using spatial features and convolutional
+  recurrent neural network.
+\newblock In {\em 2017 IEEE International Conference on Audio, Speech and
+  Signal Processing (ICASSP)}, 2017.
+
+\bibitem{boes2018}
+M.~Boes, K.~Filipan, B.~De Coensel, and D.~Botteldooren.
+\newblock Machine listening for park soundscape quality assessment.
+\newblock {\em Acta Acust. unit. Acust.}, 104:121--30, 2018.
+
+\bibitem{barchiesi2015}
+D.~Barchiesi, D.~Giannoulis, D.~Stowell, and M.~D. Plumbley.
+\newblock Acoustic scene classification.
+\newblock {\em IEEE Signal Processing Magazine}, 32:16--34, 2015.
+
+\bibitem{sun2019}
+K.~Sun, B.~De Coensel, K.~Filipan, F.~Aletta, T.~Van Renterghem, T.~De
+  Pessemier, W.~Joseph, and D.~Botteldooren.
+\newblock Classification of soundscapes of urban public open spaces.
+\newblock {\em Landscape and Urban Planning}, 189:139--55, 2019.
+
+\bibitem{valenti2016}
+M.~Valenti, A.~Diment, G.~Parascandolo, S.~Squartini, and T.~Virtanen.
+\newblock {DCASE} 2016 acoustic scene classification using convolutional neural
+  networks.
+\newblock In {\em Detection and Classification of Acoustic Scenes and Events
+  (DCASE) 2016 workshop}, 2016.
+
+\bibitem{salamon2017-2}
+J.~Salamon and J.~P. Bello.
+\newblock Deep convolutional neural networks and data augmentation for
+  environmental sound classification.
+\newblock {\em IEEE Signal Processing Letters}, 24, 2017.
+
+\bibitem{yu2009}
+L.~Yu and J.~Kang.
+\newblock Modeling subjective evaluation of soundscape quality in urban open
+  spaces: An artificial neural network approach.
+\newblock {\em J. Ac. Soc. Am.}, 126:1163--74, 2009.
+
+\bibitem{mesaros2019}
+A.~Mesaros, A.~Diment, B.~Elizalde, T.~Heitolla, E.~Vincent, B.~Raj, and
+  T.~Virtanen.
+\newblock Sound event detection in the {DCASE} 2017 challenge.
+\newblock {\em IEEE Transactions on Audio, Speech and Language Processing},
+  27:992--1006, 2019.
+
+\bibitem{salamon2014}
+J.~Salamon, C.~Jacoby, and J.~P. Bello.
+\newblock A dataset and taxonomy for urban sound research.
+\newblock In {\em 22nd ACM International Conference on Multimedia}, 2014.
+
+\bibitem{gemmeke2017}
+J.~F. Gemmeke, D.~Ellis, D.~Freedman, A.~Jansen, W.~Lawrence, R.~Moore,
+  M.~Plakal, and M.~Ritter.
+\newblock Audio {S}et: an ontology and human-labeled dataset for audio events.
+\newblock In {\em 2017 IEEE International Conference on Audio, Speech and
+  Signal Processing (ICASSP)}, 2017.
+
+\bibitem{lafay2016}
+Gr{\'e}goire Lafay, Mathieu Lagrange, Mathias Rossignol, Emmanouil Benetos, and
+  Axel Roebel.
+\newblock {A morphological model for simulating acoustic scenes and its
+  application to sound event detection}.
+\newblock {\em {IEEE/ACM Transactions on Audio, Speech and Language
+  Processing}}, 24(10):1854--1864, 2016.
+
+\bibitem{salamon2017}
+J.~Salamon, D.~MacConnell, M.~Cartwright, P.~Li, and J.~P. Bello.
+\newblock Scaper: A library for soundscape synthesis and augmentation.
+\newblock In {\em 2017 IEEE Workshop on Applications of Signal Processing to
+  Audio and Acoustics (WASPAA)}, 2017.
+
+\bibitem{panayotov2015}
+V.~Panayotov, G.~Chen, D.~Povey, and S.~Khudanpur.
+\newblock Librispeech: an {ASR} corpus based on public domain audio books.
+\newblock In {\em 2015 IEEE International Conference on Acoustics, Speech and
+  Signal Processing (ICASSP)}, 2015.
+
+\bibitem{gloaguen2017}
+J.R. Gloaguen, A.~Can, M.~Lagrange, and J.F. Petiot.
+\newblock Creation of a corpus of realistic urban sound scenes with controlled
+  acoustic properties.
+\newblock In {\em Meetings on Acoustics}, 2017.
+
+\bibitem{gontier2018}
+F.~Gontier, P.~Aumond, M.~Lagrange, C.~Lavandier, and J.-F. Petiot.
+\newblock Towards perceptual soundscape characterization using event detection
+  algorithms.
+\newblock In {\em Detection and Classification of Acoustic Scenes and Events
+  (DCASE) 2018 workshop}, 2018.
+
+\bibitem{dagnelie2003}
+P.~Dagnelie.
+\newblock {\em Principes d'exp\'erimentation: planification des exp\'eriences
+  et analyse de leurs r\'esultats}.
+\newblock Presses Agronomiques de Gembloux, 2003.
+
+\bibitem{moller1992}
+H.~M{\o}ller.
+\newblock Fundamentals of binaural technology.
+\newblock {\em Applied Acoustics}, 36:171--218, 1992.
+
+\bibitem{wilcoxon1945}
+F.~Wilcoxon.
+\newblock Individual comparisons by ranking methods.
+\newblock {\em Biometrics Bulletin}, 1:80--3, 1945.
+
+\bibitem{pratt1959}
+J.~W. Pratt.
+\newblock Remarks on zeros and ties in the {W}ilcoxon signed rank procedures.
+\newblock {\em Journal of the American Statistical Association}, 54:655--67,
+  1959.
+
+\bibitem{itatoolbox2017}
+M.~Berzborn, R.~Bomhardt, J.~Klein, J.G. Richter, and M.~Vorlander.
+\newblock The ita-toolbox: an open source matlab toolbox for acoustic
+  measurements and signal processing.
+\newblock In {\em 43th Annual German Congress on Acoustics}, 2017.
+
+\bibitem{paszke2017}
+A.~Paszke, S.~Gross, S.~Chintala, G.~Chanan, E.~Yang, Z.~DeVito, Z.~Lin,
+  A.~Desmaison, L.~Antiga, and A.~Lerer.
+\newblock Automatic differentiation in pytorch.
+\newblock In {\em 2017 Conference Neural Information Processing Systems
+  (NIPS)}, 2017.
+
+\bibitem{kingma2015}
+D.~P. Kingma and J.~Lei Ba.
+\newblock Adam: a method for stochastic optimization.
+\newblock In {\em 2015 International Conference on Learning Representations
+  (ICLR)}, 2015.
+
+\bibitem{blauert1997}
+J.~Blauert and U.~Jekosch.
+\newblock Sound quality evaluation: a multi layered problem.
+\newblock {\em Acta Acust. unit. Acust.}, 83:747--53, 1997.
+
+\bibitem{jekosch2004}
+U.~Jekosch.
+\newblock Basic concepts and terms quality, reconsidered in the context of
+  product-sound quality.
+\newblock {\em Acta Acust. unit. Acust.}, 90:999--1006, 2004.
+
+\bibitem{lasseck2018}
+M.~Lasseck.
+\newblock Acoustic bird detection with deep convolutional neural networks.
+\newblock In {\em Detection and Classification of Acoustic Scenes and Events
+  (DCASE) 2016 workshop}, 2018.
+
+\end{thebibliography}
 

### 2019-11-28 paper/paperV3.tex
Final V3

@@
 To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}.
-These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~mn in duration.
+These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, as seen in Figure~\ref{fig:path}.
+The original recordings range from 55~s to 4.5~mn in duration.
 In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}).
@@
 Though many studies dealing with the environmental acoustic quality are associated with acoustic measurements ranging from a few seconds~\cite{paulsen1997, brambilla2006} to fifteen minutes~\cite{kuwano1997, decoensel2007} and even to eighty minutes~\cite{namba1988}, stimuli between 30 seconds and 1 minute are often preferred for laboratory tests.
-To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence for each type of sources (ranging from no presence to 100\% of presence) without redundancy.
+To avoid simulated scenes to be too similar in the listening experiment corpus, a selection of 75 simulated scenes is carried out from the initial 200 scenes, in order to cover a large diversity of presence (ranging from no presence to 100\% of presence) for each type of sources without redundancy.
+In the case of three sources, a 3-dimensional space is constructed where each dimension corresponds to one of the three sources.
+The selection is then operated by taking the most isolated sound scenes, where the euclidean distance in this space is considered as a measure of the differences between sound scenes.
+
 Playback sound levels are realistically drawn for each scene, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.
@@
 The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception.
-The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode.
+The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene} in the \textit{replicate} mode, using the same isolated samples database as for the 75 simulated scenes.
 The 19 resulting sound scenes are thus simulated, but their scenarios follow those of the reference recordings to the extent of annotation precision.
 Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization.
-The 6 recorded and 19 replicated scenes are scaled using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.
+The 6 recorded and 19 replicated scenes are normalized using the relation found in Section~\ref{sec:exp_equi} so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL over the 19 locations.
 
@@
 By repeating this procedure for different settings of the generator level, the relation between the logarithm of the generator output voltage and headphones playback sound level in dB is approximated as a linear function.
-From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener, when using the calibrated headphones and a specific sound card configuration for which the relation between the numeric level and output voltage is known.
+From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.
 
@@
 \caption{Mean differences of perceptual assessments (resp.
-pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) between original and replicated sound scenes.
+Pleasantness, Liveliness, Overall Loudness, Interest, Calmness, Time of presence of Traffic, Voices and Birds) between original and replicated sound scenes.
 Significant differences as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
@@
 
-Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point scales (0-10).
-These attributes are presented in French and translated in this report using standard terminology.
+Participants are asked to evaluate sound scenes using 8 perceptual attributes represented by 11-point semantic differential rating scales (0-10).
+These attributes are presented in French and translated in this paper using standard terminology.
 The first 5 semantic differential scales relate to general properties of the scene:
@@
 These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}.
-Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point scales:
+Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11-point semantic differential scales:
 \begin{itemize}
@@
 
-The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab.
+The statistic analyses carried out in this paper (statistic comparison tests, principal component analysis) are done with Matlab R2015b (Statistics and Machine Learning Toolbox v10.1).
 The effect of the \textit{simScene} generation procedure on perception is first investigated.
@@
 Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios.
-Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold.
+Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and semantic differential scale to outline significant differences between assessment distributions, which are shown in bold.
 As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}.
-On the first five scales, all mean differences are lower than 2 points on the 11-point discrete scale.
+On the first five scales, all mean differences are lower than 2 points.
 Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters.
@@
 Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy.
-For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.
+For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness: they share a common \textit{voice} class which is characterized in the isolated samples database by recordings of read texts.
 Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.
@@
 The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of perceived source activity. Its expression is:
+
 \begin{equation}
@@
 \end{equation}
+
 where $L(f, t)$ is the third-octave spectrum of the signal.
@@
 \end{equation}
+
 where $s$ denotes the sound source, $N_t$ is the number of considered time frames, $N_f$ is the number of frequency bands.
@@
 \caption{Pearson's correlation coefficients between physical and perceptual (resp.
-pleasantness, liveliness, overall loudness, interest, calmness, time of presence of traffic, voices and birds) indicators (n = 92).}
+Pleasantness, Liveliness, Overall Loudness, Interest, Calmness, Time of presence of Traffic, Voices and Birds) indicators (n = 92).}
 \label{tab:physc}
@@
 
-Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments, with two-tailed significance tests.
+Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments, with two-tailed significance tests (see Matlab \textit{corrcoef}).
 First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness.
@@
 \begin{threeparttable}
-\caption{Pearson's correlation coefficients between perceptual scales (resp. pleasantness, overall loudness, time of presence of traffic, voices and birds) at the scene level (n=100)}
+\caption{Pearson's correlation coefficients between perceptual scales (resp. Pleasantness, Overall Loudness, Time of presence of Traffic, Voices and Birds) at the scene level (n=100)}
 \label{tab:percc}

### 2019-11-29 paper/submitted/coverLetter2.tex
sub

--- 2019-11-26 paper/coverLetter2.tex
@@
 $\rightarrow$ All elements provided to the reviewers as part of the previous response letter are now presented in the manuscript.
-In particular, the choice of 45~s long extracts for the listening test and absence of controlled background noise in simulated sound scenes are now included in Section 2.1.
+In particular, the choice of 45~s long extracts for the listening test and absence of controlled background noise in simulated sound scenes are now included in Section 2.1, on rows 255-262 and 204-207 respectively.
 
@@
 However, the isolated samples database is different.
-Details of the isolated samples database have been added in Section 2.1 and 3.1  for the listening test corpus and the deep learning corpus respectively.
+Details on the contents the isolated samples database have been added in Section 2.1 (rows 197-204) and 3.1 (rows 508-512) for the listening test corpus and the deep learning corpus respectively.
 
@@
 
-$\rightarrow$ The description of the generation procedure from distributions extracted from the GRAFIC project recordings using simScene has been extended in Section 2.1.
+$\rightarrow$ The description of the generation procedure from distributions extracted from the GRAFIC project recordings using simScene has been extended in Section 2.1, with a presentation of simScene simulation modes and inputs on rows 162-193.
 
@@
 
-$\rightarrow$ The introduction of Section 4 has been extended to better motivate the objective of this application.
+$\rightarrow$ The objective of Section 4 is to present a potential application of the proposed detection model to the prediction of higher-level perceptual attributes, and assess the gain of such an approach with respect to known models.
+The introduction of Section 4 has been extended to better present this motivation.
 
@@
 
-$\rightarrow$ As is now more clearly described in Section 2.1, replicated sound scenes are generated using simScene, with scene composition (scenarios) constrained by annotations of the reference recordings.
-This differs from the 75 simulated sound scenes for which the scene composition is sampled from gaussian distributions.
+$\rightarrow$ As is now more clearly described in Section 2.1 (rows 292-299), replicated sound scenes are generated using simScene, with scene composition (scenarios) constrained by annotations of the reference recordings.
+This differs from the 75 simulated sound scenes for which the scene composition is obtained by sampling from gaussian distributions.
 
@@
 To avoid this phenomenon in the listening test corpus, 200 scenes are first generated, among which 75 scenes are selected to maximize content diversity.
+This is described on rows 262-274 of the revised manuscript.
 
@@
 
-$\rightarrow$ The selection procedure description has been simplified and Figure 1 has been removed to enhance the reader's comprehension, as the considered indicator for the selection of the 75 simulated scenes was not presented until Section 2.4 (now 3.2).
+$\rightarrow$ The dimensions of the mentioned 3D space correspond to the presence (0-100\%) of traffic, voices and birds sources.
+The 75 simulated scenes of the listening test corpus are selected to be the most isolated, that is distant from others with regard to the euclidean distance, from the original pool of 200 simulated scenes.
+The selection procedure description has been rewritten on rows 262-274 and Figure 1 has been removed as it did not directly contribute to this description.
 
@@
 
-$\rightarrow$ The sound scenes are scaled so that they are heard at the desired sound level (as measured in the case of recorded and replicated scenes) when using the specific configuration (headphones, sound card) presented in Section 2.2.
-The term "normalized" has been replaced with "scaled" to avoid confusion.
+$\rightarrow$ The sound scenes are normalized so that they are heard at the desired sound level (as measured in the case of recorded and replicated scenes, or as sampled conditionnally to the ambiance in the case of simulated scenes).
+This requires the relation between the electrical level at the output of the sound card and the resulting sound level produced by the headphones, which is obtained in Section 2.2.
 
@@
 
-$\rightarrow$ The term "Likert scale" has been removed throughout the manuscript.
+$\rightarrow$ The term "Likert scale" has been replaced with "semantic differential rating scale" throughout the manuscript.
 
@@
 
-$\rightarrow$ The steps undertaken during the calibration procedure are now detailed in Section 2.2.
-From this procedure, a simple (linear) relation is obtained between the voltage at the output of the sound card and the sound level produced by the headphones.
-Sound scenes can be scaled using this information so that they are played at the desired sound level using the specific headphones and sound card configuration.
+$\rightarrow$ From the calibration procedure, a simple (linear) relation is obtained between the voltage at the output of the sound card and the sound level produced by the headphones.
+Sound scenes can be normalized using this information so that they are played at the desired sound level using the specific headphones and sound card configuration.
+The steps undertaken during the calibration procedure are now detailed in Section 2.2 (rows 315-347).
 
@@
 
-$\rightarrow$ The 19 replicated scenes are generated by simScene, and thus are not recordings. This part has been rewritten to avoid any confusion in this regard.
+$\rightarrow$ The 19 replicated scenes are generated by simScene, and thus are not recordings. This part has been rewritten to avoid any confusion in this regard (rows 543-455).
 
@@
 
-$\rightarrow$ All statistical analyses are conducted using Matlab.
-This is emphasized in the introduction as well as in the beginning of Section 2.5.
+$\rightarrow$ All statistical analyses are conducted using Matlab R2015b with the Statistics and Machine Learning Toolbox v10.1.
+This is now emphasized in Section 2.5, and reminded where appropriate.
 The code is also made available for further information about the specific functions used.
@@
 
-$\rightarrow$ The description of both equations has been extended to enhance the reader's comprehension.
+$\rightarrow$ Both equations are now presented separately, and each one is more thoroughly explained.
 
@@
 The corresponding significance test is two-tailed.
-This is now stated in Section 3.2 and 4.1, where correlation coefficients are presented.
+This is now stated in Section 3.2 (rows 632-635) and 4.1 (rows 666-669), where correlation coefficients are presented.
 
@@
 The baseline model from acoustic indices is computed on a subset of 92 sound scenes as the computation of time of presence estimations requires separated source contributions, which are not available in the 6 recorded scenes.
-These elements are now more clearly described in Section 4.1.
+These elements are now more clearly described in Section 4.1 (rows 783-787 and 809-817).
 
